---
title: "Registration accuracy — ANHIR challenge"
author: "sceriff0"
date: "2026-09-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(readr)
  library(ggplot2)
  library(here)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the plotting module is never sourced — which would silently leave the
# empty-state plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

SLUG       <- "benchmark_anhir"
anhir_dir  <- here("data", "benchmark")
have_anhir <- file.exists(file.path(anhir_dir, "anhir_cases.csv"))
```

## Starting data and notation

**ANHIR** (Automatic Non-rigid Histological Image Registration,
[anhir.grand-challenge.org](https://anhir.grand-challenge.org/)) is the public
benchmark for registering serially-sectioned, differently-stained histology
images. It ships pairs of whole-slide images at several downsampled scales with
pathologist-placed **landmarks**, so a method's error is measured against human
correspondences rather than against its own features — the one thing neither the
[synthetic sweep](benchmark_registration.html) nor the
[run QC](registration_run_qc.html) can offer. Here mirage's two backends,
**VALIS** and **STARE** (`tiled`), are scored on it beside the challenge's own
**bUnwarpJ** baseline and the **initial** (unregistered) state.

**Where the numbers come from.** mirage's `benchmarks/anhir/` harness registers
every ANHIR pair with each method and scores the result; its
`benchmarks/pull_to_ihc_method.sh` copies two tables into `data/benchmark/`:

- **`anhir_cases.csv`** — one row per (case, method): the case's `tissue`,
  `scale` and `status`, then `n_landmarks`, `rtre_median` / `rtre_mean` /
  `rtre_max`, `tre_median_px`, `robustness`, `rank_median_rtre` and `time_min`.
- **`anhir_aggregates.csv`** — one row per (method, subset), for the subsets
  `all`, `training`, `evaluation`, `robust` (cases with robustness > 0.5) and one
  `tissue:<name>` per tissue: the means and medians of the per-case metrics plus
  `avg_rank_median_rtre`.

**The metrics**, as the challenge defines them
([Performance Metrics](https://anhir.grand-challenge.org/Performance_Metrics/)):

- **TRE** — target registration error: for each landmark, the distance in pixels
  between the warped source landmark and its target.
- **rTRE** — *relative* TRE: TRE divided by the target image's diagonal. Unitless,
  so it compares across image sizes and scales; a good non-rigid result sits
  around 10^-3^, no registration around 10^-1^. Every figure below uses rTRE.
- **Robustness** — the fraction of a case's landmarks whose rTRE *improved* over
  the initial (unregistered) state. A case is "robust" above 0.5.
- **Ranking** — the challenge's primary metric. On each case the methods are
  ranked by their **median rTRE** (1 = best), and each method's rank is averaged
  over cases; lower is better. `avg_rank_median_rtre` for `subset == "all"` is
  that number.

**Only training cases are scored here.** ANHIR releases target landmarks for the
*training* split only; *evaluation* cases are scored server-side on
grand-challenge.org from submitted warped landmarks. So in `anhir_cases.csv`
evaluation rows carry `scored = FALSE` and NA metrics, every figure draws the
training cases, and the `evaluation` aggregate row is empty until a submission
is scored on the server.

```{r build}
anhir_figs <- list()
anhir <- NULL
if (have_anhir) {
  source(here("code", "anhir_plots.R"))
  anhir <- anhir_load(anhir_dir)
  anhir_figs <- list(
    "01_rtre_by_method" = plot_anhir_rtre_by_method(anhir$cases),
    "02_rtre_by_tissue" = plot_anhir_rtre_by_tissue(anhir$cases),
    "03_robustness"     = plot_anhir_robustness(anhir$cases),
    "04_rank"           = plot_anhir_rank(anhir$aggregates))
  cat(nrow(anhir$cases), "case rows (", sum(anhir$cases$scored %in% TRUE), "scored ) and",
      nrow(anhir$aggregates), "aggregate rows loaded from", basename(anhir_dir), ".\n")
} else {
  cat("No ANHIR benchmark data found — the tables have not been pulled yet.",
      "Run mirage's `benchmarks/pull_to_ihc_method.sh` to copy `anhir_cases.csv` and",
      "`anhir_aggregates.csv` into `data/benchmark/`, then re-knit.")
}
```

```{r fig-notes}
fig_notes <- c(
  "01_rtre_by_method" = paste(
    "Per-case **median rTRE** by method (`anhir_cases.csv`, `rtre_median`, rows with",
    "`scored = TRUE`), one point per case, one facet per `status`. rTRE is TRE over the",
    "image diagonal, so it is unitless; lower = better. **Log y axis** — the unregistered",
    "state is one to two orders of magnitude worse than any registered result, and on a",
    "linear axis every method collapses onto zero."),
  "02_rtre_by_tissue" = paste(
    "The same `rtre_median`, summarised as the median over cases per (method, tissue).",
    "Tissues are ordered hardest-first by their pooled median, so the profile reads as a",
    "difficulty ladder and a method that only wins on easy tissue is visible. n per tissue",
    "counts cases, not (case, method) rows."),
  "03_robustness" = paste(
    "`robustness` per method: the fraction of a case's landmarks whose rTRE improved over",
    "no registration. `initial` is omitted because robustness is defined against it (its",
    "own value is identically 0). Dashed line = the challenge's robust cut, 0.5, the same",
    "threshold that defines the `robust` subset in `anhir_aggregates.csv`."),
  "04_rank" = paste(
    "The challenge's primary metric: `avg_rank_median_rtre` for `subset == \"all\"` in",
    "`anhir_aggregates.csv` — each method's rank on per-case median rTRE, averaged over",
    "the cases it was scored on. 1 = best; lower = better. n is the aggregate's `n_cases`.")
)
```

## Figures

Each figure is preceded by its derivation note. A figure whose input has no
scored rows renders as an empty panel that says so.

```{r figures, results="asis", fig.width=fig_width("double"), fig.height=5.5}
if (length(anhir_figs)) {
  for (nm in names(anhir_figs)) {
    title <- gsub("_", " ", sub("^[0-9]+[a-z]?_", "", nm))
    cat("\n\n### ", title, "\n\n", sep = "")
    if (!is.na(fig_notes[nm])) cat(fig_notes[[nm]], "\n\n")
    print(anhir_figs[[nm]])
    cat("\n\n")
  }
}
```

## Summary table

One row per (subset, method) from `anhir_aggregates.csv`, for the `all` and
`training` subsets, methods in rank order. `mean_rank` is
`avg_rank_median_rtre`; the rTRE columns are means and medians of the per-case
values; `avg_time_min` is wall time per case where the harness recorded it.

```{r summary-table, results="asis"}
if (have_anhir) {
  knitr::kable(anhir_summary_table(anhir$aggregates), digits = 4,
               caption = "ANHIR aggregates by method (lower rank and rTRE = better; higher robustness = better).")
}
```

## Notes

- **Landmarks, not self-assessment.** The sweep page and the run-QC page score a
  registration with the features the method itself matched (VALIS rTRE, STARE
  tile error) or with an independent DAPI overlap; ANHIR scores it against
  pathologist-placed correspondences, so it is the one benchmark here where the
  error is external to every method.
- **Training split only.** The evaluation split's landmarks are withheld by the
  challenge; its rows are present in `anhir_cases.csv` so the case inventory is
  complete, but unscored. A server-side score would arrive as a leaderboard
  entry, not through this page.
- **Rank is the headline, rTRE is the reason.** The challenge ranks on mean rank
  of per-case median rTRE (§4); the distributions (§1–§2) and robustness (§3) say
  *why* a method ranks where it does — a method can win the rank while losing a
  tissue.
- **This is not the study's tissue.** ANHIR's pairs are differently-stained
  serial sections at reduced scale; the study's cyclic-IF rounds are the same
  section re-imaged. For the registration accuracy of the run that produced
  this study's cells, see [run QC](registration_run_qc.html) and
  [registration arms](registration_arms.html).

```{r export-pdfs, include=FALSE}
# Every plot above is also rendered to PDF by the dev = c("png", "pdf") chunk option;
# this collects them into output/figures/<SLUG>/. Runs last, so they all exist by now.
export_pdf_figures(SLUG)
```
