---
title: "clinical_data (mirage phenotyping)"
author: "sceriff0"
date: "2026-08-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the shared
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R
# pulls in cell_tables.R (export-schema adapter) and plot_theme.R (house figure style).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "membership.R"))           # membership_data(): the one variant knob
source(here("code", "aggregation_compare.R"))  # annotation-aggregation sensitivity grid
source(here("code", "pdf_export.R"))           # export_pdf_figures(): plots -> output/figures/

SLUG <- "clinical_data_mirage"
```

## Starting data and notation

This is the **mirage phenotyping** variant. It is the same analysis as
[`clinical_data`](clinical_data.html) with the same polygons and the same
statistics — the only thing that changes is **who called the phenotypes**.
`clinical_data` uses the FlowPath gate tree drawn interactively in QuPath; this page
uses mirage's own constrained phenotyping (`bin/phenotype_cells.py`, driven by
`panel.yaml`), which runs inside the pipeline with no human in the loop. The diff
between the two pages is therefore attributable to the phenotyping method alone,
just as the diff between the two flag pages isolates the re-exported annotations.

Every value below is derived from three inputs:

- `mem$cells` — mirage single cells, one row per cell, assembled by
  `load_mirage_cells()` from **two** per-patient tables joined on the segmentation
  `label`:
    - `data/mirage/<patient>/phenotyping/phenotypes.csv` — the `phenotype` call plus
      the per-marker `sign:<M>` / `state:<M>` columns and the `pheno_score:<TYPE>`
      posteriors.
    - `data/mirage/<patient>/cell_properties/morphology.csv` — `label`, the centroid
      (`x`/`y`, in **pixels**, renamed `x_px`/`y_px` on load) and the shape
      descriptors.
    - optionally `quantification/merged_quant.csv` for the raw intensities, whose
      columns are named for the measurement (`"CD3: Cytoplasm: Median"`).

  `<patient>` is the directory name — nothing inside the files identifies the
  patient — normalised through `slide_key()`.
- `neoplastic_data` — pathologist tumour-content percentage per annotation,
  `ANNOTATION_1..3`.
- `clinical_data` — the clinical CRF (`HOT score`, `Immuno-phenotype`, blood counts,
  `RESPONSE`, `OS`/`PFS`).

Membership (which cells count as "inside the tumour annotation"):

```
cell is INSIDE  = its centroid falls in the pathologist geojson polygon
                  (sf point-in-polygon), the same rule and the same polygons
                  clinical_data.Rmd uses
per annotation  = inside one polygon (ANNOTATION_k)
union           = inside the dissolved union of the patient's polygons

NO FALLBACK: a patient with mirage cells but no geojson contributes nothing.
```

**There is no fallback here, by design.** The flag variants can fall back to the
exporter's `Out_of_annotation` column when a polygon is missing; mirage emits no
such flag, so the only honest alternative would be to count the whole slide as
"inside", which would silently inflate that patient's denominators. Those patients
are listed below instead.

Because a polygon *is* read, this variant has an annotation **area**, so
`area_mm2` and the `dens_*_per_mm2` densities carry real values — they are
comparable with `clinical_data` and absent from the two flag pages.

**Coordinates are already pixels.** mirage's morphology centroids are regionprops
coordinates on the SEGMENT mask, i.e. the registered reference frame the QuPath
geojson is drawn in. `cell_centroids_px()` passes them through untouched, where a
FlowPath micron export is divided by 0.325. Applying both conversions would put
every cell outside its annotation, so an implausibly low in-annotation count is the
symptom to look for — see [annotation membership QC](annotation_membership_qc.html).

**The two tools name the same taxonomy differently.** FlowPath's tree writes
`PANCK+Tumor`, `T helper`, `CD8+ T reg`; mirage's panel writes `PANCK_Tumor`,
`T_helper`, `CD8_Treg`. `phenotype_lineage` is joined on a punctuation-free key
(`pheno_join_key()`), so both collapse to the same lineages and the two pages are
directly comparable. Two differences are real rather than cosmetic, and are worth
holding in mind when reading any composition panel:

- mirage has **`Myeloid`** and **`Macrophage_M2`** leaves that FlowPath's tree does
  not; both fold into `Immune_other`, where FlowPath would have called the same
  cells plain `Immune`.
- mirage never writes `Unknown`. It emits four reserved outcomes —
  `Unclassified`, `Ambiguous`, `Conflict`, `Artefact` — for cells its constraint
  solver could not commit. All four count as unresolved, exactly as FlowPath's
  `Unknown` does, so the QC-filtered ("clean") denominators mean the same thing on
  both pages. The resolved percentage per patient is in the inventory below; a low
  one is the headline caveat for every panel that follows.

## Membership

```{r annotations}
# Pathologist geojson polygons, then the membership frames the whole report runs on.
# mode = "mirage": mirage's own phenotyped cells, geometric membership, no fallback.
annots <- load_annotations(patient_ids = neoplastic_data$SAMPLE)
mem    <- membership_data("mirage", ihc_data, neoplastic_data, annots = annots)
inv    <- mem$inventory

neo_ids      <- slide_key(neoplastic_data$SAMPLE)
mirage_ids   <- sort(unique(inv$patient_id))
have_geojson <- unique(slide_key(annots$patient_id))
no_polygon   <- setdiff(mirage_ids, have_geojson)   # cells, but nothing to place them in
no_cells     <- setdiff(neo_ids, mirage_ids)        # scored, but mirage produced nothing

cat("mirage patients loaded:", length(mirage_ids),
    "[", paste(mirage_ids, collapse = ", "), "]\n")
cat("patients with a pathologist score AND mirage cells:",
    length(intersect(neo_ids, mirage_ids)), "\n")
if (length(no_polygon))
  cat("!! mirage cells but NO geojson (absent from every metric, no fallback exists):",
      paste(no_polygon, collapse = ", "), "\n")
if (length(no_cells))
  cat("pathologist patients with NO mirage output:",
      paste(sort(no_cells), collapse = ", "), "\n")

knitr::kable(inv, caption = paste(
  "mirage cell inventory. n_resolved / pct_resolved exclude the four reserved",
  "outcomes (Unclassified, Ambiguous, Conflict, Artefact) — those cells stay in",
  "every denominator but carry no cell type, so a low pct_resolved dilutes each",
  "composition fraction. n_phenotypes counts the distinct resolved labels."))
```

```{r variant-hooks, include=FALSE}
# Everything past this point is shared with the other three clinical_data reports.
# These are the only values the shared body reads out of this one.
IMMUNE_HEADING   <- "whole-slide mirage immune readout vs clinical table"
IMMUNE_PROSE     <- "These use every mirage-phenotyped cell of every patient (whole slide, no annotation boundary) — the mirage analogue of the geojson report's whole-slide readout."
MEMBERSHIP_NOTES <- c(
  "Phenotypes are **mirage's own** (`bin/phenotype_cells.py` + `panel.yaml`), not FlowPath's gate tree. Membership, polygons, statistics and figures are identical to [`clinical_data`](clinical_data.html), so the difference between the two pages is the phenotyping method.",
  "Membership is **sf point-in-polygon** on the pathologist geojson, with **no fallback**: mirage carries no out-of-annotation flag, so a patient with no polygon is absent from every metric rather than counted whole-slide. Check the list above before concluding a patient is missing for a biological reason.",
  "Centroids are read as **pixels** (mirage's regionprops frame), not rescaled from microns. `area_mm2` and the `dens_*_per_mm2` columns carry real values here, as on the geojson page.",
  "mirage's `Myeloid` / `Macrophage_M2` leaves have no FlowPath counterpart and fold into `Immune_other`; its four unresolved outcomes (Unclassified, Ambiguous, Conflict, Artefact) all count as unresolved, matching FlowPath's `Unknown`.")
```

```{r body, child = "_clinical_data_body.Rmd"}
```
