---
title: "clinical_data (per-annotation flag)"
author: "sceriff0"
date: "2026-07-13"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the shared
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R pulls in
# cell_tables.R (export-schema adapter) and plot_theme.R (house figure style).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "membership.R"))           # membership_data(): the one variant knob
source(here("code", "aggregation_compare.R"))  # annotation-aggregation sensitivity grid
source(here("code", "pdf_export.R"))           # export_pdf_figures(): plots -> output/figures/

SLUG <- "clinical_data_per_annotation"
```

## Starting data and notation

This is the **per-annotation flag** variant: cells and tumour membership come from
the per-annotation exports rather than the whole-slide files plus geojson. It is the
`old/`-**ignoring** of the two flag variants — see
[`clinical_data_per_annotation_old`](clinical_data_per_annotation_old.html) for the
one that layers `per_annotation/old/` on top, and
[`clinical_data`](clinical_data.html) for the geometric reference.

Every value below is derived from three inputs:

- `ann_cells` — single cells, one row per cell, assembled by `membership_data()`
  from **two** places:
    - *primary* — the per-annotation exports
      `data/flowpath/per_annotation/<patient>_a<k>.csv` (**one csv per (patient,
      annotation)**). Each carries `phenotype`, per-marker sign columns, centroids
      and the precomputed out-of-annotation flag for that annotation. `patient_id` =
      the stem before `_a`; `annotation` = `ANNOTATION_<k>`.
    - *fallback* — for a patient with **no** per-annotation csv at all, the whole-slide
      file `data/flowpath/<patient>.csv`. Its single flag was computed over all that
      patient's annotations at once, so it gives **one** region. When the patient has
      exactly one scored annotation that region *is* that annotation and is labelled
      accordingly (`ANNOTATION_1` for both `10338` and `15897`), so the patient
      reaches every panel; otherwise it is labelled `"csv"` — see the note below.
- `neoplastic_data` — pathologist tumour-content percentage per annotation,
  `ANNOTATION_1..3`.
- `clinical_data` — the clinical CRF (`HOT score`, `Immuno-phenotype`, blood counts,
  `RESPONSE`, `OS`/`PFS`).

Membership (which cells count as "inside the tumour annotation"):

```
cell is INSIDE = the out-of-annotation flag is FALSE   (precomputed by the exporter)

per-annotation patients                    source = "flag"
  per annotation = inside cells of file <patient>_a<k>.csv   (annotation ANNOTATION_k)
  union          = inside cells pooled across a patient's annotation files, deduped
                   to one row per cell (segmentation id + centroid)

fallback patients (no per-annotation csv)  source = "csv"
  one region only = inside cells of <patient>.csv
  labelled        = the patient's sole scored ANNOTATION_k, when it has exactly one
                    (true for 10338 and 15897); otherwise "csv"
  union           = that same single region
```

There is no sf point-in-polygon and no geojson anywhere in this variant. With no
polygon there is no area, so every `dens_*_per_mm2` column (and `area_mm2`) is `NA`:
the density panels are empty by construction and only the fraction-based comparisons
carry values.

**Why the fallback region can be labelled with an annotation.** The per-annotation
panels join IHC to the pathologist on `(patient_id, annotation)`, so a region labelled
`"csv"` would never match `ANNOTATION_1..3` and the patient would silently drop out.
That is avoidable exactly when the patient has **one** scored annotation: the flag's
region and the pathologist's single scored annotation are then the same region, so
labelling it `ANNOTATION_1` compares like with like. Both fallback patients here
(`10338`, `15897`) are in that case, so **all six patients appear in every panel**.

The guard matters for future data. If a fallback patient ever has two or more scored
annotations, its one flag region spans all of them and cannot be attributed to any
single `ANNOTATION_k`. `load_flag_fallback_cells()` then falls back to the `"csv"`
label **and warns**, and that patient appears in the union panels only (the union
score is the mean over its annotations, which does match). Check the knit log for
such a warning whenever the cohort changes.

Either way `source` stays `"csv"`, so a fallback patient is never disguised as a
per-annotation one.

## Membership

```{r annotations}
# The membership frames the whole report runs on. mode = "flag": per-annotation files
# (top level only — old/ and new/ are NOT read) plus the whole-slide fallback for any
# patient with no per-annotation csv. neoplastic_data lets a fallback patient with a
# single scored annotation be labelled with that annotation, so it reaches every panel.
mem <- membership_data("flag", ihc_data, neoplastic_data)
inv <- mem$inventory

neo_ids  <- slide_key(neoplastic_data$SAMPLE)
per_ann  <- sort(unique(inv$patient_id[inv$file_origin == "per_annotation"]))
fallback <- sort(unique(inv$patient_id[inv$file_origin == "whole_slide"]))
missing  <- setdiff(neo_ids, unique(inv$patient_id))  # scored, but no cells at all

cat("per-annotation patients:", length(per_ann), "[", paste(per_ann, collapse = ", "), "]",
    "across", sum(inv$file_origin == "per_annotation"), "annotation file(s)\n")
cat("whole-slide FALLBACK patients (no per-annotation csv):", length(fallback),
    "[", paste(fallback, collapse = ", "), "]\n")
cat("patients with a pathologist score AND cells:",
    length(intersect(neo_ids, unique(inv$patient_id))), "\n")
if (length(missing))
  cat("pathologist patients with NO csv at all (absent from every IHC metric):",
      paste(sort(missing), collapse = ", "), "\n")

knitr::kable(inv, caption = paste(
  "Cell inventory. file_origin = per_annotation (a <patient>_a<k>.csv) or",
  "whole_slide (the fallback <patient>.csv, annotation \"csv\").",
  "n_inside = cells the flag puts inside."))
```

```{r variant-hooks, include=FALSE}
# Everything past this point is shared with the other two clinical_data reports.
# These are the only values the shared body reads out of this one.
IMMUNE_HEADING   <- "in-annotation IHC immune readout vs clinical table"
IMMUNE_PROSE     <- "These use the **per-patient union of in-annotation cells** (flag-inside cells pooled across a patient\'s annotation files, deduped to one row per cell) — the in-annotation analogue of the geojson report\'s whole-slide readout. For a fallback patient there is only one region, so its \"union\" is simply the flag-inside cells of the whole-slide csv."
MEMBERSHIP_NOTES <- c(
  "Membership is the exporter's **out-of-annotation flag** everywhere (`inside = flag FALSE`); no geojson, no sf point-in-polygon. `area_mm2` / `dens_*_per_mm2` are `NA` throughout.",
  "`source` is `flag` for patients read from a per-annotation `<patient>_a<k>.csv` and `csv` for fallback patients read from the whole-slide `<patient>.csv`. Check the inventory table above before reading any panel: a `csv` row is one region for the whole slide, not a single pathologist annotation.",
  "A fallback patient has ONE region, not a per-annotation breakdown. It is labelled with the patient's sole scored annotation when it has exactly one — so all six patients reach every panel — and `\"csv\"` (union panels only, with a warning) when it has more. Read a fallback patient's per-annotation point as \"the whole slide\", which for a single-annotation patient is the annotation.",
  "The `old/` subfolder is **ignored** here. `clinical_data_per_annotation_old.Rmd` is identical except that it layers `per_annotation/old/` over the top-level files, so differences between the two pages are attributable to those re-exported annotations alone.")
```

```{r body, child = "_clinical_data_body.Rmd"}
```
