---
title: "Annotation membership QC — sf point-in-polygon vs Out_of_annotation flag"
author: "sceriff0"
date: "2026-07-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(forcats)
  library(here)
  library(sf)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the analysis script that used to pull plot_theme.R in never runs —
# which silently left those plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (gives the per-arm neoplastic_* tables + the helpers). load_data.R prints
# colnames(); silence it.
# validation_helpers.R also sources code/plot_theme.R, which applies the house
# figure style (theme + palettes + geom defaults) for every plot below.
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

SLUG <- "clinical_membership_qc"

# WHICH ARM THIS PAGE CHECKS. The flag-vs-sf comparison is only meaningful WITHIN one
# arm: the arms' regions are drawn independently, so pairing one arm's csv with
# another's polygon would report a disagreement that is real geometry rather than the
# coordinate-scale problem this page exists to detect. Passing `spec` rather than a
# directory is what makes that pairing structural.
ARM        <- "massimo2"
spec       <- arm_spec(ARM)
neoplastic <- neoplastic_for(ARM)
```

## Starting data and notation

Every value below is derived from two inputs, compared **per cell**:

- the arm's per-region cell files (`spec$region_csv`), one row per cell, carrying
  **both** the cell `centroid` and the exporter's precomputed `Out_of_annotation` flag
  for the same cells. That both live in one file is what makes this a per-cell
  comparison rather than a cross-file match.
- the pathologist annotation **geojson** polygons (`ANNOTATION_k`, in pixels) from the
  SAME arm (`spec$region_poly`), for the patients in `neoplastic$SAMPLE`.

Both trees are keyed by the arm's own naming pattern (`arm_parse_name()`), which is
what guarantees a cell file and the polygon it is compared against refer to the same
region. For `massimo2` that pattern is a LETTER read by alphabet position (`A` is
`ANNOTATION_1`); for `massimo1` it is the literal `k` in an `_a<k>` suffix.

The main pipeline decides "inside the annotation" by **sf point-in-polygon** on the
pathologist geojson, mapping cell centroids into the pixel frame with the fixed
`centroid / 0.325`. This report cross-checks that against the shipped
`Out_of_annotation` flag. Because the region csv carries *both* the
centroid and the flag for the **same cells**, agreement is a direct per-cell
comparison (no cross-file matching):

```
flag inside = Out_of_annotation is FALSE
sf inside   = st_within( centroid / 0.325 , the ANNOTATION_k geojson polygon )
```

Per `(patient, annotation)` we report the two set sizes, their overlap (Jaccard and
% of cells that agree), and the resulting tumour fraction under each method. A large
flag-vs-sf gap on a slide means the fixed `0.325` conversion is wrong **for that
slide** — sf is landing on the wrong cells.

## Compute agreement

```{r qc}
# The region csvs and the polygons come from the same ARM, and are read through the
# same spec — so there is no way for this page to compare one arm's cells against
# another arm's regions.
REGION_DIR <- spec$region_csv$path

annots <- arm_annotations(spec, "region", patient_ids = neoplastic$SAMPLE)

if (is.null(annots)) {
  qc <- tibble::tibble()
  cat("No annotation polygons found for", ARM, "— nothing to compare the flag against.\n")
} else {
  qc <- annotation_membership_qc(spec, annots, um_per_px = 0.325)
  if (nrow(qc) == 0) cat("No region csvs found under", REGION_DIR, "\n")
}

# `qc` is empty when the arm's tree is absent or no csv carries both a centroid and
# the flag. Gate the table and the scatter on it rather than letting each fail on a
# column that was never created.
have_qc <- nrow(qc) > 0 && "pct_cells_agree" %in% names(qc)
if (!have_qc)
  cat("No comparable region csvs for arm `", ARM, "` — the table and scatter below",
      " are skipped.\n", sep = "")

if (nrow(qc)) {
  cat("annotations checked:", nrow(qc),
      "| mean cell-level agreement:",
      sprintf("%.1f%%", 100 * mean(qc$pct_cells_agree, na.rm = TRUE)),
      "| mean Jaccard:", sprintf("%.2f", mean(qc$jaccard, na.rm = TRUE)), "\n")
}
```

## Per-annotation agreement table

`pct_cells_agree` = fraction of cells given the same inside/outside call by both
methods. `jaccard` = |flag ∩ sf| / |flag ∪ sf| of the inside sets. The last two
columns are the tumour-fraction-inside each method would feed into the concordance
plots — they should be close if membership agrees.

```{r qc-table, eval = have_qc}
if (nrow(qc)) {
  qc |>
    mutate(across(c(jaccard, pct_cells_agree, tumor_frac_flag, tumor_frac_sf),
                  ~ round(.x, 3))) |>
    arrange(jaccard) |>
    knitr::kable(caption = paste("sf vs Out_of_annotation flag, per annotation.",
                                 "Low jaccard / pct_cells_agree = the /0.325 sf",
                                 "disagrees with the flag on that slide."))
}
```

## Tumour fraction: flag vs sf

Each point is one annotation; the dashed line is perfect agreement. Points off the
line are annotations where the two membership definitions give a different tumour
fraction — those are the slides to distrust the sf conversion on.

```{r qc-scatter, eval = have_qc}
if (nrow(qc) && sum(is.finite(qc$tumor_frac_flag) & is.finite(qc$tumor_frac_sf)) > 1) {
  d  <- qc |> filter(is.finite(tumor_frac_flag), is.finite(tumor_frac_sf))
  cc <- paired_cor3(d$tumor_frac_flag, d$tumor_frac_sf)   # Pearson + Spearman + Kendall
  ggplot(d, aes(tumor_frac_flag, tumor_frac_sf, colour = patient_id)) +
    geom_abline(slope = 1, linetype = "dashed", colour = REF_LINE) +
    geom_point(size = 3, alpha = 0.85) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    annotate("text", x = 0, y = 1, hjust = 0, vjust = 1,
             label = sprintf("Pearson r = %.2f\nSpearman rho = %.2f\nKendall tau = %.2f\n(n = %d)",
                             cc$pearson, cc$spearman, cc$kendall, cc$n)) +
    labs(title = "Tumour fraction inside: flag vs sf point-in-polygon",
         x = "Tumour fraction, export flag (unitless, 0-1)",
         y = "Tumour fraction, polygon membership (unitless, 0-1)",
         colour = "Patient")
}
```

## Notes

- Only patients with per-annotation csvs **and** a matching geojson are checked; a
  `note` column flags the ones skipped (missing geojson or columns).
- High agreement here validates the fixed `0.325` conversion for those slides; a
  slide with low agreement should either be excluded or given its own conversion.
- This compares *membership*, not the pathologist — that concordance is in
  the `clinical_massimo*.Rmd` pages.

```{r export-pdfs, include=FALSE}
# Every plot above is also rendered to PDF by the dev = c("png", "pdf") chunk option;
# this collects them into output/figures/<SLUG>/. Runs last, so they all exist by now.
export_pdf_figures(SLUG)
```
