---
title: "Clinical concordance — FlowPath phenotyping"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the shared
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R pulls in
# cell_tables.R (export-schema adapter) and plot_theme.R (house figure style).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "membership.R"))           # membership_data(): the one variant knob
source(here("code", "aggregation_compare.R"))  # annotation-aggregation sensitivity grid
source(here("code", "pdf_export.R"))           # export_pdf_figures(): plots -> output/figures/

SLUG <- "clinical_flowpath"
```

## Starting data and notation

Membership is decided **geometrically**, from the pathologist's own polygons — and in
this export, from the polygon belonging to the *same region* as the cell file it
scores. Its companion [`clinical_mirage`](clinical_mirage.html) runs the
identical analysis on cells MIRAGE phenotyped itself, over the same slides and the
same polygons, so the difference between the two pages is attributable to the
**phenotyping method** alone.

Every value below is derived from three inputs:

- `ihc_data` — single cells from the all-slide export,
  `data/all_slide/csv/<patient>/<patient>_<A|B|C>.csv`: **one csv per annotation
  region**, each covering the whole slide. `ihc_data` is the de-duplicated union of a
  patient's region files (one row per physical cell), so cohort-level fractions are not
  computed over an inflated denominator. Carries `phenotype`, per-marker sign columns,
  centroids and `patient_id`.
- `neoplastic_data` — the pathologist's neoplastic-cellularity percentage, **one row
  per scored region** (`ANNOTATION_1..3`, or `whole_slide` for a patient with no
  annotation drawn). Long rather than wide, because a whole-slide score cannot be put
  in an `ANNOTATION_k` column without claiming a region that was never drawn.
- `clinical_data` — the clinical CRF (`HOT score`, `Immuno-phenotype`, blood counts,
  `RESPONSE`, `OS`/`PFS`).

Membership (which cells count as "inside the tumour annotation"):

```
PER REGION (patient HAS an annotation directory):
  cell is INSIDE  = its centroid falls in THAT REGION'S polygon (sf point-in-polygon),
                    centroids mapped to the geojson pixel frame by
                    x_px = centroid_x / 0.325  (0.325 microns/pixel)
  union           = inside the dissolved union of the patient's polygons, scored on
                    the DE-DUPLICATED slide so a cell in two regions counts once

WHOLE SLIDE (patient has NO annotation directory):
  every cell is INSIDE — the layout's own convention, not a fallback guess
```

The `source` column on every row records which applied (`sf` or `whole_slide`). Region
`A` is `ANNOTATION_1`, `B` is `ANNOTATION_2` and so on **by alphabet position**, so the
regions line up with `neoplastic_data`'s `ANNOTATION_1..3` columns.

Because a polygon is read, `area_mm2` and every `dens_*_per_mm2` column carry real
values — except for a whole-slide patient, which has no polygon and therefore no area.

## Membership

```{r annotations}
# "all_slide" reads BOTH its cells and its polygons — they are paired region by
# region, so an externally supplied annotation set would not line up, and passing one
# is an error rather than being silently ignored.
mem <- membership_data("all_slide", ihc_data, neoplastic_data)

neo_ids   <- unique(slide_key(neoplastic_data$SAMPLE))
ihc_ids   <- unique(slide_key(ihc_data$patient_id))
have_poly <- unique(mem$per_annotation$patient_id[mem$per_annotation$source == "sf"])
missing   <- setdiff(neo_ids, ihc_ids)                        # no cell csv at all
whole     <- setdiff(intersect(neo_ids, ihc_ids), have_poly)  # cells but no polygon

cat("patients loaded:", length(ihc_ids),
    "[", paste(sort(ihc_ids), collapse = ", "), "]\n")
cat("patients with a pathologist score AND IHC cells:",
    length(intersect(neo_ids, ihc_ids)),
    "| scored per region against a polygon:", length(intersect(neo_ids, have_poly)),
    "| whole-slide (no annotation directory):", length(whole), "\n")
if (length(whole))
  cat("whole-slide patients (every cell counts, by the layout convention):",
      paste(sort(whole), collapse = ", "), "\n")
if (length(missing))
  cat("pathologist patients with NO cell csv (absent from every IHC metric):",
      paste(sort(missing), collapse = ", "), "\n")
```

```{r region-reconciliation}
# DOES THE PATHOLOGIST'S REGION LIST MATCH THE EXPORT'S? The concordance below is an
# inner join on (patient_id, annotation), so a region scored but not exported — or
# exported but not scored — drops out silently and the correlation is computed over
# fewer pairs than anyone intended. Print the comparison instead.
.scored   <- neoplastic_data |>
  transmute(patient_id = slide_key(SAMPLE), annotation, path_pct)
.exported <- mem$per_annotation |> select(patient_id, annotation, source, n_inside)
.recon <- full_join(.scored, .exported, by = c("patient_id", "annotation")) |>
  mutate(status = case_when(
    is.na(path_pct) ~ "exported, NOT scored",
    is.na(source)   ~ "scored, NOT exported",
    TRUE            ~ "matched")) |>
  arrange(patient_id, annotation)

knitr::kable(.recon, caption = paste(
  "Pathologist-scored regions vs regions the export produced. Only `matched` rows",
  "reach the concordance below."))

.bad <- filter(.recon, status != "matched")
if (nrow(.bad))
  cat("\n> **", nrow(.bad), " region(s) do not line up** and are excluded from every",
      " correlation on this page. Reconcile `neoplastic_data` in `code/load_data.R`",
      " against the export before reading the numbers.\n", sep = "")
```

```{r export-shape}
# WHAT THE REGION FILES CONTAIN — printed, not assumed. Either each <pid>_<L>.csv holds
# the whole slide with Out_of_annotation computed for region L, or it holds only that
# region's cells. The layout cannot say which, the two give different cohort-level
# denominators, and both look plausible — so the answer is reported here on every knit.
# `ihc_data` is de-duplicated either way, which is a no-op in the second case.
.rep <- attr(ihc_data, "overlap_report")
if (!is.null(.rep) && nrow(.rep)) {
  cat("**Export shape:**", attr(.rep, "verdict"), "\n\n")
  knitr::kable(.rep, caption = paste(
    "Rows read per patient vs distinct cells after de-duplication.",
    "rows_per_cell near 1 means the region files partition the slide;",
    "near the file count means each file exports the whole slide."))
}
```

```{r variant-hooks, include=FALSE}
# Everything past this point is shared with clinical_data_mirage.
# These are the only values the shared body reads out of this one.
IMMUNE_HEADING   <- "whole-slide IHC immune readout vs clinical table"
IMMUNE_PROSE     <- "These use every imaged cell of every patient (`ihc_data`, the de-duplicated union of the region files, no annotation boundary)."
MEMBERSHIP_NOTES <- c(
  "Membership is **sf point-in-polygon**, and each region's cells are scored against **that region's own polygon** — the export pairs them, so no cell is tested against a region it was not exported for. Centroids are mapped to the geojson pixel frame by `/0.325`.",
  "A patient with **no annotation directory** is entirely inside: the export layout's stated convention, marked `whole_slide` in the `source` column. That is the one case with no `area_mm2`, hence no densities.")
```

```{r body, child = here::here("analysis", "_children", "clinical_body.Rmd")}
```
