---
title: "clinical_data (geojson membership)"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the shared
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R pulls in
# cell_tables.R (export-schema adapter) and plot_theme.R (house figure style).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "membership.R"))           # membership_data(): the one variant knob
source(here("code", "aggregation_compare.R"))  # annotation-aggregation sensitivity grid
source(here("code", "pdf_export.R"))           # export_pdf_figures(): plots -> output/figures/

SLUG <- "clinical_data"
```

## Starting data and notation

This is the **geojson** variant, the reference report of the three: membership is
decided geometrically, from the pathologist's own polygons. The companions
[`clinical_data_per_annotation`](clinical_data_per_annotation.html) and
[`clinical_data_per_annotation_old`](clinical_data_per_annotation_old.html) run the
identical analysis off the exporter's precomputed flag instead, so any difference
between the three pages is attributable to the membership rule alone.

Every value below is derived from three inputs:

- `ihc_data` — single cells, **one csv per patient** (`<patient>.csv`), one row per
  cell. Carries `phenotype`, per-marker sign columns, centroids, `patient_id`, and
  (optionally) an out-of-annotation flag used only as a membership fallback.
- `neoplastic_data` — pathologist tumour-content percentage per annotation,
  `ANNOTATION_1..3`.
- `clinical_data` — the clinical CRF (`HOT score`, `Immuno-phenotype`, blood counts,
  `RESPONSE`, `OS`/`PFS`).

Membership (which cells count as "inside the tumour annotation"):

```
PRIMARY (patient HAS a geojson):
  cell is INSIDE  = its centroid falls in the pathologist geojson polygon
                    (sf point-in-polygon), with centroids mapped to the geojson
                    pixel frame by  x_px = centroid_x / 0.325  (0.325 microns/pixel)
  per annotation  = inside one polygon (ANNOTATION_k)
  union           = inside the dissolved union of the patient's polygons

FALLBACK (patient has NO geojson):
  cell is INSIDE = the out-of-annotation flag is FALSE  (one region, labelled "csv")
```

The `source` column on every row records which was used (`sf` or `csv`). This is the
only variant that reads a polygon, so it is the only one with an annotation **area**
— `area_mm2` and every `dens_*_per_mm2` column carry values here and are `NA` in the
flag variants.

## Membership

```{r annotations}
# Pathologist geojson polygons, then the membership frames the whole report runs on.
annots <- load_annotations(patient_ids = neoplastic_data$SAMPLE)
mem    <- membership_data("geojson", ihc_data, neoplastic_data, annots = annots)

neo_ids      <- slide_key(neoplastic_data$SAMPLE)
ihc_ids      <- unique(slide_key(ihc_data$patient_id))
have_geojson <- unique(slide_key(annots$patient_id))
missing      <- setdiff(neo_ids, ihc_ids)                           # no cell csv at all
csv_only     <- setdiff(intersect(neo_ids, ihc_ids), have_geojson)  # cells but no geojson

cat("patients loaded:", length(ihc_ids),
    "[", paste(sort(ihc_ids), collapse = ", "), "]\n")
cat("patients with a pathologist score AND IHC cells:",
    length(intersect(neo_ids, ihc_ids)),
    "| via sf geojson:", length(intersect(intersect(neo_ids, ihc_ids), have_geojson)),
    "| via CSV fallback:", length(csv_only), "\n")
if (length(missing))
  cat("pathologist patients with NO cell csv (absent from every IHC metric):",
      paste(sort(missing), collapse = ", "), "\n")
```

```{r variant-hooks, include=FALSE}
# Everything past this point is shared with the other two clinical_data reports.
# These are the only values the shared body reads out of this one.
IMMUNE_HEADING   <- "whole-slide IHC immune readout vs clinical table"
IMMUNE_PROSE     <- "These use every imaged cell of every patient (`ihc_data`, whole slide, no annotation boundary)."
MEMBERSHIP_NOTES <- c(
  "Membership is **sf point-in-polygon** on the pathologist geojson (centroids mapped to the pixel frame by `/0.325`); the exporter's out-of-annotation flag is used only for patients with no geojson. The `source` column marks which per row.",
  "Because a polygon is read, `area_mm2` and the `dens_*_per_mm2` columns carry real values here — the one thing the flag variants cannot report.")
```

```{r body, child = "_clinical_data_body.Rmd"}
```
