---
title: "Clinical concordance — FlowPath phenotyping"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the shared
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R pulls in
# cell_tables.R (export-schema adapter) and plot_theme.R (house figure style).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "membership.R"))           # membership_data(): the one variant knob
source(here("code", "aggregation_compare.R"))  # annotation-aggregation sensitivity grid
source(here("code", "pdf_export.R"))           # export_pdf_figures(): plots -> output/figures/

SLUG <- "clinical_flowpath"
```

## Starting data and notation

Membership is decided **geometrically**, from the pathologist's own polygons — and in
this export, from the polygon belonging to the *same region* as the cell file it
scores.

Phenotypes here are **FlowPath's**, from the QuPath gate tree. A companion page ran
the identical analysis on cells MIRAGE phenotyped itself — same slides, same polygons,
so the two differed by the phenotyping method alone — but it is withdrawn for now and
FlowPath is the only phenotyping on the site.

Every value below is derived from three inputs:

- `ihc_data` — single cells from the all-slide export,
  `data/all_slide/csv/<patient>/<patient>_<A|B|C>.csv`: **one csv per annotation
  region**, each covering the whole slide. `ihc_data` is the de-duplicated union of a
  patient's region files (one row per physical cell), so cohort-level fractions are not
  computed over an inflated denominator. Carries `phenotype`, per-marker sign columns,
  centroids and `patient_id`.
- `neoplastic_data` — the pathologist's neoplastic-cellularity percentage, **one row
  per scored region** (`ANNOTATION_1..3`, or `whole_slide` for a patient with no
  annotation drawn). Long rather than wide, because a whole-slide score cannot be put
  in an `ANNOTATION_k` column without claiming a region that was never drawn.
- `clinical_data` — the clinical CRF (`HOT score`, `Immuno-phenotype`, blood counts,
  `RESPONSE`, `OS`/`PFS`).

Membership (which cells count as "inside the tumour annotation"):

```
PER REGION (patient HAS an annotation directory):
  cell is INSIDE  = its centroid falls in THAT REGION'S polygon (sf point-in-polygon),
                    centroids mapped to the geojson pixel frame by
                    x_px = centroid_x / 0.325  (0.325 microns/pixel)
  union           = inside the dissolved union of the patient's polygons, scored on
                    the DE-DUPLICATED slide so a cell in two regions counts once

WHOLE SLIDE (patient has NO annotation directory):
  every cell is INSIDE — the layout's own convention, not a fallback guess
```

The `source` column on every row records which applied (`sf` or `whole_slide`). Region
`A` is `ANNOTATION_1`, `B` is `ANNOTATION_2` and so on **by alphabet position**, so the
regions line up with `neoplastic_data`'s `ANNOTATION_1..3` columns.

Because a polygon is read, `area_mm2` and every `dens_*_per_mm2` column carry real
values — except for a whole-slide patient, which has no polygon and therefore no area.

## Membership

```{r annotations}
# "all_slide" reads BOTH its cells and its polygons — they are paired region by
# region, so an externally supplied annotation set would not line up, and passing one
# is an error rather than being silently ignored.
mem <- membership_data("all_slide", ihc_data, neoplastic_data)

neo_ids   <- unique(slide_key(neoplastic_data$SAMPLE))
ihc_ids   <- unique(slide_key(ihc_data$patient_id))
have_poly <- unique(mem$per_annotation$patient_id[mem$per_annotation$source == "sf"])
missing   <- setdiff(neo_ids, ihc_ids)                        # no cell csv at all
whole     <- setdiff(intersect(neo_ids, ihc_ids), have_poly)  # cells but no polygon

cat("patients loaded:", length(ihc_ids),
    "[", paste(sort(ihc_ids), collapse = ", "), "]\n")
cat("patients with a pathologist score AND IHC cells:",
    length(intersect(neo_ids, ihc_ids)),
    "| scored per region against a polygon:", length(intersect(neo_ids, have_poly)),
    "| whole-slide (no annotation directory):", length(whole), "\n")
if (length(whole))
  cat("whole-slide patients (every cell counts, by the layout convention):",
      paste(sort(whole), collapse = ", "), "\n")
if (length(missing))
  cat("pathologist patients with NO cell csv (absent from every IHC metric):",
      paste(sort(missing), collapse = ", "), "\n")
```

```{r region-reconciliation}
# DOES THE PATHOLOGIST'S REGION LIST MATCH THE EXPORT'S? The concordance below is an
# inner join on (patient_id, annotation), so a region scored but not exported — or
# exported but not scored — drops out silently and the correlation is computed over
# fewer pairs than anyone intended. Print the comparison instead.
.scored   <- neoplastic_data |>
  transmute(patient_id = slide_key(SAMPLE), annotation, path_pct)
.exported <- mem$per_annotation |> select(patient_id, annotation, source, n_inside)
.recon <- full_join(.scored, .exported, by = c("patient_id", "annotation")) |>
  mutate(status = case_when(
    is.na(path_pct) ~ "exported, NOT scored",
    is.na(source)   ~ "scored, NOT exported",
    TRUE            ~ "matched")) |>
  arrange(patient_id, annotation)

knitr::kable(.recon, caption = paste(
  "Pathologist-scored regions vs regions the export produced. Only `matched` rows",
  "reach the concordance below."))

.bad <- filter(.recon, status != "matched")
if (nrow(.bad))
  cat("\n> **", nrow(.bad), " region(s) do not line up** and are excluded from every",
      " correlation on this page. Reconcile `neoplastic_data` in `code/load_data.R`",
      " against the export before reading the numbers.\n", sep = "")
```

```{r export-shape}
# WHAT THE REGION FILES CONTAIN — printed, not assumed. Either each <pid>_<L>.csv holds
# the whole slide with Out_of_annotation computed for region L, or it holds only that
# region's cells. The layout cannot say which, the two give different cohort-level
# denominators, and both look plausible — so the answer is reported here on every knit.
# `ihc_data` is de-duplicated either way, which is a no-op in the second case.
.rep <- attr(ihc_data, "overlap_report")
if (!is.null(.rep) && nrow(.rep)) {
  cat("**Export shape:**", attr(.rep, "verdict"), "\n\n")
  knitr::kable(.rep, caption = paste(
    "Rows read per patient vs distinct cells after de-duplication.",
    "rows_per_cell near 1 means the region files partition the slide;",
    "near the file count means each file exports the whole slide."))
}
```

```{r variant-hooks, include=FALSE}
# Scope wording for the sections below, kept as variables rather than inlined prose.
# They were the contract between this page and the mirage-phenotyping variant, which
# read the same body as a child document; that page is withdrawn and the body is
# inlined here now, but the seam is cheap to keep and is where a second phenotyping
# arm would plug back in.
IMMUNE_HEADING   <- "whole-slide IHC immune readout vs clinical table"
IMMUNE_PROSE     <- "These use every imaged cell of every patient (`ihc_data`, the de-duplicated union of the region files, no annotation boundary)."
MEMBERSHIP_NOTES <- c(
  "Membership is **sf point-in-polygon**, and each region's cells are scored against **that region's own polygon** — the export pairs them, so no cell is tested against a region it was not exported for. Centroids are mapped to the geojson pixel frame by `/0.325`.",
  "A patient with **no annotation directory** is entirely inside: the export layout's stated convention, marked `whole_slide` in the `source` column. That is the one case with no `area_mm2`, hence no densities.")
```

Cell labels:

```
tumour cell = phenotype_clean contains "Tumor"
              (phenotype_clean = clean_phenotype(phenotype): the text inside the
               parentheses when the gate label has any, the whole label otherwise)
CD45+ cell  = marker_pos(cells, "CD45") — the CD45 sign column, whichever of
              CD45_sign / sign:CD45 / CD45_positive the export carries
lineage(cell) = phenotype_clean collapsed via the phenotype_lineage table
```

`slide_key()` maps every id to the shared numeric slide space (strip non-digits,
apply the verified typo fix `15879` -> `15897`).

## Membership counts

For a set of cells (either one annotation's inside cells or the union inside
cells), `region_ratios()` returns:

```
n_inside          = number of cells in the set
n_inside_clean    = cells in the set after dropping Outlier == TRUE and
                    Unknown-phenotype cells (QC-filtered denominator)
n_tumor_inside    = tumour cells in the set
n_tumor_clean     = tumour cells in the set that survive the same clean filter
n_cd45_inside     = CD45+ cells in the set
n_cd3cd45_inside  = CD45+ AND CD3+ cells (marker-gated T cells)
n_gzmb_nk_inside  = NK-lineage cells that are GZMB+ (cytotoxic/activated NK)
n_<lineage>       = cells in the set with that lineage
                    (lineage in: CD8T, CD4T, Treg, NK, Immune_other, Stroma)

tumor_over_inside       = n_tumor_inside / n_inside
tumor_over_inside_clean = n_tumor_clean  / n_inside_clean  (outliers + Unknown removed)
cd45_over_inside    = n_cd45_inside    / n_inside
cd3cd45_over_inside = n_cd3cd45_inside / n_inside
cd3cd45_over_cd45   = n_cd3cd45_inside / n_cd45_inside   (T cells among immune)
gzmb_nk_over_inside = n_gzmb_nk_inside / n_inside
gzmb_nk_over_nk     = n_gzmb_nk_inside / n_NK            (GZMB+ fraction of NK)
frac_<lineage>      = n_<lineage>      / n_inside
```

Any fraction with a zero denominator is `NA`. The marker-gated subsets read the
marker sign columns via `marker_pos()`, not the phenotype label; a marker the export
never gated contributes no cells rather than aborting the report.

## Tumour content — IHC vs pathologist

The IHC tumour metric is shown under **three normalizations**, all restricted to the
cells inside the annotation:

```
tumor_over_inside       = (tumour cells inside) / (all cells inside)     -> 0..1
tumor_over_inside_clean = (tumour cells inside) / (cells inside)  -> 0..1
                          with the cell set first cleaned: FlowPath Outlier == TRUE
                          cells and Unknown-phenotype cells dropped from BOTH the
                          numerator and the denominator (a QC-filtered version of
                          tumor_over_inside; if a csv has no `Outlier` column no cell
                          is treated as an outlier)
tumor_over_cd45         = (tumour cells inside) / (CD45+ cells inside)   -> can exceed 1
dens_tumor_per_mm2      = (tumour cells inside) / (annotation area in mm^2)
                          area_mm2 = polygon pixel area * 0.325^2 / 1e6

pathologist (per annotation) = path_pct / 100, per scored region
pathologist (union)          = mean of that patient's scored-region values / 100
```

`tumor_over_inside_clean` answers the same IHC-vs-pathologist question as
`tumor_over_inside` but removes debris/segmentation outliers and unclassified cells
that dilute the denominator, so it is expected to track the pathologist's tumour
fraction more closely.

`dens_tumor_per_mm2` is the same tumour-cell density (cells/mm²) that periphery
reports, computed here inside the whole annotation. It needs the polygon area, so it
exists only where a polygon was read:

```{r density-scope, echo=FALSE, results="asis"}
# The flag variants read no polygon, so their density columns are structurally NA —
# an empty panel there is the expected outcome, not a data problem.
cat(if (mem$has_area)
      "it is defined for every patient with a geojson and `NA` for the fallback ones."
    else
      paste("it is **`NA` throughout this variant** — no polygon is read, so the",
            "density panels are empty and only the fraction-based comparisons carry",
            "values. The columns are kept so the schema matches the geojson report."),
    "\n")
```

Being a density rather than a 0..1 fraction, it is compared to the pathologist by
**rank** (Spearman/Kendall), not the x=y line.

Each normalization is computed both per-annotation (one polygon) and over the union
of a patient's polygons.

```{r tumor-content}
# Membership was resolved in the parent report; from here on the two frames have the
# same schema whichever mode produced them (see code/membership.R).
ihc_tumor_per_ann <- mem$per_annotation
ihc_tumor_union   <- mem$union

# neoplastic_data is already long — one row per scored region — so there is no pivot.
# Its `annotation` labels are the same vocabulary the metrics frames use, including
# `whole_slide` for a patient with no annotation drawn, so the join below is direct.
path_long <- neoplastic_data |>
  mutate(patient_id = slide_key(SAMPLE), path_frac = path_pct / 100) |>
  filter(!is.na(path_pct)) |>
  select(patient_id, annotation, path_pct, path_frac)

cc_ann <- ihc_tumor_per_ann |>
  inner_join(path_long, by = c("patient_id", "annotation")) |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

path_union <- path_long |>
  group_by(patient_id) |>
  summarise(path_frac = mean(path_frac), .groups = "drop")
cc_union <- ihc_tumor_union |>
  inner_join(path_union, by = "patient_id") |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

cat("per-annotation pairs:", nrow(cc_ann), " | union pairs:", nrow(cc_union), "\n")

# Diagnostic: the ACTUAL pairs that enter the correlation, largest gap first. Read
# the top rows to tell WHY the correlation is what it is:
#   - big gap + fallback `source` or tiny n_inside -> suspect membership, not biology
#   - big gap + a healthy primary-source patient   -> genuine IHC/pathologist disagreement
#   - per-annotation: verify each region letter lines up with the pathologist's score
cc_union |>
  transmute(patient_id, source, n_inside,
            area_mm2       = round(area_mm2, 3),
            ihc_tumor_frac = round(tumor_over_inside, 3),
            dens_tumor_mm2 = round(dens_tumor_per_mm2, 1),
            path_frac      = round(path_frac, 3),
            abs_gap        = round(abs(tumor_over_inside - path_frac), 3)) |>
  arrange(desc(abs_gap)) |>
  knitr::kable(caption = "UNION concordance pairs (largest IHC-vs-pathologist gap first).")

cc_ann |>
  transmute(patient_id, annotation, source, n_inside,
            ihc_tumor_frac = round(tumor_over_inside, 3),
            path_frac      = round(path_frac, 3),
            abs_gap        = round(abs(tumor_over_inside - path_frac), 3)) |>
  arrange(desc(abs_gap)) |>
  knitr::kable(caption = "PER-ANNOTATION concordance pairs (largest gap first). Confirm each region letter maps to the pathologist's score for that region.")
```

**Concordance statistics.** `paired_cor3(x, y)` reports three correlations between
IHC and pathologist across the shared patients, each `NA` when `n < 3` or either
vector is constant:

```
Pearson r    = linear correlation
Spearman rho = rank correlation (monotonic, robust to outliers)
Kendall tau  = rank concordance (pairwise, best-behaved at small n)
```

```{r tumor-content-plot}
# `metric` selects the IHC normalization; `unit_scale` draws the x=y line and forces
# a 0..1 square (only meaningful for tumor_over_inside — tumor_over_cd45 is unbounded
# and pathologist is 0..1, so their scatter is judged by RANK, not the diagonal).
# Points are coloured by `patient_id` (from cc_ann / cc_union) so a single patient's
# annotations read as one colour — useful in the per-annotation panel where a patient
# contributes several points.
plot_concordance <- function(df, metric, subtitle, xlab, unit_scale = TRUE) {
  d  <- dplyr::filter(df, is.finite(.data[[metric]]), is.finite(path_frac))
  cc <- paired_cor3(d[[metric]], d$path_frac)   # Pearson + Spearman + Kendall
  p <- ggplot(d, aes(.data[[metric]], path_frac)) +
    geom_point(aes(colour = patient_id), alpha = 0.7, size = 2) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
             label = sprintf("Pearson r = %.2f\nSpearman rho = %.2f\nKendall tau = %.2f\n(n = %d)",
                             cc$pearson, cc$spearman, cc$kendall, cc$n)) +
    labs(title = "Neoplastic cellularity: IHC vs pathologist", subtitle = subtitle,
         x = xlab, y = "Pathologist tumour fraction", colour = "Patient")
  if (unit_scale)
    p <- p + geom_abline(slope = 1, linetype = "dashed", colour = REF_LINE) +
             coord_equal(xlim = c(0, 1), ylim = c(0, 1))
  p
}

norm_specs <- list(
  list(metric = "tumor_over_inside",       xlab = "IHC tumour / all cells inside",  unit = TRUE),
  list(metric = "tumor_over_inside_clean", xlab = "IHC tumour / cells inside (excl. outliers + Unknown)", unit = TRUE),
  list(metric = "tumor_over_cd45",         xlab = "IHC tumour / CD45+ cells inside", unit = FALSE),
  list(metric = "dens_tumor_per_mm2",      xlab = "IHC tumour density (cells / mm²) inside", unit = FALSE)
)

for (s in norm_specs) {
  if (nrow(cc_ann) > 0)
    print(plot_concordance(cc_ann, s$metric, paste0(s$metric, " — per annotation"),
                           s$xlab, s$unit))
  if (nrow(cc_union) > 0)
    print(plot_concordance(cc_union, s$metric, paste0(s$metric, " — union"),
                           s$xlab, s$unit))
}
```

**Bland–Altman (union).** Per patient:

```
mean = (IHC + pathologist) / 2
diff = IHC - pathologist

bias   = average of diff over patients
LoA    = 1.96 * standard deviation of diff
limits = bias plus/minus LoA
```

Bland–Altman is strictly interpretable only when both quantities share the 0..1
scale (i.e. `tumor_over_inside`); for `tumor_over_cd45` it is shown for consistency,
but read it as descriptive since the CD45 ratio is unbounded.

```{r tumor-content-bland-altman}
plot_bland_altman <- function(df, metric, title) {
  d <- dplyr::filter(df, is.finite(.data[[metric]]), is.finite(path_frac))
  if (nrow(d) <= 2) return(invisible())
  ba <- d |> transmute(mean_val = (.data[[metric]] + path_frac) / 2,
                       diff_val = .data[[metric]] - path_frac)
  bias <- mean(ba$diff_val); loa <- 1.96 * sd(ba$diff_val)
  print(
    ggplot(ba, aes(mean_val, diff_val)) +
      geom_hline(yintercept = bias, colour = FIT_LINE) +
      geom_hline(yintercept = bias + c(-loa, loa), linetype = "dashed", colour = REF_LINE) +
      geom_point(alpha = 0.7, size = 2) +
      labs(title = title,
           subtitle = sprintf("bias = %.3f, limits of agreement +/- %.3f", bias, loa),
           x = "mean of the two", y = "IHC − pathologist")
  )
}

for (s in norm_specs)
  plot_bland_altman(cc_union, s$metric,
                    paste0("Bland–Altman (union): ", s$metric, " − pathologist"))
```

## Aggregation strategy — how collapsing the annotations moves the answer

The concordance panels above compare neoplastic cellularity at two fixed levels:
**per annotation** (one point per `ANNOTATION_k`) and **union** (one point per
patient). The union level hides a choice, because a patient with several annotations
has several IHC values *and* several pathologist values, and each side has to be
collapsed to one number before they can be paired. That collapse is not neutral.

**The mismatch this section exists to expose.** The union pair above collapses the
two sides by *different* rules:

```
pathologist   path_union      = mean(ANNOTATION_k / 100)          UNWEIGHTED mean
IHC           ihc_tumor_union = tumour cells / all cells,
                                over the POOLED cells of every
                                annotation of that patient        CELL-WEIGHTED mean
```

For a patient whose annotations differ in size these are different summaries of the
same slide. Take two annotations, 40 000 cells at 80% tumour and 4 000 cells at 50%:
the cell-weighted IHC value is 0.77 while the unweighted pathologist value is 0.65.
That 0.12 gap is **aggregation artefact**, not IHC-vs-pathologist disagreement — and
the existing union panels cannot tell the two apart.

So below, every IHC aggregator is paired against every pathologist aggregator:

```
IHC side (per patient)
  pooled      cells of every annotation pooled and deduped, ratio recomputed
              (exactly what the union panels above use)          cell-weighted
  wmean_ann   per-annotation fractions averaged, weights = that annotation's
              cell count                                          cell-weighted
  mean_ann    per-annotation fractions averaged, every annotation equal
  median_ann  median of the per-annotation fractions (only differs from mean_ann
              at 3+ annotations)

pathologist side (per patient)
  wmean_cells ANNOTATION_k scores averaged, weights = that annotation's IHC cell
              count — the pathologist analogue of pooling cells   cell-weighted
  mean        plain mean over the patient's annotations (used by the union panels)
  median      median over the annotations
  max         the highest-scoring annotation ("worst area governs")
  min         the lowest-scoring annotation (most conservative)
```

The weight is always the denominator of the metric being aggregated: `n_inside` for
`tumor_over_inside`, `n_inside_clean` for `tumor_over_inside_clean`. Weighting a
clean-set metric by the unclean count would over-weight annotations whose cells were
mostly discarded as outliers or Unknown.

One consequence to watch in the tables: a patient whose IHC region is labelled
`"csv"` rather than an `ANNOTATION_k` cannot be matched to a per-annotation
pathologist score, so it contributes no weight and drops out of the `wmean_cells`
rows while still appearing in the `mean` / `median` / `max` / `min` ones. That is why
`n` is printed on every row of the grid — compare rows only at equal `n`.

**Only two of the twenty combinations weight both sides the same way** —
`(pooled, wmean_cells)` and `(mean_ann, mean)`. Those are the pairs whose gap is
interpretable as genuine disagreement; everywhere else part of the gap is the
aggregation mismatch. They are shaded in the scatter grids and boxed in the heatmap.

**A built-in check.** `wmean_ann` should reproduce `pooled` almost exactly, since
pooling cells *is* cell-weighting them. The only thing separating those two rows is
the union's dedup of cells falling inside two overlapping annotations — so a visible
gap between them is a measurement of **annotation overlap**, not an aggregation
error. Read those two rows against each other before trusting anything else here.

```{r aggregation-grid}
# Every IHC aggregator x every pathologist aggregator, for both tumour-fraction
# normalizations. $pairs = the per-patient points; $stats = paired_cor3 + bias per
# combination. `bias` = mean(IHC - pathologist): the systematic offset that the
# correlation coefficients are deliberately blind to (a method can correlate
# perfectly and still read 15 points high).
agg      <- aggregation_grid(ihc_tumor_per_ann, ihc_tumor_union, path_long)
baseline <- per_annotation_baseline(ihc_tumor_per_ann, path_long)

agg$stats |>
  transmute(metric, ihc = ihc_lab, pathologist = path_lab,
            matched = if_else(matched, "yes", ""),
            n, pearson = round(pearson, 2), spearman = round(spearman, 2),
            kendall = round(kendall, 2), bias = round(bias, 3)) |>
  knitr::kable(caption = paste(
    "Aggregation grid, best Spearman first within each metric.",
    "matched = both sides weighted the same way (the interpretable rows).",
    "bias = mean(IHC - pathologist)."))

baseline |>
  transmute(metric, n, pearson = round(pearson, 2), spearman = round(spearman, 2),
            kendall = round(kendall, 2), bias = round(bias, 3)) |>
  knitr::kable(caption = paste(
    "PER-ANNOTATION baseline: no aggregation at all, one point per",
    "(patient, annotation). The reference the aggregated rows above should be read",
    "against — it neither gains aggregation's smoothing nor loses the",
    "within-patient signal."))
```

**Read the correlations with the `n` beside them.** This cohort gives a handful of
patients, so a Spearman rho here has a confidence interval spanning most of the
range, and the *ranking* of the twenty combinations is far less stable than the
numbers suggest. The point of the grid is not to pick the winner by rho — that would
be selecting on noise — but to see **whether the choice matters at all**. A flat grid
means the conclusion is robust to the aggregation; a grid with a large spread means
any single reported correlation is partly an artefact of the collapse rule.

```{r aggregation-heatmap, fig.width=8, fig.height=8}
# Reading a ROW shows how much the pathologist aggregator matters; a COLUMN, how much
# the IHC one does. Flat = the choice is immaterial for this cohort (the hoped-for
# outcome). Boxed cells are the matched-weighting pairs.
plot_aggregation_heatmap(agg$stats, "spearman")
```

```{r aggregation-scatter-raw, fig.width=11, fig.height=9}
# tumour / all cells inside. Dashed line is x = y; vertical distance from it is the
# disagreement. Shaded panels weight both sides the same way.
plot_aggregation_grid(agg$pairs, "tumor_over_inside")
```

```{r aggregation-scatter-clean, fig.width=11, fig.height=9}
# tumour / cells inside, with FlowPath Outlier == TRUE and Unknown-phenotype cells
# removed from both numerator and denominator.
plot_aggregation_grid(agg$pairs, "tumor_over_inside_clean")
```

Finally, which patients the choice actually bites on. A patient with one annotation
has zero spread by construction; the spread is the range a patient's value takes
across the aggregators, so large spreads mark the patients whose annotations
disagree most internally.

```{r aggregation-spread}
aggregation_spread(agg$pairs) |>
  transmute(metric, patient_id, n_pairs,
            ihc_min = round(ihc_min, 3), ihc_max = round(ihc_max, 3),
            ihc_spread = round(ihc_spread, 3),
            path_min = round(path_min, 3), path_max = round(path_max, 3),
            path_spread = round(path_spread, 3)) |>
  knitr::kable(caption = paste(
    "Per-patient spread across the aggregators (largest IHC spread first).",
    "Spread 0 = a single-annotation patient, where every aggregator coincides."))
```

### Recommendation

**For method validation, prefer the per-annotation level over any aggregation.** The
pathologist reports one score per annotation, so the annotation — not the patient —
is the natural unit at which the two methods can be compared like for like. It also
uses every observation instead of averaging most of them away, which matters at this
sample size. Aggregation is a compromise made necessary by the *clinical* joins
(survival, response, blood counts), which are per patient.

**When a patient-level number is required, use a matched pair.** Reporting a
cell-weighted IHC value against an unweighted pathologist value — what the union
panels currently do — mixes a real effect with an artefact. Of the two matched
options:

- `(mean_ann, mean)` — treats each annotation as one observation on both sides.
  Preferred as the primary: it mirrors how the pathologist's own numbers are
  produced (a per-region score, not a cell census), and one very large annotation
  cannot dominate the patient's value.
- `(pooled, wmean_cells)` — treats the patient's tissue as a single specimen and
  asks what fraction of its cells are neoplastic. Report alongside as the
  sensitivity analysis. It is the more natural quantity if the annotations are
  arbitrary subdivisions of one tumour rather than distinct regions.

If those two disagree materially, the disagreement is itself the finding: it means
the patient's annotations are heterogeneous in both size and tumour content, and no
single patient-level number represents that slide well.

## Immune infiltrate — IHC vs clinical immune phenotype

Per patient, over the **union** inside cells:

```
cd45_over_inside    = n_cd45_inside    / n_inside   (all CD45+ cells)
frac_<lineage>      = n_<lineage>      / n_inside   (lineage in CD8T, CD4T, Treg, NK)
cd3cd45_over_inside = n_cd3cd45_inside / n_inside   (marker CD3+CD45+ T cells)
gzmb_nk_over_inside = n_gzmb_nk_inside / n_inside   (GZMB+ NK cells)
```

`cd3cd45` (CD45+ AND CD3+) and `gzmb_nk` (NK lineage AND GZMB+) are **marker-gated**
subsets — defined by `<marker>_sign`, not the phenotype label — so they sit
alongside the phenotype lineages as extra readouts of overall immune (CD45+),
T-cell (CD3+CD45+) and cytotoxic-NK (GZMB+ NK) content.

Clinical references: `HOT score` (continuous) and `Immuno-phenotype` (categorical),
joined on `slide_key(`ID PATIENT`)`.

```{r infiltrate-join}
clin <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`)) |>
  select(patient_id,
         hot_score  = any_of("HOT score"),
         immuno_phe = any_of("Immuno-phenotype"))

infiltrate <- ihc_tumor_union |>
  select(patient_id, cd45_over_inside, frac_CD8T, frac_CD4T, frac_Treg, frac_NK,
         cd3cd45_over_inside, gzmb_nk_over_inside) |>
  inner_join(clin, by = "patient_id")

cat("patients with IHC infiltrate + clinical immune phenotype:", nrow(infiltrate), "\n")
```

Statistic: Spearman correlation between `cd45_over_inside` and the numeric `HOT score`.

```{r infiltrate-hotscore}
if ("hot_score" %in% names(infiltrate) &&
    sum(is.finite(suppressWarnings(as.numeric(infiltrate$hot_score)))) > 2) {
  df <- infiltrate |> mutate(hot_score = suppressWarnings(as.numeric(hot_score)))
  st <- paired_spearman(df$cd45_over_inside, df$hot_score)
  ggplot(df, aes(cd45_over_inside, hot_score)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = FIT_LINE, formula = y ~ x) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
             label = sprintf("Spearman rho = %.2f (p = %.3g, n = %d)",
                             st$rho, st$p, st$n)) +
    labs(title = "Immune infiltrate vs clinical HOT score",
         x = "IHC CD45+ fraction inside tumour region", y = "clinical HOT score")
}
```

Grouped by `Immuno-phenotype` (ordered cold -> intermediate -> hot); each box is
the distribution of the IHC fraction within a phenotype group. Metrics now include
the marker-gated CD45+, CD3+CD45+ and GZMB+ NK readouts.

```{r infiltrate-phenotype}
# The metrics carried into every immuno-phenotype comparison (all are 0..1
# fractions of the cells inside the tumour region, so the arcsin panel below can
# transform them all).
infiltrate_metrics <- c("cd45_over_inside", "cd3cd45_over_inside", "gzmb_nk_over_inside",
                        "frac_CD8T", "frac_CD4T", "frac_Treg", "frac_NK")

# One boxplot-by-phenotype figure; `transform` is applied to the fraction so the
# same code draws the raw and the arcsin-sqrt panels (see next chunk).
plot_infiltrate_phenotype <- function(infiltrate, ylab, transform = identity) {
  if (!("immuno_phe" %in% names(infiltrate)) ||
      dplyr::n_distinct(na.omit(infiltrate$immuno_phe)) <= 1) return(invisible())
  df <- infiltrate |>
    pivot_longer(all_of(infiltrate_metrics), names_to = "ihc_metric", values_to = "frac") |>
    filter(!is.na(immuno_phe), is.finite(frac)) |>
    mutate(immuno_phe = hotcold_order(immuno_phe), value = transform(frac)) |>
    filter(is.finite(value))
  ggplot(df, aes(immuno_phe, value, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    facet_wrap(~ ihc_metric, scales = "free_y") +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        name = "Immuno-phenotype") +
    labs(title = "IHC infiltrate metrics by clinical immuno-phenotype",
         x = NULL, y = ylab, colour = "Immuno-phenotype") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

plot_infiltrate_phenotype(infiltrate, "fraction inside tumour region")
```

**Arcsin-transformed.** The same metrics under the arcsine square-root transform
(`asin(sqrt(fraction))`), which stabilises the variance of a proportion by
stretching the crowded 0/1 tails — read the boxes as the transformed scale, not raw
fractions.

```{r infiltrate-phenotype-arcsin}
plot_infiltrate_phenotype(infiltrate, "arcsin(sqrt(fraction))", transform = asin_sqrt)
```

### Composition under three denominators

Same population counts, three denominators (each `NA` when its denominator is 0).
Alongside the phenotype lineages, the marker-gated **CD45+**, **CD3+ CD45+** and
**GZMB+ NK** populations are included:

```
frac_inside = n_<population> / n_inside         (of all cells inside)
frac_tumor  = n_<population> / n_tumor_inside   (of tumour cells inside)
frac_cd45   = n_<population> / n_cd45_inside    (of CD45+ cells inside)

populations = CD8T, CD4T, Treg, NK, Immune_other        (phenotype lineages)
            + CD45+, CD3+ CD45+, GZMB+ NK               (marker-gated subsets)
```

Note `frac_cd45` for the CD45+ population is 1 by construction (a self-normalisation
check), and marker-gated subsets are shown under all three denominators for
consistency with the lineages.

```{r infiltrate-composition, fig.height=5}
# Marker-gated populations to add to the phenotype lineages (label -> count column
# from region_ratios). CD45+ = all immune; CD3+CD45+ = marker T cells; GZMB+ NK.
comp_markers <- c("CD45+"      = "n_cd45_inside",
                  "CD3+ CD45+" = "n_cd3cd45_inside",
                  "GZMB+ NK"   = "n_gzmb_nk_inside")

# One composition figure (population x denominator); `transform` applied to the
# fraction so the same code draws the raw and the arcsin-sqrt panels.
plot_composition <- function(region_metrics, ylab, caption, transform = identity) {
  comp <- region_composition(region_metrics, markers = comp_markers) |>
    pivot_longer(c(frac_inside, frac_tumor, frac_cd45),
                 names_to = "normalization", values_to = "frac") |>
    mutate(normalization = factor(
             c(frac_inside = "of all cells", frac_tumor = "of tumour cells",
               frac_cd45 = "of CD45+ cells")[normalization],
             levels = c("of all cells", "of tumour cells", "of CD45+ cells")),
           value = transform(frac)) |>
    filter(is.finite(value))
  if (nrow(comp) == 0) return(invisible())
  ggplot(comp, aes(lineage, value, fill = lineage)) +
    geom_boxplot(outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = .15, alpha = .5, size = 1) +
    facet_wrap(~ normalization, scales = "free_y") +
    scale_fill_manual(values = oi, guide = "none") +
    labs(title = "IHC immune composition inside the tumour region",
         x = NULL, y = ylab, caption = caption) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

plot_composition(ihc_tumor_union, "fraction",
                 "denominators: all cells / tumour cells / CD45+ cells inside")
```

**Arcsin-transformed.** Same composition under the arcsine square-root transform
of each fraction.

```{r infiltrate-composition-arcsin, fig.height=5}
plot_composition(ihc_tumor_union, "arcsin(sqrt(fraction))",
                 "arcsin-sqrt of: all cells / tumour cells / CD45+ cells inside",
                 transform = asin_sqrt)
```

```{r immune-scope, echo=FALSE, results="asis"}
# Heading and scope sentence come from the parent report: the geojson variant reads
# every imaged cell, the flag variants the deduped in-annotation union.
cat("## Biological-validity QC — ", IMMUNE_HEADING, "\n\n", IMMUNE_PROSE, "\n", sep = "")
```

Per patient p, over that patient's cells in scope:

```
lineage fraction:  frac(p, lineage) = (cells of p with that lineage) / (all cells of p)
                                       [ihc_lineage_fraction]

marker pos-fraction: <marker>_posfrac(p) =
     (cells of p positive for <marker>) / (all in-scope cells of p)
                                       [ihc_marker_fraction; a cell no gate
                                        evaluated counts as not-positive]
```

```{r whole-slide-immune}
ihc_lin <- ihc_lineage_fraction(mem$cells) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, frac) |>
  pivot_wider(names_from = lineage, values_from = frac, values_fill = 0)

ihc_mk <- ihc_marker_fraction(mem$cells) |>
  select(patient_id, any_of(c("CD45_posfrac", "CD8_posfrac", "FOXP3_posfrac")))

ihc_immune <- full_join(ihc_lin, ihc_mk, by = "patient_id")

clin_full <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`))

ihc_clin <- ihc_immune |> inner_join(clin_full, by = "patient_id")
cat("patients with IHC immune metrics + a clinical row:", nrow(ihc_clin), "\n")

# ihc_immune_score = CD45_posfrac if present, else the CD8T lineage fraction.
score_col <- intersect(c("CD45_posfrac", "CD8T"), names(ihc_clin))[1]
ihc_clin$ihc_immune_score <- if (!is.na(score_col)) ihc_clin[[score_col]] else NA_real_
```

### Tissue vs peripheral-blood immunity

```
NLR(p) = BASAL VALUE OF NEUTROPHILS / BASAL VALUE OF LYMPHOCYTES
```

compared against `ihc_immune_score` by Spearman correlation (also for lymphocytes
and PNI).

```{r blood-immunity}
blood <- ihc_clin |>
  transmute(
    patient_id, ihc_immune_score,
    lymphocytes = suppressWarnings(as.numeric(`BASAL VALUE OF LYMPHOCYTES`)),
    neutrophils = suppressWarnings(as.numeric(`BASAL VALUE OF NEUTROPHILS`)),
    pni         = suppressWarnings(as.numeric(`PROGNOSTIC NUTRITIONAL INDEX`))
  ) |>
  mutate(nlr = neutrophils / lymphocytes)

blood_long <- blood |>
  pivot_longer(c(lymphocytes, nlr, pni), names_to = "blood_metric", values_to = "value") |>
  filter(is.finite(ihc_immune_score), is.finite(value))

if (nrow(blood_long) > 2) {
  labs_tbl <- blood_long |>
    group_by(blood_metric) |>
    group_modify(~ paired_spearman(.x$ihc_immune_score, .x$value)) |>
    mutate(lab = sprintf("rho = %.2f (p = %.3g, n = %d)", rho, p, n))
  ggplot(blood_long, aes(ihc_immune_score, value)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = FIT_LINE, formula = y ~ x) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.4, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ blood_metric, scales = "free_y") +
    labs(title = "IHC immune score vs peripheral-blood immunity",
         x = "IHC immune score (CD45+ fraction)", y = NULL)
}
```

### Induction-treatment response

`ihc_immune_score` distribution grouped by `RESPONSE` (factor ordered by ascending
class value, 1 -> 3); no summary statistic beyond the group boxplots.

```{r response}
if ("RESPONSE" %in% names(ihc_clin)) {
  df <- ihc_clin |>
    filter(!is.na(RESPONSE), is.finite(ihc_immune_score))
  df$RESPONSE <- factor(as.character(df$RESPONSE),
                        levels = as.character(sort(unique(as.numeric(as.character(df$RESPONSE))))))
  if (dplyr::n_distinct(df$RESPONSE) > 1 && nrow(df) > 3) {
    ggplot(df, aes(RESPONSE, ihc_immune_score, colour = RESPONSE)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.6) +
      labs(title = "IHC immune score by induction response",
           x = NULL, y = "IHC immune score (CD45+ fraction)") +
      theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
  }
}
```

### Survival

```
group(p) = "high" if ihc_immune_score(p) >= median, else "low"   -> Kaplan-Meier
HR       = exp( Cox coefficient of ihc_immune_score )            (univariable coxph)
```

Event coding (`coerce_event`): yes/dead/relapse/1 -> 1; no/alive/censored/0 -> 0.
Time = `OS`/`PFS` (months); event = `FOLLOW UP` (death) / `PERSISTENCE OR RELAPSE?`
(progression).

```{r survival}
surv_cfg <- list(
  OS  = list(time = "OS",  event = "FOLLOW UP"),
  PFS = list(time = "PFS", event = "PERSISTENCE OR RELAPSE?")
)

coerce_event <- function(x) {
  if (is.numeric(x)) return(as.integer(x > 0))
  v <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    v %in% c("1","yes","y","dead","death","deceased","event","relapse",
             "progression","progressed","true") ~ 1L,
    v %in% c("0","no","n","alive","ned","censored","false") ~ 0L,
    TRUE ~ NA_integer_
  )
}

run_survival <- function(cfg, label) {
  if (!requireNamespace("survival", quietly = TRUE)) {
    message("survival package not installed — skipping ", label); return(invisible())
  }
  if (!all(c(cfg$time, cfg$event) %in% names(ihc_clin))) {
    message("missing ", cfg$time, "/", cfg$event, " — skipping ", label); return(invisible())
  }
  d <- ihc_clin |>
    transmute(ihc_immune_score,
              .time  = suppressWarnings(as.numeric(.data[[cfg$time]])),
              .event = coerce_event(.data[[cfg$event]])) |>
    filter(is.finite(.time), !is.na(.event), is.finite(ihc_immune_score))
  if (nrow(d) < 5 || dplyr::n_distinct(d$.event) < 2) {
    message("too few informative rows for ", label, " (n = ", nrow(d), ")"); return(invisible())
  }
  d <- d |> mutate(immune_group = if_else(ihc_immune_score >= median(ihc_immune_score),
                                          "IHC-immune high", "IHC-immune low"))
  fit  <- survival::survfit(survival::Surv(.time, .event) ~ immune_group, data = d)
  cox  <- survival::coxph(survival::Surv(.time, .event) ~ ihc_immune_score, data = d)
  hr   <- exp(stats::coef(cox))[1]; pcox <- summary(cox)$coefficients[1, "Pr(>|z|)"]
  cat(sprintf("%s: Cox HR per unit IHC immune score = %.2f (p = %.3g, n = %d)\n",
              label, hr, pcox, nrow(d)))
  if (requireNamespace("survminer", quietly = TRUE)) {
    # survminer defaults to its own theme_survminer (grey grid, large type), which
    # would make these the only figures on the site off-style — force the house
    # theme and the house categorical palette onto it.
    print(survminer::ggsurvplot(fit, data = d, pval = TRUE, risk.table = TRUE,
                                title = paste(label, "by IHC immune score (median split)"),
                                legend.title = "", palette = unname(oi),
                                ggtheme = theme_paper())$plot)
  } else {
    plot(fit, col = c(oi[2], oi[1]), xlab = "time", ylab = "survival",
         main = paste(label, "by IHC immune score"))
    legend("topright", levels(factor(d$immune_group)),
           col = c(oi[2], oi[1]), lty = 1, bty = "n")
  }
}

invisible(purrr::iwalk(surv_cfg, ~ run_survival(.x, .y)))
```

## Notes

```{r notes, echo=FALSE, results="asis"}
# MEMBERSHIP_NOTES: the bullets that only hold for this variant (parent report).
cat(paste0("- ", MEMBERSHIP_NOTES, collapse = "\n"), "\n", sep = "")
```
- Marker fractions count a cell no gate evaluated for that marker as not-positive,
  not as missing, so a sparsely-gated marker reads low rather than dropping out.
  `marker_gated()` (see `code/cell_tables.R`) is what separates the two.
- Cohorts are small; every correlation is Spearman with its `n` reported. Read
  associations as descriptive.

```{r export-pdfs, include=FALSE}
# Copy every plot from this report (each rendered as a PDF via the
# dev = c("png", "pdf") chunk option) into output/figures/<SLUG>/. Runs last so that
# every figure above already exists on disk.
export_pdf_figures(SLUG)
```
