---
title: "Clinical concordance — Massimo1 inverted — modified PANCK threshold"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(forcats)
  library(here)
  library(sf)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the analysis script that used to pull plot_theme.R in never runs —
# which silently left those plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, the per-arm neoplastic_* / ihc_* objects, counts_data)
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R pulls in
# cell_tables.R (export-schema adapter) and plot_theme.R (house figure style).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "membership.R"))           # membership_data(): the one variant knob
source(here("code", "aggregation_compare.R"))  # annotation-aggregation sensitivity grid
source(here("code", "pdf_export.R"))           # export_pdf_figures(): plots -> output/figures/

SLUG <- "clinical_massimo1_inverted"

# WHICH ARM THIS PAGE IS ABOUT. Three arms phenotype the same slides three ways
# (code/arms.R), and their regions are drawn INDEPENDENTLY of one another — so the
# cells, the polygons AND the pathologist table all have to come from the same arm
# or the page correlates one arm's tumour content against another arm's regions.
# There is no default anywhere in the loaders precisely so this constant cannot be
# left unset: change it here and the whole page follows.
ARM        <- "massimo1_inverted"
ihc_data   <- ihc_for(ARM)
neoplastic <- neoplastic_for(ARM)
```

> **Arm: `massimo1_inverted`.** Massimo1's OWN regions and polygons, with the cells re-classified at a modified PANCK threshold. Covers 052 and 5456 only, and shares arm 1's pathologist scores — the tissue is the same, only the classifier differs.
>
> This page is one of three, identical in construction and differing only in which
> arm it reads. The arms' regions were drawn in **separate annotation sessions**, so
> `ANNOTATION_2` here is not `ANNOTATION_2` on the other pages — compare them by
> PATIENT, never region by region. `code/arms.R` states the rule in full.

```{r body, results = "asis", echo = FALSE}
# THE BODY IS SPLICED WITH knit_child(text = ...), NOT THE `child =` CHUNK OPTION.
#
# That is not a style choice. workflowr's fig.path option hook does
#   options$fig.path <- create_figure_path(knitr::current_input())
# and inside a `child =` chunk knitr::current_input() returns the CHILD. So every
# arm's figures would land in figure/clinical_body/ — ONE directory shared by all
# three parents — and whichever page knitted last would silently overwrite the
# others. export_pdf_figures(SLUG) then copies those same overwritten files into
# all three output/figures/<slug>/ directories, so a panel labelled massimo1 would
# show massimo2's cells with nothing to indicate it.
#
# knit_child(text = readLines(...)) has no input file of its own, so
# current_input() stays the PARENT: figures go to figure/<parent>/ as they should,
# and workflowr's "custom fig.path was ignored" warning disappears too — that
# warning was the visible symptom of the collision, not the problem itself.
# tests/testthat/test-child-documents.R pins this idiom.
cat(knitr::knit_child(
      text  = readLines(here::here("analysis", "_children", "clinical_body.Rmd")),
      quiet = TRUE),
    sep = "\n")
```

```{r export-pdfs, include=FALSE}
# Copy every plot from this report (each rendered as a PDF via the
# dev = c("png", "pdf") chunk option) into output/figures/<SLUG>/. Runs last so that
# every figure above already exists on disk.
export_pdf_figures(SLUG)
```
