---
title: "Marker QC — channel usability, phenotype agreement and lineage leakage"
author: "sceriff0"
date: "2026-08-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (in the lockfile). Optional: diptest for Hartigan's dip
# test (a bimodality-coefficient fallback runs if it is absent).
#   renv::install("diptest"); renv::snapshot()
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(forcats)
  library(here)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the analysis script that used to pull plot_theme.R in never runs —
# which silently left those plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

SLUG <- "marker_qc"
CAP  <- "ihc_method internal QC · FlowPath marker & lineage QC"
```

Internal QC of the FlowPath gating, with **no external reference**: every check
below asks whether the marker calls are self-consistent, not whether they match
bulk RNA, deconvolution or the clinical record. Those comparisons live on
[the molecular page](molecular_hot_cold.html); a disagreement there is only
interpretable once the checks here pass.

Two parts, both reading the whole-slide `ihc_data`:

| part | asks | fails when |
|---|---|---|
| 1 | is each channel usable, and does it agree with the phenotype it defines? | a channel is unimodal, unused, or a phenotype's defining marker is not enriched in it |
| 2 | do the lineages stay exclusive, or do markers leak across them? | CD45 fires inside tumour phenotypes, PANCK inside immune ones, or a marker pair co-fires above chance |

`CAP` is the shared figure caption; both parts stamp it on every plot so an exported
PDF still says which report it came from.

## Part 1 — Channel usability and phenotype agreement


### Starting data and notation

Every value below is derived from a single input, with **no external reference**:

- `ihc_data` — FlowPath single cells, deduplicated to **one copy per patient**
  (whole slide). Each cell carries a `phenotype` label and, for each marker, a
  value / z-score / sign triple — under whichever of the three export spellings the
  file uses (`code/cell_tables.R`).

Notation used throughout. `ihc_marker_long(ihc_data)` reshapes the whole set of
imaged cells to **one row per (cell, marker) pair**, over the markers in
`ihc_markers` (= `marker_gene_map$marker`, excluding `wrongL1CAM`/DAPI):

```
raw    = marker_value(cell, marker)     the intensity the export carries
zscore = marker_zscore(cell, marker)
sign   = marker_sign(cell, marker)      verbatim, for display
gated  = marker_gated(cell, marker)     FALSE = a gate never evaluated this marker
                                        on this cell — NOT the same as negative
pos    = marker_pos(cell, marker)       TRUE only for an explicitly positive sign
```

`gated` is the distinction the sign string alone hides: FlowPath leaves an ungated
column blank and mirage writes `·`, both of which are "never measured", not
"negative". Panels that would otherwise count an unmeasured cell as a negative one
filter on `gated` first.

`phenotype_clean` is the cell type. All patient ids are mapped to the shared slide
space by `slide_key()`. The report restricts to **typed** cells: the cell type is
not NA and not "Unknown".

```{r marker-long}
mk_long <- ihc_marker_long(ihc_data) |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown")
cat("cells:", dplyr::n_distinct(mk_long$cell_id),
    "| markers:", dplyr::n_distinct(mk_long$marker),
    "| phenotypes:", dplyr::n_distinct(mk_long$phenotype_clean), "\n")
```

### Phenotype ↔ marker concordance

For each phenotype and marker, over the typed cells of that phenotype:

```
mean_z(type, marker)  = average of <marker>_zscore over cells of that type
pct_pos(type, marker) = fraction of those cells positive for the marker
```

where:
- both averages use na.rm = TRUE (a positive cell counts as 1, negative as 0).

The heatmap fills each (marker, type) tile with `mean_z` (diverging scale centred
at 0).

```{r concordance-matrix}
conc <- mk_long |>
  group_by(phenotype_clean, marker) |>
  summarise(mean_z = mean(zscore, na.rm = TRUE),
            pct_pos = mean(pos,  na.rm = TRUE), .groups = "drop")

ggplot(conc, aes(marker, phenotype_clean, fill = mean_z)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_div(midpoint = 0, name = "mean z", guide = guide_cbar()) +
  labs(title = "Phenotype vs marker: mean z-score",
       subtitle = "Expect each phenotype to peak on its defining marker(s); off-diagonal signal = mislabelled cells.",
       # Both axes are categorical and their ticks name the variable already
       # (marker symbols, phenotype labels), so a title would only repeat them.
       x = NULL, y = NULL, caption = CAP) +
  theme_paper_tile() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Defining-marker check.** `expected` is a fixed table giving the defining marker
for each phenotype. Within each phenotype the markers are ranked by descending
mean z-score:

```
rank_z(type, marker) = rank of the marker's mean_z within the phenotype
                       (rank 1 = largest mean_z)
rank_of_defining     = rank_z of that phenotype's defining marker
pass                 = rank_of_defining <= 2
```

where:
- the defining-marker row is kept by joining on the `expected` table;
- pass means the defining marker is among the phenotype's top-2 mean-z signals.

`mean_z` is rounded to 2 dp; rows are sorted by descending `rank_of_defining`.

```{r concordance-expected}
# Encode the expected defining marker(s) per phenotype and check it is the top
# (or near-top) signal. Flags phenotypes whose defining marker is not their max.
expected <- tibble::tribble(
  ~phenotype_clean,           ~marker,
  "PANCK+Tumor",              "PANCK",
  "VIM+Tumor",                "VIMENTIN",
  "T helper",                 "CD4",
  "T cytotoxic",              "CD8",
  "Activated T cytotoxic",    "CD8",
  "CD4+ Treg",                "FOXP3",
  "CD8+ T reg",               "FOXP3",
  "Natural Killer",           "CD56",
  "Activated Natural Killer", "CD56",
  "Immune",                   "CD45",
  "Stroma",                   "SMA"
)

check <- conc |>
  group_by(phenotype_clean) |>
  mutate(rank_z = rank(-mean_z)) |>
  inner_join(expected, by = c("phenotype_clean", "marker")) |>
  transmute(phenotype_clean, defining_marker = marker,
            mean_z = round(mean_z, 2), rank_of_defining = as.integer(rank_z),
            pass = rank_z <= 2) |>
  arrange(desc(rank_of_defining))
knitr::kable(check, caption = "Is each phenotype's defining marker among its top-2 signals?")
```

### Per-channel usability

Per marker, over all cells with a finite z-score (typed and untyped —
`ihc_marker_long(ihc_data)` is re-derived here without the phenotype filter).

**Bimodality.** Sarle's bimodality coefficient on the marker's z-scores:

```
BC = (skew^2 + 1) / (kurtosis_excess + 3*(n-1)^2 / ((n-2)*(n-3)))
```

where:
- skew = skewness of the finite z-scores (mean of the cubed standardised values);
- kurtosis_excess = excess kurtosis of those z-scores (mean of the 4th-power
  standardised values, minus 3);
- n = number of finite z-scores;
- BC is NA if n < 4 or the z-score sd is 0.

`dip_p` is Hartigan's dip-test p-value (`diptest::dip.test`) when the package is
installed and n >= 4, else NA.

**Sign-vs-zscore separability.** With the positive/negative split from
`pos(cell, marker)`:

```
neg_max        = largest z-score among the marker's sign-negative cells
pos_min        = smallest z-score among the marker's sign-positive cells
separable      = pos_min > neg_max
cutoff         = (neg_max + pos_min) / 2 when separable, else NA
likely_bimodal = BC > 0.555 OR dip_p < 0.05
```

Rows are sorted by descending BC.

```{r channel-qc}
bimodality_coef <- function(x) {                       # Sarle's BC
  x <- x[is.finite(x)]; n <- length(x)
  if (n < 4) return(NA_real_)
  m <- mean(x); s <- sd(x); if (s == 0) return(NA_real_)
  g <- mean(((x - m) / s)^3)                           # skewness
  k <- mean(((x - m) / s)^4) - 3                        # excess kurtosis
  (g^2 + 1) / (k + 3 * (n - 1)^2 / ((n - 2) * (n - 3)))
}
dip_p <- function(x) {
  x <- x[is.finite(x)]
  if (!requireNamespace("diptest", quietly = TRUE) || length(x) < 4) return(NA_real_)
  suppressWarnings(diptest::dip.test(x)$p.value)
}

channel_qc <- ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore)) |>
  group_by(marker) |>
  summarise(
    n        = n(),
    bc       = bimodality_coef(zscore),
    dip_p    = dip_p(zscore),
    neg_max  = suppressWarnings(max(zscore[!pos], na.rm = TRUE)),
    pos_min  = suppressWarnings(min(zscore[pos],  na.rm = TRUE)),
    .groups  = "drop"
  ) |>
  mutate(
    separable = is.finite(neg_max) & is.finite(pos_min) & pos_min > neg_max,
    cutoff    = if_else(separable, (neg_max + pos_min) / 2, NA_real_),
    likely_bimodal = bc > 0.555 | (!is.na(dip_p) & dip_p < 0.05)
  ) |>
  arrange(desc(bc))
knitr::kable(channel_qc, digits = 3,
             caption = "Per-channel bimodality and sign-vs-zscore separability")
```

Density of the z-scores per marker, over cells with a finite z-score that a gate
actually evaluated (`gated`), split by the positive call (`sign +` / `sign -`);
facets are free-scaled.

```{r channel-qc-plot, fig.height=8}
ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore), gated) |>
  mutate(call = if_else(pos, "sign +", "sign -")) |>
  ggplot(aes(zscore, fill = call)) +
    geom_density(alpha = 0.55, colour = NA) +
    facet_wrap(~ marker, scales = "free") +
    scale_fill_manual(values = c("sign +" = oi[1], "sign -" = oi[2]), name = NULL) +
    labs(title = "Marker z-score by sign call",
         subtitle = "Clean channels show two separated humps; failed channels overlap or are unimodal.",
         x = "Marker intensity (z-score)", y = "Density (unitless)", caption = CAP)
```

### Morphology by phenotype

Morphology features are whichever of `area`, `perimeter`, `eccentricity`,
`solidity` are present in `ihc_data`, plus circularity when both `area` and
`perimeter` exist:

```
circularity = 4 * pi * area / perimeter^2
```

Over typed cells, each feature is pivoted long and filtered to finite values; the
plot is a per-phenotype boxplot (phenotypes ordered by median feature value,
outliers hidden, facets free-scaled).

```{r morphology}
morph_feats <- intersect(c("area", "perimeter", "eccentricity", "solidity"),
                         names(ihc_data))

morph <- ihc_data |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown") |>
  mutate(circularity = if (all(c("area", "perimeter") %in% names(ihc_data)))
           4 * pi * area / perimeter^2 else NA_real_) |>
  select(phenotype_clean, any_of(c(morph_feats, "circularity"))) |>
  pivot_longer(-phenotype_clean, names_to = "feature", values_to = "value") |>
  filter(is.finite(value))

ggplot(morph, aes(reorder(phenotype_clean, value, FUN = median), value,
                  fill = phenotype_clean)) +
  geom_boxplot(outlier.shape = NA, width = .6, colour = "grey30") +
  facet_wrap(~ feature, scales = "free") +
  coord_flip() +
  scale_fill_ordinal(guide = "none") +
  labs(title = "Cell morphology by phenotype",
       subtitle = "Shapes should track cell type: tumour largest, lymphocytes small/round, stroma elongated.",
       x = NULL, y = NULL, caption = CAP)
```

### Notes

- All three checks use only `ihc_data`; no external reference is joined.
- `mean` reductions carry `na.rm = TRUE`; `neg_max`/`pos_min` return non-finite
  (`±Inf`) when a marker has no negative (resp. positive) cell, which drives
  `separable = FALSE` via the `is.finite` guards.
- `separable = FALSE` means the sign call is not a pure threshold on the z-score (the
  `_sign+` and `_sign−` z-score ranges overlap — likely thresholded on `_raw`);
  its `cutoff` is then `NA`.
- `likely_bimodal` uses an inclusive-or of BC > 0.555 and dip-p < 0.05; if
  `diptest` is absent the dip-p term is `NA` and only BC decides.


## Part 2 — Lineage leakage and marker exclusivity


### Starting data and notation

Single input: `ihc_data`, the FlowPath single cells deduplicated to **one copy per
patient** (whole slide). "Patient p's cells" means every row of `ihc_data` for that
patient; "the cohort" means all cells pooled across patients. All patient ids are
mapped to slide space by `slide_key()` (strip non-digits, apply the typo fix
`15879` -> `15897`).

Cell labels:

```
marker+ cell = marker_pos(cell, marker) — the marker's sign column, whichever of
               <marker>_sign / sign:<marker> / <marker>_positive the export carries
tumour-phenotype cell = phenotype_clean contains "Tumor"
                        (phenotype_clean = the text inside the parentheses of `phenotype`)
immune-phenotype cell = phenotype_clean is one of the literal immune labels below
```

The immune-phenotype labels are: `Immune`, `T helper`, `T cytotoxic`,
`Activated T cytotoxic`, `CD4+ Treg`, `CD8+ T reg`, `Natural Killer`,
`Activated Natural Killer` (defined in the chunk below; excludes Stroma / Unknown /
Tumor). These two phenotype sets are read directly from `phenotype_clean`, **not**
through the `lineage()` collapse. Missing `phenotype_clean` is treated as the empty
string, so it matches neither set.

### Phenotype-level leakage

Per patient p, the fraction of one lineage's phenotyped cells that call the
opposite lineage's marker:

```
cd45pos_in_tumorph(p) = (tumour-phenotype cells of p that are CD45+) / (tumour-phenotype cells of p)
panckpos_in_immune(p) = (immune-phenotype cells of p that are PANCK+) / (immune-phenotype cells of p)

n_cells(p) = all cells of p
where:
  tumour-phenotype cell = phenotype_clean contains "Tumor"
  immune-phenotype cell = phenotype_clean is one of the literal immune labels
  CD45+  = marker_pos(cell, "CD45")
  PANCK+ = marker_pos(cell, "PANCK")
```

Each ratio is `NA` when its patient has no cell in the corresponding phenotype set
(`any(...)` guard); positivity is never `NA`.

```{r leakage}
immune_ph <- c("Immune", "T helper", "T cytotoxic", "Activated T cytotoxic",
               "CD4+ Treg", "CD8+ T reg", "Natural Killer", "Activated Natural Killer")

excl <- ihc_data |>
  mutate(patient_id  = slide_key(patient_id),
         is_tumor_ph = str_detect(replace_na(phenotype_clean, ""), "Tumor"),
         is_immune_ph = phenotype_clean %in% immune_ph,
         panck_pos = marker_pos(ihc_data, "PANCK"),
         cd45_pos  = marker_pos(ihc_data, "CD45"))

spill <- excl |>
  group_by(patient_id) |>
  summarise(
    n_cells            = n(),
    cd45pos_in_tumorph = if (any(is_tumor_ph))  mean(cd45_pos[is_tumor_ph],  na.rm = TRUE) else NA_real_,
    panckpos_in_immune = if (any(is_immune_ph)) mean(panck_pos[is_immune_ph], na.rm = TRUE) else NA_real_,
    .groups = "drop"
  )
knitr::kable(spill, digits = 3,
             caption = "Per-patient phenotype-level leakage (want ~0)")
```

Bars are the two ratios above, per patient, dropping `NA` entries.

```{r leakage-plot, fig.height=4.5}
spill |>
  pivot_longer(c(cd45pos_in_tumorph, panckpos_in_immune),
               names_to = "leak", values_to = "rate") |>
  filter(!is.na(rate)) |>
  mutate(leak = c(cd45pos_in_tumorph = "CD45+ among tumour-typed",
                  panckpos_in_immune = "PANCK+ among immune-typed")[leak]) |>
  ggplot(aes(reorder(patient_id, rate), rate, fill = leak)) +
    geom_col(position = position_dodge(width = .7), width = .65) +
    coord_flip() +
    scale_fill_manual(values = oi[c(1, 2)], name = NULL) +
    labs(title = "Phenotype-level leakage per patient",
         subtitle = "Fraction of a lineage's cells calling the opposite lineage's marker — lower is better.",
         x = "Patient", y = "Fraction of the lineage's cells (unitless, 0-1)",
         caption = CAP)
```

### Marker-pair exclusivity panel

`sign_pos` is the per-cell positivity matrix: one logical column per marker in
`ihc_markers`, obtained by applying `is_pos` to every `<marker>_sign` column and
stripping the `_sign` suffix. Each row is a cell of the cohort; `patient_id` is
`slide_key`-normalised.

For a candidate pair of markers (a, b), retained only if both columns exist in
`sign_pos`, all counts are **pooled over the whole cohort**:

```
n_a    = cohort cells that are a+
n_b    = cohort cells that are b+
n_both = cohort cells that are both a+ and b+   (double-positive)

pct_dp_all  = n_both / (all cohort cells)
p(b | a)    = n_both / n_a       (share of a+ cells that are also b+)
p(a | b)    = n_both / n_b       (share of b+ cells that are also a+)
where:
  marker+ cell = marker_pos(cell, marker)
```

The two conditionals are set to `NA` when their denominator (`n_a` or `n_b`) is 0.
`pct_dp_all` uses the all-cell denominator; the conditionals use the single-marker
positive counts.

```{r exclusivity-pairs}
# Per-cell marker positivity matrix: one logical column per marker, named for the
# marker rather than for whichever sign column the export used. A marker the panel
# does not carry comes back all-FALSE, so its pairs report 0% co-positivity instead
# of aborting the chunk.
sign_pos <- marker_matrix(ihc_data, ihc_markers) |>
  mutate(patient_id = slide_key(ihc_data$patient_id), .before = 1)

# Literature-backed mutually-exclusive lineage marker pairs (+ known exception).
exclusivity_pairs <- tibble::tribble(
  ~a,      ~b,      ~pair_label,                              ~expected_copos, ~reference,
  "PANCK", "CD45",  "epithelial/tumour vs leukocyte",         "~0 (rare hybrid CTCs)",     "Nandedkar 1998",
  "CD4",   "CD8",   "helper vs cytotoxic T",                  "~3% peripheral DP-T",       "Overgaard 2023 (Biomedicines)",
  "CD3",   "CD56",  "T cell vs NK cell",                      "rare CD3+CD56+ NKT",        "std NK/NKT gating",
  "CD3",   "CD14",  "lymphoid vs monocyte/macrophage",        "~0 distinct lineages",      "flow-panel convention",
  "CD3",   "CD163", "lymphoid vs M2 macrophage",              "~0 (lymphocytes CD163-)",   "Roberts 2004",
  "CD45",  "SMA",   "leukocyte vs smooth-muscle/myofibroblast","~0 distinct lineages",     "stromal-marker convention"
) |>
  # keep only pairs whose markers are both present
  filter(a %in% names(sign_pos), b %in% names(sign_pos))

pair_stats <- exclusivity_pairs |>
  rowwise() |>
  mutate(
    n_a          = sum(sign_pos[[a]], na.rm = TRUE),
    n_b          = sum(sign_pos[[b]], na.rm = TRUE),
    n_both       = sum(sign_pos[[a]] & sign_pos[[b]], na.rm = TRUE),
    pct_dp_all   = n_both / nrow(sign_pos),
    p_b_given_a  = ifelse(n_a > 0, n_both / n_a, NA_real_),
    p_a_given_b  = ifelse(n_b > 0, n_both / n_b, NA_real_)
  ) |>
  ungroup() |>
  select(pair = pair_label, a, b, expected_copos, reference,
         n_both, pct_dp_all, p_b_given_a, p_a_given_b)

knitr::kable(pair_stats, digits = 4,
             caption = "Mutually-exclusive marker pairs: observed vs expected co-positivity")
```

Worst-direction co-positivity per pair:

```
max_cond = larger of  p(b | a)  and  p(a | b)
```

plotted for the pairs with a finite value.

```{r exclusivity-summary, fig.height=4.5}
# The worst-case conditional co-positivity per pair (max of the two directions).
pair_stats |>
  mutate(max_cond = pmax(p_b_given_a, p_a_given_b, na.rm = TRUE),
         pair = sprintf("%s / %s", a, b)) |>
  filter(is.finite(max_cond)) |>
  ggplot(aes(reorder(pair, max_cond), max_cond)) +
    geom_col(fill = oi[2], width = .65) +
    coord_flip() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(title = "Worst-direction co-positivity per exclusive pair",
         subtitle = "max( P(B+|A+), P(A+|B+) ) — high bars = the pair most likely showing spillover.",
         x = NULL, y = "Conditional co-positivity (unitless, 0-1)", caption = CAP)
```

Per-patient double-positive fraction, one value per (patient, pair), over **all**
of that patient's cells:

```
dp(p, a, b) = (cells of p that are both a+ and b+) / (all cells of p)
```

This denominator is all of patient p's cells, i.e. the per-patient analogue of
`pct_dp_all`, not of the conditional rates above.

```{r exclusivity-heatmap}
per_pat <- exclusivity_pairs |>
  select(a, b, pair_label) |>
  purrr::pmap_dfr(function(a, b, pair_label) {
    sign_pos |>
      group_by(patient_id) |>
      summarise(dp = mean(.data[[a]] & .data[[b]], na.rm = TRUE), .groups = "drop") |>
      mutate(pair = pair_label)
  })

if (nrow(per_pat) > 0) {
  ggplot(per_pat, aes(patient_id, pair, fill = dp)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    scale_fill_seq(labels = scales::percent_format(accuracy = .1),
                   name = "double-pos fraction", guide = guide_cbar()) +
    labs(title = "Marker-pair co-positivity per patient",
         subtitle = "Bright = high double-positive rate = likely spillover in that patient.",
         x = "Patient", y = NULL, caption = CAP) +
    theme_paper_tile() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

Pairs held out of the tribble because their phenotypes are defined to overlap
(co-positivity is expected, not an artifact): PANCK/VIMENTIN (the `VIM+Tumor` EMT
phenotype) and CD8/FOXP3 (the `CD8+ T reg` phenotype).

### Notes

- Single source `ihc_data` (whole slide); every metric is a count ratio, no
  external reference.
- Denominator audit: leakage rates divide by the phenotype-set size within a
  patient's cells; `pct_dp_all` and heatmap `dp` divide by all cells (whole cohort /
  the patient); `p_b_given_a` / `p_a_given_b` divide by the single-marker positive
  counts (`n_a` / `n_b`). The conditional rates and pair counts are pooled
  cohort-wide, whereas leakage and the heatmap are per patient.
- Leakage membership uses `phenotype_clean` text and the literal immune-phenotype
  list, not the `phenotype_lineage` collapse.


```{r export-pdfs, include=FALSE}
# Every plot above is also rendered to PDF by the dev = c("png", "pdf") chunk option;
# this collects them into output/figures/<SLUG>/. Runs last, so they all exist by now.
export_pdf_figures(SLUG)
```
