---
title: "Bulk RNA, deconvolution and the hot/cold axis"
author: "sceriff0"
date: "2026-08-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (in the lockfile), sf + jsonlite for the margin geometry in
# part 3, and immunedeconv (GitHub: omnideconv/immunedeconv, NOT Bioconductor) for
# part 2 — see the method-configuration chunk there for the install recipe. Each part
# degrades to a message if its optional input is absent; none of them is required.
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the analysis script that used to pull plot_theme.R in never runs —
# which silently left those plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the shared
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R pulls
# in cell_tables.R (export-schema adapter) and plot_theme.R (house figure style, which
# is where hotcold_cols()/hotcold_order() come from).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

SLUG <- "molecular_hot_cold"
```

Everything on this page asks one question three ways: **does a bulk, cohort-level
readout agree with what the slide shows, and does that agreement depend on whether
the tumour is immunologically hot or cold?**

| part | bulk readout | compared against | grain |
|---|---|---|---|
| 1 | per-gene DESeq2-normalised expression | IHC marker-positive cell fraction | marker |
| 2 | immunedeconv cell-type scores | IHC lineage fraction | lineage |
| 3 | — (spatial, not bulk) | clinical `HOT score` / immuno-phenotype | invasive margin |

Parts 1 and 2 were separate pages (`bulkRna`, `deconvolution`) and part 3 was a third
(`periphery`). They shared the CRF-to-slide id bridge, the clinical hot/cold lookup and
the whole-slide `ihc_data` reduction, and re-derived all three independently — three
places for the same mapping to drift. They are one page now, and those shared keys are
built **once**, immediately below, by the only chunk allowed to build them.

## Shared keys: the id bridge and the hot/cold call

Two lookups are used by all three parts, so they are defined here and nowhere else.

**The id bridge.** Bulk RNA lives in **CRF-id** space; IHC lives in **slide-id**
space. `slide_key()` maps any id to the shared numeric slide space (strip non-digits,
apply the verified typo fix `15879` -> `15897`); `norm_id()` upper-cases and strips
non-alphanumerics. `crf_to_slide_map(clinical_data)` returns a named vector:

```
crf2slide[crf] = slide
where: crf   = norm_id of `ID CRF PRESERVE`
       slide = slide_key of `ID PATIENT`
```

It is a base-R named-vector lookup, deliberately **not** a join: a join on a
duplicated CRF id silently multiplies rows, and this table has one row per patient.

**The hot/cold call.** `clin` carries the two clinical immune columns, keyed into
slide space so it joins onto every per-patient frame below:

```
hot_score(p)  = numeric `HOT score`        from the clinical CRF
immuno_phe(p) = `Immuno-phenotype`         from the clinical CRF (cold / intermediate / hot)
```

`hot_score` is coerced with `as.numeric()`; a non-numeric entry becomes `NA` rather
than stopping the page. `any_of()` means a clinical export missing either column
yields a `clin` without it instead of an error — the hot/cold colouring then simply
falls back to grey.

```{r shared-keys}
# ONE owner for both lookups. Parts 1-3 all consume these; none of them rebuilds them.
crf2slide <- crf_to_slide_map(clinical_data)

clin <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`)) |>
  select(patient_id,
         hot_score  = any_of("HOT score"),
         immuno_phe = any_of("Immuno-phenotype")) |>
  mutate(hot_score = suppressWarnings(as.numeric(hot_score)))

cat("clinical rows:", nrow(clin),
    "| with a numeric HOT score:", sum(is.finite(clin$hot_score)),
    "| with an immuno-phenotype:", sum(!is.na(clin$immuno_phe)), "\n")
```

## Part 1 — Bulk RNA expression vs IHC marker positivity

Marker by marker, across patients: does a marker's gene expression in bulk RNA track
the fraction of cells the IHC gate calls positive for it? Points are coloured by the
clinical immuno-phenotype so a marker that only agrees in hot tumours is visible.


### Starting data and notation

Every value below is derived from three inputs:

- `ihc_data` — FlowPath single cells, deduplicated to **one copy per patient**
  (whole slide, no annotation boundary). Carries per-marker `<marker>_sign` and
  `<marker>_zscore` columns and `patient_id`.
- `counts_data` — DESeq2-normalised **bulk RNA**, one row per RNA sample, one
  column per gene, plus the CRF-space sample id `Sample`.
- `crf2slide` and `clin` — the id bridge and the hot/cold call, both built once in
  **Shared keys** above and only read here.

Notation and mappings used throughout:

```
cells of a patient p = all of p's cells in `ihc_data`
marker-positive cell (marker <marker>) = marker_pos(cell, <marker>) is TRUE
  (is_pos accepts +, pos, positive, yes, true, 1)
genes of a marker = marker_gene_map[<marker>]: a comma-separated gene list
  (e.g. CD3 -> CD3D, CD3E, CD3G; PANCK -> EPCAM, KRT5, KRT8, KRT14, KRT18, KRT19);
  multi-gene markers are averaged
```

Paired associations are Spearman correlations via `paired_spearman`.

### Per-patient bulk expression of each marker's gene(s)

RNA samples are keyed in CRF space; `crf2slide` bridges them to slide space.
`counts_data` is melted long to one row per (sample, gene), with columns
`Sample`, `crf_id = norm_id(Sample)`, `gene`, and `expr`. `gene_rows` expands
`marker_gene_map` to one row per (marker, gene) over each marker's gene list
(comma-split, trimmed); `marker_genes` keeps only the (marker, gene) rows whose
gene appears in the RNA matrix:

```
available genes of a marker = its genes that are also columns of `counts_data`
```

Per marker and CRF sample, expression is the mean over that marker's available
genes:

```
rna_expr(<marker>, s) = (sum over the marker's available genes of expr(s, gene))
                        / (number of the marker's available genes)
where:
  expr(s, gene) = the DESeq2-normalised value for gene in sample s
```

The CRF id is then mapped to slide space with
`patient_id = crf2slide[crf_id(s)]`, and rows with no crosswalk match (NA) are
dropped.

```{r rna-marker-expression}
expr_long <- counts_data |>
  mutate(crf_id = norm_id(Sample)) |>
  pivot_longer(-c(Sample, crf_id), names_to = "gene", values_to = "expr")

# marker -> its genes, one row per (marker, gene). Explicit $-access + strsplit
# (no NSE) so it is robust to the base pipe / tidyr separate_rows() quirks.
gene_rows <- do.call(rbind, Map(function(m, g)
  data.frame(marker = m, gene = trimws(strsplit(g, ",")[[1]]), stringsAsFactors = FALSE),
  marker_gene_map$marker, marker_gene_map$genes))

missing_genes <- setdiff(gene_rows$gene, unique(expr_long$gene))
if (length(missing_genes)) message("genes absent from RNA matrix: ", paste(missing_genes, collapse = ", "))

# keep only genes present in the RNA matrix
marker_genes <- gene_rows |> filter(gene %in% unique(expr_long$gene))

# Per patient, average expression across a marker's available genes, then map the
# CRF sample id to the IHC slide id via the crosswalk.
marker_expr <- expr_long |>
  inner_join(marker_genes, by = "gene", relationship = "many-to-many") |>
  group_by(marker, crf_id) |>
  summarise(rna_expr = mean(expr), .groups = "drop") |>
  mutate(patient_id = unname(crf2slide[crf_id])) |>   # crf id -> slide id
  filter(!is.na(patient_id)) |>
  select(marker, patient_id, rna_expr)
```

### Per-patient IHC positive-cell fraction

`ihc_marker_fraction(ihc_data)` groups cells by `slide_key(patient_id)` and, for
each marker, returns the whole-slide positive fraction (mean of `is_pos` over
non-NA sign values):

```
<marker>_posfrac(p) = (cells of p positive for <marker>) / (all cells of p)
where:
  positive = marker_pos(cell, marker) — TRUE only for an explicitly positive sign.
             A cell no gate evaluated for this marker (blank / "·" sign, or a marker
             the panel lacks) counts in the denominator as not-positive, so a
             sparsely-gated marker reads LOW rather than dropping out. Check
             marker_gated() before reading a near-zero posfrac as biology.
```

These `_posfrac` columns are reshaped long to `patient_id`, `marker`, and
`ihc_posfrac`.

```{r rna-ihc-fraction}
ihc_marker_frac <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_posfrac")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_posfrac") |>
  mutate(marker = sub("_posfrac$", "", marker))

cat("IHC patients:", dplyr::n_distinct(ihc_marker_frac$patient_id),
    "| RNA patients:", dplyr::n_distinct(marker_expr$patient_id), "\n")
```

### Marker-by-marker concordance

Inner-join `marker_expr` and `ihc_marker_frac` on `marker` and `patient_id`, keeping
finite pairs. Per marker, over its shared patients, the statistic is the Spearman
correlation between `ihc_posfrac` and `rna_expr` across patients (`paired_spearman`,
via `cor.test(method = "spearman")`), reported with its p-value and the pair count
`n`; it is `NA` when `n < 3` or either vector is constant. Rows are sorted by
descending correlation.

```{r rna-concordance}
rna_paired <- marker_expr |>
  inner_join(ihc_marker_frac, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_posfrac))

cat("shared patients (RNA + IHC):",
    dplyr::n_distinct(rna_paired$patient_id), "\n")

rna_summary <- rna_paired |>
  group_by(marker) |>
  group_modify(~ paired_spearman(.x$ihc_posfrac, .x$rna_expr)) |>
  ungroup() |>
  arrange(desc(rho))

knitr::kable(rna_summary, digits = 3,
             caption = "Spearman: IHC positive-cell fraction vs bulk-RNA expression")
```

Scatter per marker: `ihc_posfrac` against `log1p(rna_expr)` (`log1p` keeps zeros),
with an OLS fit line.

```{r rna-concordance-plot, fig.height=8}
if (nrow(rna_paired) > 0) {
  # Colour by the clinical hot/cold call so a marker whose RNA-IHC agreement holds
  # only in one immune stratum is visible as such. left_join, not inner_join: a
  # patient with no immuno-phenotype must stay in the plot (grey) rather than
  # silently shrink the panel the fit is drawn through.
  df <- rna_paired |>
    left_join(clin, by = "patient_id") |>
    mutate(immuno_phe = hotcold_order(immuno_phe))
  # log-scale expression (counts span orders of magnitude); +1 to keep zeros.
  # The fit goes under the points and is over ALL patients, coloured or not.
  ggplot(df, aes(ihc_posfrac, log1p(rna_expr))) +
    geom_smooth(method = "lm", se = FALSE, colour = FIT_LINE, formula = y ~ x) +
    geom_point(aes(colour = immuno_phe), alpha = 0.75, size = 1.8) +
    facet_wrap(~ marker, scales = "free") +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        na.value = "grey70", name = "Immuno-phenotype") +
    labs(title = "IHC marker+ fraction vs bulk-RNA expression",
         subtitle = "points coloured by the clinical immuno-phenotype; the fit is over all patients",
         x = "IHC fraction of cells positive", y = "log1p(bulk expression)")
}
```

Sensitivity: replace the hard-threshold positive fraction with the per-patient
mean staining intensity (mean z-score over non-NA cells):

```
<marker>_z(p) = (sum over p's cells with a non-NA <marker>_zscore of that zscore)
                / (cells of p with a non-NA <marker>_zscore)     (ihc_marker_fraction)
```

Then correlate `<marker>_z` against `rna_expr` per marker with the same statistic
(Spearman correlation across patients).

```{r rna-zscore-sensitivity}
ihc_z <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_z")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_z") |>
  mutate(marker = sub("_z$", "", marker))

paired_z <- marker_expr |>
  inner_join(ihc_z, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_z))

if (nrow(paired_z) > 0) {
  paired_z |>
    group_by(marker) |>
    group_modify(~ paired_spearman(.x$ihc_z, .x$rna_expr)) |>
    ungroup() |>
    arrange(desc(rho)) |>
    knitr::kable(digits = 3,
                 caption = "Spearman: IHC mean z-score vs bulk-RNA expression")
}
```

### Notes

- Both IHC reductions use `na.rm = TRUE`: `<marker>_posfrac` and `<marker>_z`
  average over the non-NA sign / z-score cells of a patient, so the effective
  denominator is that count, not necessarily all of the patient's cells.
- `rna_expr` averages only genes actually present in `counts_data` (a marker's
  available genes); genes in a marker's list that are absent from the matrix are
  reported by the `missing_genes` message and excluded from the mean. A marker with no
  available genes contributes no RNA row.
- The gene join is `relationship = "many-to-many"`, but each (sample, gene) has a
  single `expr`, so the per-(marker, sample) mean is over distinct genes.
- `PANCK` averages `EPCAM` + keratins; `SMA` -> `ACTA2`, `CD56` -> `NCAM1`,
  `PDL1` -> `CD274`, `PD1` -> `PDCD1`, `FSP1` -> `S100A4`. `wrongL1CAM` is
  excluded by construction (not in `marker_gene_map`).
- Correlations are cross-patient and small-n; every row carries its `n`.


## Part 2 — Deconvolution scores vs IHC lineage fractions

The same question one level up: instead of a marker, a whole cell type. `immunedeconv`
estimates cell-type abundance from the same bulk matrix; the IHC panel counts those
cells directly. Restricted to the four lineages both sides can name (CD8T, CD4T, Treg,
NK), and coloured by the same hot/cold call as part 1.


### Starting data and notation

Every value below is derived from two inputs, bridged by the clinical table:

- `ihc_data` — FlowPath single cells deduplicated to **one copy per patient**
  (whole slide, no annotation boundary). Each cell carries a `lineage` =
  `phenotype_clean` collapsed via the `phenotype_lineage` table.
- `counts_data` — DESeq2-normalized bulk RNA, one row per sample, one column per
  gene, keyed by `Sample` (a CRF id).
- `crf2slide` and `clin` — the id bridge and the hot/cold call, both built once in
  **Shared keys** above and only read here.

Comparison is anchored to the four comparable lineages CD8T, CD4T, Treg, NK
(`comparable_lineages`) — the only IHC phenotype lineages with a clean
deconvolution counterpart. `deconv_to_lineage()` maps each method's cell-type
string to one of these (case-insensitive pattern match: `regulatory|treg` ->
Treg, `cd8` -> CD8T, `cd4` -> CD4T, `^nk|nk cell|natural killer` -> NK),
everything else to `NA`. All correlations are Spearman correlation via
`paired_spearman` (rank-based; `NA` when fewer than 3 shared patients or either
vector is constant).

### Bulk-RNA expression matrix

`expr_mat` is `counts_data` (sample by gene) transposed to gene by sample, cast
double, with gene symbols as row names:

```
expr_mat[gene, sample] = counts_data[sample, gene]   (transpose)
```

```{r deconv-expr-matrix}
expr_mat <- counts_data |>
  column_to_rownames("Sample") |>
  as.matrix() |>
  t()

storage.mode(expr_mat) <- "double"
cat("expression matrix:", nrow(expr_mat), "genes x", ncol(expr_mat), "samples\n")
```

> Caveat: quanTIseq/EPIC are designed for TPM. DESeq2-normalized counts are used
> here as a pragmatic proxy; absolute fractions are therefore approximate, but the
> *cross-patient ranking* (what the Spearman validation uses) is robust to this.

### Method configuration

Each method is gated by the inputs it needs, not preference:

- `methods_free` (quanTIseq, EPIC, MCPcounter, xCell, ABIS) — run on `expr_mat`
  alone. Always enabled.
- `methods_indication` (TIMER, ConsensusTME) — need one cancer-type code per
  sample (`INDICATIONS`, length = `ncol(expr_mat)`). Enabled iff `INDICATIONS`
  is non-`NULL`; here it is `NULL`, so both are skipped.
- CIBERSORT / CIBERSORT (abs.) — license-gated `CIBERSORT.R` + `LM22.txt`, paths
  from env vars `CIBERSORT_BINARY` / `CIBERSORT_MAT`. Run iff both are set and the
  files exist.
- BayesPrism — needs an scRNA-seq reference at `data/sc_reference.rds` via
  `omnideconv`. Run iff present.

```{r deconv-method-config}
# First-generation methods that run on the bulk matrix with no extra artifacts.
methods_free <- c("quantiseq", "epic", "mcp_counter", "xcell", "abis")

# TIMER / ConsensusTME need one cancer-type code per SAMPLE (length == ncol(expr_mat),
# same order). Set INDICATIONS to enable them, e.g. rep("READ", ncol(expr_mat)) for a
# single-indication cohort, or a per-sample vector. NULL -> both skipped.
INDICATIONS        <- NULL
methods_indication <- if (!is.null(INDICATIONS)) c("timer", "consensus_tme") else character(0)

# CIBERSORT is license-gated: register at cibersort.stanford.edu, download CIBERSORT.R
# and LM22.txt, and point these at them (env vars keep the paths out of the repo).
# Absent -> CIBERSORT and CIBERSORT (abs.) are skipped.
CIBERSORT_BINARY <- Sys.getenv("CIBERSORT_BINARY", unset = NA)  # path to CIBERSORT.R
CIBERSORT_MAT    <- Sys.getenv("CIBERSORT_MAT",    unset = NA)  # path to LM22.txt

# BayesPrism (second-generation) needs a cell-type-annotated scRNA-seq reference,
# read from here() below. Absent -> BayesPrism is skipped.
SC_REFERENCE <- here("data", "sc_reference.rds")
```

### Run deconvolution methods

Each enabled method produces a cell-type by sample score table, pivoted long to
rows of (cell_type, sample, score, method). Then each row is keyed into slide
space:

```
crf_id     = norm_id(sample)
patient_id = crf2slide[crf_id]              (slide space)
lineage    = deconv_to_lineage(cell_type)
```

Rows with no slide crosswalk (`patient_id` is `NA`) are dropped.

```{r deconv-run}
deconv_long <- NULL
deconv_status <- NULL
if (requireNamespace("immunedeconv", quietly = TRUE)) {
  # Run one method, CAPTURING success/failure + a note (why it failed, or how many
  # cell types it returned). possibly()'s messages are hidden by message=FALSE, so
  # failures were previously invisible — this makes "only 3 methods ran" diagnosable.
  run_method <- function(m) {
    tryCatch({
      args <- list(gene_expression = expr_mat, method = m)
      if (m %in% c("timer", "consensus_tme")) args$indications <- INDICATIONS
      res <- do.call(immunedeconv::deconvolute, args) |>
        pivot_longer(-cell_type, names_to = "sample", values_to = "score") |>
        mutate(method = m)
      list(method = m, ok = TRUE, data = res,
           note = sprintf("%d cell types", dplyr::n_distinct(res$cell_type)))
    }, error = function(e) list(method = m, ok = FALSE, data = NULL,
                                note = conditionMessage(e)))
  }

  methods_run <- c(methods_free, methods_indication)
  # CIBERSORT (+ abs): only when the license-gated files are configured & present.
  if (!is.na(CIBERSORT_BINARY) && !is.na(CIBERSORT_MAT) &&
      file.exists(CIBERSORT_BINARY) && file.exists(CIBERSORT_MAT)) {
    immunedeconv::set_cibersort_binary(CIBERSORT_BINARY)
    immunedeconv::set_cibersort_mat(CIBERSORT_MAT)
    methods_run <- c(methods_run, "cibersort", "cibersort_abs")
  }

  runs <- purrr::map(methods_run, run_method)
  deconv_status <- purrr::map_dfr(runs, function(r)
    tibble::tibble(method = r$method,
                   status = if (r$ok) "ok" else "FAILED",
                   detail = r$note))

  ok_data <- runs |> purrr::keep(~ .x$ok) |> purrr::map("data")
  if (length(ok_data)) {
    deconv_long <- bind_rows(ok_data) |>
      mutate(crf_id     = norm_id(sample),
             patient_id = unname(crf2slide[crf_id]),   # crf id -> slide id
             lineage    = deconv_to_lineage(cell_type))
    n_unmapped  <- dplyr::n_distinct(deconv_long$crf_id[is.na(deconv_long$patient_id)])
    deconv_long <- deconv_long |> filter(!is.na(patient_id))
  } else {
    n_unmapped <- 0L
  }

  cat("methods:", sum(deconv_status$status == "ok"), "ok of",
      nrow(deconv_status), "attempted [",
      paste(sort(unique(deconv_long$method)), collapse = ", "), "]",
      "| deconvolution rows:", if (is.null(deconv_long)) 0L else nrow(deconv_long),
      "|", n_unmapped, "RNA sample(s) had no slide crosswalk\n")
} else {
  message("immunedeconv not installed — install it to run this report.")
}
```

```{r deconv-status, echo=FALSE}
if (!is.null(deconv_status))
  knitr::kable(deconv_status,
    caption = paste("Deconvolution method run status. A FAILED row is why that",
                    "method is absent downstream; the `detail` is its error message."))
```

### BayesPrism (single-cell reference; scaffolded)

`immunedeconv::deconvolute()` covers only bulk-signature methods. BayesPrism is a
second-generation method (fractions from a cell-type-annotated scRNA-seq
reference, via `omnideconv`); this project ships no reference, so the block is a
ready-to-run scaffold that activates only when `data/sc_reference.rds` is present
(an `.rds` list: `counts` = gene by cell, `labels` = per-cell type, `batch` =
per-cell sample id) and skips cleanly otherwise. Its output is mapped to
(patient_id, lineage) by the same `norm_id` -> `crf2slide` / `deconv_to_lineage`
pipeline as the other methods, then row-bound into `deconv_long`. Verify the
`omnideconv` argument names against your installed version before trusting the
numbers.

```{r deconv-bayesprism}
bayes_long <- NULL
if (file.exists(SC_REFERENCE) && requireNamespace("omnideconv", quietly = TRUE)) {
  ref <- readRDS(SC_REFERENCE)
  bayes_long <- tryCatch({
    # build_model(): learn the reference; deconvolute(): apply it to the bulk matrix.
    model <- omnideconv::build_model(
      single_cell_object    = ref$counts,
      cell_type_annotations = ref$labels,
      batch_ids             = ref$batch,
      method                = "bayesprism",
      bulk_gene_expression  = expr_mat)
    frac <- omnideconv::deconvolute(
      bulk_gene_expression  = expr_mat,
      model                 = model,
      single_cell_object    = ref$counts,
      cell_type_annotations = ref$labels,
      batch_ids             = ref$batch,
      method                = "bayesprism")
    as.data.frame(frac) |>
      tibble::rownames_to_column("sample") |>            # omnideconv returns sample x cell_type
      pivot_longer(-sample, names_to = "cell_type", values_to = "score") |>
      mutate(method = "bayesprism")
  }, error = function(e) { message("BayesPrism failed: ", conditionMessage(e)); NULL })
} else {
  message("BayesPrism skipped — needs data/sc_reference.rds + the omnideconv package.")
}

if (!is.null(bayes_long) && !is.null(deconv_long)) {
  bayes_long <- bayes_long |>
    mutate(crf_id     = norm_id(sample),
           patient_id = unname(crf2slide[crf_id]),
           lineage    = deconv_to_lineage(cell_type)) |>
    filter(!is.na(patient_id))
  deconv_long <- bind_rows(deconv_long, bayes_long)
  cat("BayesPrism added:", nrow(bayes_long), "rows\n")
}
```

### IHC lineage fractions (reference)

`ihc_lineage_fraction(ihc_data)` counts cells per patient and lineage and
normalises by the patient's total cell count (over **all** lineages, `NA`
included), giving the whole-slide fraction:

```
ihc_frac(patient, lineage) = (cells of patient with that lineage) / (all cells of patient)
where: the denominator is ALL cells of the patient (every lineage, NA included),
       before restricting to the four comparable lineages.
```

Restricted to the four comparable lineages, then **completed to 0** across the
full patient-by-lineage grid: a lineage a patient simply lacks reads as 0
(present), not a missing row — otherwise the `inner_join` below would silently
drop that patient from that lineage's panel.

```{r deconv-ihc-lineages}
lin_raw <- ihc_lineage_fraction(ihc_data) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, ihc_frac = frac)

# Complete the patient x lineage grid: a lineage a patient simply lacks must read
# as 0 (present), not a missing row. Otherwise the inner_join below silently drops
# that patient from that lineage's panel — the "plot never shows all patients" bug.
ihc_lin <- expand.grid(patient_id = unique(slide_key(ihc_data$patient_id)),
                       lineage    = comparable_lineages,
                       stringsAsFactors = FALSE) |>
  left_join(lin_raw, by = c("patient_id", "lineage")) |>
  mutate(ihc_frac = ifelse(is.na(ihc_frac), 0, ihc_frac))

cat("IHC patients with lineage fractions:", dplyr::n_distinct(ihc_lin$patient_id), "\n")
```

### Method vs IHC — per-lineage concordance

Per method, patient and lineage, the score sums a method's sub-types mapping to
the same lineage (e.g. two NK sub-types):

```
score(method, patient, lineage) = sum of that method's sub-type scores mapping to the lineage
where: the sub-types are the method's rows for that patient whose
       deconv_to_lineage(cell_type) equals the lineage.
```

This is inner-joined to `ihc_frac(patient, lineage)` on patient and lineage,
keeping finite pairs.

```{r deconv-concordance}
deconv_paired <- NULL
if (!is.null(deconv_long) && nrow(deconv_long) > 0) {
  deconv_lin <- deconv_long |>
    filter(!is.na(lineage)) |>
    group_by(method, patient_id, lineage) |>
    summarise(score = sum(score), .groups = "drop")

  deconv_paired <- deconv_lin |>
    inner_join(ihc_lin, by = c("patient_id", "lineage")) |>
    filter(is.finite(score), is.finite(ihc_frac))
  cat("deconv_paired (method x lineage x patient) rows:", nrow(deconv_paired), "\n")
}
```

Per method and lineage, concordance is the cross-patient Spearman correlation
between the method's score and the IHC fraction; method ranking averages those
across lineages:

```
concordance(method, lineage) = Spearman correlation between score and ihc_frac across patients
method ranking(method)       = average of those Spearman values across lineages (na.rm)
where: n_lineages = the number of lineages with a finite Spearman value.
```

```{r deconv-concordance-table}
if (!is.null(deconv_paired) && nrow(deconv_paired) > 0) {
  deconv_summary <- deconv_paired |>
    group_by(method, lineage) |>
    group_modify(~ paired_spearman(.x$score, .x$ihc_frac)) |>
    ungroup() |>
    arrange(lineage, desc(rho))
  knitr::kable(deconv_summary, digits = 3,
               caption = "Spearman IHC vs deconvolution, per method and lineage")

  # Which method agrees best overall (mean rho across lineages).
  method_rank <- deconv_summary |>
    group_by(method) |>
    summarise(mean_rho = mean(rho, na.rm = TRUE),
              n_lineages = sum(is.finite(rho)), .groups = "drop") |>
    arrange(desc(mean_rho))
  knitr::kable(method_rank, digits = 3,
               caption = "Method ranking by mean Spearman across lineages")
}
```

Scatter of `ihc_frac` (x) vs `score` (y), faceted method by lineage.

```{r deconv-concordance-plot}
if (!is.null(deconv_paired) && nrow(deconv_paired) > 0) {
  # Same hot/cold colouring as the marker panel above, on the same `clin` lookup:
  # a method that tracks IHC only in hot tumours shows up here as a colour-split
  # cloud rather than as a middling overall rho.
  df <- deconv_paired |>
    left_join(clin, by = "patient_id") |>
    mutate(immuno_phe = hotcold_order(immuno_phe))
  ggplot(df, aes(ihc_frac, score)) +
    geom_smooth(method = "lm", se = FALSE, colour = FIT_LINE, formula = y ~ x) +
    geom_point(aes(colour = immuno_phe), alpha = 0.75, size = 1.8) +
    facet_grid(method ~ lineage, scales = "free") +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        na.value = "grey70", name = "Immuno-phenotype") +
    # Every panel here has its OWN x and y range (scales = "free"), because the
    # methods' scores are on incomparable scales. theme_paper_panels() draws the
    # border and the wider gutter that say so — without them twenty panels sit edge
    # to edge and read as one continuous field on a shared axis, which is the single
    # wrong conclusion this figure can produce.
    theme_paper_panels() +
    labs(title = "IHC lineage fraction vs bulk-RNA deconvolution score",
         subtitle = "points coloured by the clinical immuno-phenotype; the fit is over all patients; every panel has its own scales",
         x = "IHC fraction of all cells (unitless, 0-1)",
         y = "Deconvolution score (unitless)")
}
```

### Cross-method agreement (methods vs each other and vs IHC)

Treat IHC as one more "source" alongside each method, giving one value per source,
patient and lineage: the per-lineage deconvolution score for a method, or
`ihc_frac` for source `IHC`.

```{r deconv-agreement}
agr <- NULL
if (!is.null(deconv_long) && nrow(deconv_long) > 0) {
  # per (method, patient, lineage) score, summing a method's split sub-types.
  deconv_lin <- deconv_long |>
    filter(!is.na(lineage)) |>
    group_by(method, patient_id, lineage) |>
    summarise(score = sum(score), .groups = "drop")

  agree_src <- bind_rows(
    deconv_lin |> transmute(source = method, patient_id, lineage, value = score),
    ihc_lin    |> transmute(source = "IHC",  patient_id, lineage, value = ihc_frac)
  )
  agr <- pairwise_agreement(agree_src)
}
```

`pairwise_agreement` returns a symmetric long table over every ordered source
pair (a, b):

```
pairwise_agreement(a, b):
  if a == b:             mean_rho = 1,   n_lineages = NA
  if no shared patients: mean_rho = NA,  n_lineages = 0
  otherwise:             mean_rho   = average across lineages of the per-lineage
                                      Spearman between a and b over their shared patients
                         n_lineages = the number of lineages with a finite Spearman value
where: for each lineage, the per-lineage Spearman correlates a's value against
       b's value across the patients the two sources share.
```

Rank-based throughout, so the methods' differing score scales do not matter.

```{r deconv-agreement-heatmap, fig.height=6}
if (!is.null(agr) && nrow(agr) > 0) {
  # order sources by mean off-diagonal agreement (most-consensual first); keep IHC visible.
  ord <- agr |>
    filter(a != b) |>
    group_by(a) |>
    summarise(m = mean(mean_rho, na.rm = TRUE), .groups = "drop") |>
    arrange(m) |>
    pull(a)
  agr <- agr |> mutate(a = factor(a, levels = ord), b = factor(b, levels = ord))

  ggplot(agr, aes(a, b, fill = mean_rho)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_text(aes(label = ifelse(is.na(mean_rho), "·", sprintf("%.2f", mean_rho))),
              size = pt_text(6.5)) +
    scale_fill_div(midpoint = 0, limits = c(-1, 1), name = "mean rho", guide = guide_cbar()) +
    coord_equal() +
    labs(title = "Cross-source agreement (mean Spearman across lineages)",
         subtitle = "Each deconvolution method vs every other and vs IHC; averaged over CD8T/CD4T/Treg/NK",
         x = NULL, y = NULL) +
    theme_paper_tile() +
    theme(axis.text.x = element_text(angle = 40, hjust = 1))
}
```

Per source:

```
mean_rho_vs_others(a) = average of mean_rho(a, b) over all other sources b
rho_vs_IHC(a)         = mean_rho(a, IHC)
```

`mean_rho_vs_others` is agreement with the consensus of the others; `rho_vs_IHC`
is agreement with IHC alone. A high vs-others but low vs-IHC source (or the
reverse) localises where methods and IHC diverge.

```{r deconv-agreement-ranks}
if (!is.null(agr) && nrow(agr) > 0) {
  # How much each source agrees with the consensus of the OTHERS (mean off-diagonal),
  # and separately with IHC alone.
  vs_ihc <- agr |> filter(b == "IHC", a != "IHC") |> select(source = a, rho_vs_IHC = mean_rho)
  agr |>
    filter(a != b) |>
    group_by(source = a) |>
    summarise(mean_rho_vs_others = mean(mean_rho, na.rm = TRUE), .groups = "drop") |>
    left_join(vs_ihc, by = "source") |>
    arrange(desc(mean_rho_vs_others)) |>
    knitr::kable(digits = 3,
      caption = "Per-source agreement with the other sources vs with IHC alone. A high vs-others but low vs-IHC source (or the reverse) flags where methods and IHC diverge.")
}
```

### Notes

- Relative-score methods (MCPcounter, xCell, CIBERSORT rel.) are rank-only; their
  slope against IHC is not meaningful, so every statistic here is Spearman.
  Absolute-fraction methods (quanTIseq, EPIC, CIBERSORT abs., ABIS) are the fair
  apples-to-apples comparison but are still compared by rank for uniformity.
- Only the four comparable lineages CD8T, CD4T, Treg, NK are compared: the IHC
  panel has no macrophage or B-cell phenotype, so `deconv_to_lineage` returns `NA`
  for those and they are dropped (part 1 above does a marker-level CD163/CD14
  check instead).
- Cross-patient Spearman needs at least 3 shared patients (and non-constant
  vectors); sparser method-lineage or source pairs return `NA` rather than a
  spurious correlation, and mean_rho averages over the finite lineages only.
  With this cohort's size the agreement matrix is descriptive — read it for
  *pattern*, not significance.
- The CIBERSORT and BayesPrism blocks are scaffolds — verify `set_cibersort_*()`
  and `omnideconv::build_model()/deconvolute()` argument names against your
  installed versions before trusting their output.

```{r deconv-cache-paired, include=FALSE}
# Hand `deconv_paired` to the figure-assembly page rather than making it recompute.
# Deconvolution needs immunedeconv, counts.RData and several minutes; the paper
# panel is a re-cut of THIS frame, so re-deriving it there would be both slow and
# a second place for the mapping to drift. Written only when there is something to
# write, so a skipped deconvolution run leaves the previous cache alone instead of
# replacing it with an empty one.
if (!is.null(deconv_paired) && nrow(deconv_paired) > 0) {
  dir.create(here("output"), showWarnings = FALSE, recursive = TRUE)
  # Cached WITH the hot/cold call joined on. Figure 5(c) colours by it, and deriving
  # it there would mean the paper page reopening the clinical table and re-deriving
  # slide keys — a second place for that mapping to drift, which is exactly what
  # this page exists to prevent.
  saveRDS(left_join(deconv_paired, clin, by = "patient_id"),
          here("output", "paired_deconv.rds"))
  cat("cached", nrow(deconv_paired), "deconv_paired rows for paper_figures.Rmd\n")
}
```


## Part 3 — Invasive-margin immunity vs the hot/cold call

Neither bulk readout above is spatial, and "hot" is a spatial claim: the clinical
immuno-phenotype is a statement about where the immune cells are, not just how many
there are. This part tests it directly — a physical band around the pathologist
polygon, the CD45+ fraction and density inside it, and those against `HOT score` and
`Immuno-phenotype`.

This is the only part that needs real geometry (a distance band cannot come from a
binary in-annotation flag), so it keeps `sf` and is gated on the geojson tree being
present; parts 1 and 2 run without it.


### Starting data and notation

Inputs: `ihc_data` (whole-slide FlowPath cells, one copy per patient, with
`centroid_x`/`centroid_y` in **microns**), the pathologist annotation **geojson**
polygons (in **pixels**), and the clinical `HOT score` / `Immuno-phenotype`.

Unlike the inside-the-annotation checks, membership here needs real geometry (a
distance band cannot come from the binary `Out_of_annotation` flag), so this report
keeps `sf`. Cells are placed in the polygon pixel frame by a **fixed** conversion:

```
x_px = centroid_x / 0.325
y_px = centroid_y / 0.325           (0.325 microns per pixel)
```

So 1 pixel = 0.325 microns, which gives the two conversions used below:

```
threshold d microns  ->  d / 0.325 pixels
pixel area A         ->  A * 0.325^2 / 1e6  square millimetres
```

### Margin and core regions

For a core polygon P (the union of a patient's annotations, or each single
annotation) and a half-width `d` microns, with `d_px = d / 0.325`:

```
margin band = (P grown by d_px)  minus  (P shrunk by d_px)
deep core   =  P shrunk by d_px
```

(grow/shrink are `st_buffer` by +/- `d_px`; `margin_regions`.) If `d_px` exceeds
the core's inradius the shrunk core is empty, and the margin band falls back to the
full grown buffer. Thresholds analysed: `d` = 50, 100, 200 microns. These are
narrower than the 500 µm consensus-Immunoscore half-width on purpose: at 500 µm the
grown buffer swallows most of a small annotation and the "margin" stops being a
margin, so the band is read at widths where it still isolates the boundary. The
widest, `D_MAIN`, is the one every single-threshold panel below uses.

```{r margin-metrics}
THRESHOLDS_UM <- c(50, 100, 200)
UM_PER_PX     <- 0.325

# The headline width, derived rather than repeated. Every panel that shows one
# threshold reads D_MAIN, so changing THRESHOLDS_UM moves them all together — the
# previous version spelled `500` into eight filters and four captions, which is four
# captions that can disagree with the data they label.
D_MAIN        <- max(THRESHOLDS_UM)

# load_annotations() returns NULL when there is no geojson tree at all. This part is
# the only one of the three that needs real polygon geometry, so the absence is
# recorded as a flag and every chunk below carries `eval = have_margin`. Letting the
# first filter() hit a NULL frame instead would take parts 1 and 2 down with it, and
# those read bulk RNA and whole-slide cells — nothing to do with the annotations.
annots      <- load_annotations(patient_ids = neoplastic_data$SAMPLE)
have_margin <- !is.null(annots) && nrow(annots) > 0

if (!have_margin)
  cat("No pathologist geojson under", ANNOTATION_DIR(),
      "- the invasive-margin section is skipped.\n")

if (have_margin) {
  peri_union   <- ihc_periphery_metrics(ihc_data, annots, scope = "union",
                                        thresholds_um = THRESHOLDS_UM, um_per_px = UM_PER_PX)
  peri_per_ann <- ihc_periphery_metrics(ihc_data, annots, scope = "per_annotation",
                                        thresholds_um = THRESHOLDS_UM, um_per_px = UM_PER_PX)

  cat("union rows:", nrow(peri_union),
      "| patients:", dplyr::n_distinct(peri_union$patient_id),
      "| per-annotation rows:", nrow(peri_per_ann), "\n")
}
```

### Per-region quantities

For the cells inside a region (`st_within`), `region_ratios_area()` returns the
membership counts (as in `clinical_flowpath.Rmd`) plus:

```
area_mm2            = region pixel area * 0.325^2 / 1e6
dens_all_per_mm2    = n_inside      / area_mm2
dens_cd45_per_mm2   = n_cd45_inside / area_mm2
dens_<lineage>_per_mm2 = n_<lineage> / area_mm2    (CD8T, CD4T, Treg, NK)

cd45_over_inside    = n_cd45_inside / n_inside
```

Densities are `NA` when the region is empty.

```{r margin-coverage, eval = have_margin, echo=FALSE}
peri_union |>
  filter(threshold_um == D_MAIN) |>
  select(patient_id, region, source, um_per_unit, area_mm2,
         n_inside, n_cd45_inside, cd45_over_inside) |>
  arrange(patient_id, region) |>
  knitr::kable(digits = 3,
    caption = paste0("Cells and area per region at d = ", D_MAIN,
                     " µm (union). n_inside = 0 ⇒ empty region."))
```

### Margin immunity vs clinical HOT score

Statistic: Spearman correlation (per threshold) between the margin-band value and
the numeric `HOT score`, for both:

```
fraction = cd45_over_inside   = n_cd45_inside / n_inside
density  = dens_cd45_per_mm2  = n_cd45_inside / area_mm2
```

```{r margin-hot-join, eval = have_margin}
margin_union <- peri_union |>
  filter(region == "margin") |>
  inner_join(clin, by = "patient_id")

cat("patients with margin metrics + clinical immune phenotype:",
    dplyr::n_distinct(margin_union$patient_id), "\n")
```

```{r margin-hot-score-fraction, eval = have_margin, fig.height=4}
df <- margin_union |>
  filter(is.finite(cd45_over_inside), is.finite(hot_score))
if (nrow(df) > 2) {
  labs_tbl <- df |>
    group_by(threshold_um) |>
    group_modify(~ paired_spearman(.x$cd45_over_inside, .x$hot_score)) |>
    mutate(lab = sprintf("rho = %.2f\np = %.3g (n = %d)", rho, p, n))
  # Coloured by the categorical immuno-phenotype while the axis is the continuous
  # HOT score: the two are separate clinical columns, and seeing them together is
  # the point — a cold-labelled case sitting high on the score axis is a
  # disagreement inside the clinical record, not in the imaging.
  df <- df |> mutate(immuno_phe = hotcold_order(immuno_phe))
  ggplot(df, aes(cd45_over_inside, hot_score)) +
    geom_smooth(method = "lm", se = FALSE, colour = FIT_LINE, formula = y ~ x) +
    geom_point(aes(colour = immuno_phe), alpha = 0.85, size = 2) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.2, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ threshold_um, labeller = labeller(threshold_um = ~ paste0("d = ", .x, " µm"))) +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        na.value = "grey70", name = "Immuno-phenotype") +
    labs(title = "Invasive-margin immune fraction vs clinical HOT score",
         x = "CD45+ fraction in margin band (unitless, 0-1)",
         y = "Clinical HOT score (unitless)")
}
```

```{r margin-hot-score-density, eval = have_margin, fig.height=4}
df <- margin_union |>
  filter(is.finite(dens_cd45_per_mm2), is.finite(hot_score))
if (nrow(df) > 2) {
  labs_tbl <- df |>
    group_by(threshold_um) |>
    group_modify(~ paired_spearman(.x$dens_cd45_per_mm2, .x$hot_score)) |>
    mutate(lab = sprintf("rho = %.2f\np = %.3g (n = %d)", rho, p, n))
  # Coloured by the categorical immuno-phenotype while the axis is the continuous
  # HOT score: the two are separate clinical columns, and seeing them together is
  # the point — a cold-labelled case sitting high on the score axis is a
  # disagreement inside the clinical record, not in the imaging.
  df <- df |> mutate(immuno_phe = hotcold_order(immuno_phe))
  ggplot(df, aes(dens_cd45_per_mm2, hot_score)) +
    geom_smooth(method = "lm", se = FALSE, colour = FIT_LINE, formula = y ~ x) +
    geom_point(aes(colour = immuno_phe), alpha = 0.85, size = 2) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.2, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ threshold_um, labeller = labeller(threshold_um = ~ paste0("d = ", .x, " µm"))) +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        na.value = "grey70", name = "Immuno-phenotype") +
    labs(title = "Invasive-margin immune density vs clinical HOT score",
         x = "CD45+ density in margin band (cells/mm²)",
         y = "Clinical HOT score (unitless)")
}
```

### Margin immunity by clinical immuno-phenotype

Distribution of the margin-band `cd45_over_inside` within each `Immuno-phenotype`
group (ordered cold -> intermediate -> hot), faceted by threshold.

```{r margin-phenotype-boxplot, eval = have_margin, fig.height=4}
df <- margin_union |>
  filter(!is.na(immuno_phe), is.finite(cd45_over_inside)) |>
  mutate(immuno_phe = hotcold_order(immuno_phe))
if (dplyr::n_distinct(na.omit(df$immuno_phe)) > 1 && nrow(df) > 3) {
  ggplot(df, aes(immuno_phe, cd45_over_inside, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    facet_wrap(~ threshold_um, labeller = labeller(threshold_um = ~ paste0("d = ", .x, " µm"))) +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        name = "Immuno-phenotype") +
    labs(title = "Invasive-margin immune fraction by clinical immuno-phenotype",
         x = NULL, y = "CD45+ fraction in margin band") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

### Margin-vs-core exclusion score

Per patient at the widest band, `d = D_MAIN` (200) microns:

```
exclusion_frac = cd45_over_inside (margin)  -  cd45_over_inside (core)
exclusion_dens = dens_cd45_per_mm2 (margin) -  dens_cd45_per_mm2 (core)
```

Positive means immunity is concentrated at the border.

```{r margin-core, eval = have_margin}
mc <- peri_union |>
  filter(threshold_um == D_MAIN) |>
  select(patient_id, region, cd45_over_inside,
         dens_cd45_per_mm2, dens_CD8T_per_mm2, n_inside) |>
  pivot_wider(names_from = region,
              values_from = c(cd45_over_inside, dens_cd45_per_mm2, dens_CD8T_per_mm2, n_inside)) |>
  mutate(
    exclusion_frac = cd45_over_inside_margin - cd45_over_inside_core,
    exclusion_dens = dens_cd45_per_mm2_margin - dens_cd45_per_mm2_core
  ) |>
  inner_join(clin, by = "patient_id")

mc |>
  select(patient_id, immuno_phe,
         margin_cd45 = cd45_over_inside_margin, core_cd45 = cd45_over_inside_core,
         exclusion_frac, hot_score) |>
  arrange(desc(exclusion_frac)) |>
  knitr::kable(digits = 3,
    caption = paste0("Margin-vs-core CD45+ fraction at d = ", D_MAIN,
                     " µm. Positive exclusion_frac ⇒ border-concentrated immunity."))
```

```{r margin-core-scatter, eval = have_margin, fig.height=4.5}
sc <- mc |> filter(is.finite(cd45_over_inside_margin), is.finite(cd45_over_inside_core))
if (nrow(sc) > 1) {
  aes_col <- if ("immuno_phe" %in% names(sc) && any(!is.na(sc$immuno_phe)))
    aes(cd45_over_inside_core, cd45_over_inside_margin, colour = hotcold_order(immuno_phe))
  else aes(cd45_over_inside_core, cd45_over_inside_margin)
  p <- ggplot(sc, aes_col) +
    geom_abline(slope = 1, linetype = "dashed", colour = "grey50") +
    geom_point(size = 3, alpha = 0.85) +
    coord_equal() +
    labs(title = paste0("Margin vs core CD45+ fraction (d = ", D_MAIN, " µm)"),
         x = "CD45+ fraction, deep core", y = "CD45+ fraction, invasive margin",
         colour = "Immuno-phenotype")
  if ("colour" %in% names(sc) || any(!is.na(sc$immuno_phe)))
    p <- p + scale_colour_manual(values = hotcold_cols(levels(hotcold_order(sc$immuno_phe))),
                                 name = "Immuno-phenotype", na.value = "grey70")
  print(p)
}
```

```{r margin-exclusion-by-phenotype, eval = have_margin, fig.height=4}
df <- mc |>
  filter(!is.na(immuno_phe), is.finite(exclusion_frac)) |>
  mutate(immuno_phe = hotcold_order(immuno_phe))
if (dplyr::n_distinct(na.omit(df$immuno_phe)) > 1 && nrow(df) > 3) {
  ggplot(df, aes(immuno_phe, exclusion_frac, colour = immuno_phe)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)), name = "Immuno-phenotype") +
    labs(title = paste0("Margin-minus-core exclusion score by clinical phenotype (d = ", D_MAIN, " µm)"),
         x = NULL, y = "Exclusion score: margin − core CD45+ fraction (unitless)") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

### Single annotations vs union

Per-annotation margin fraction (each single-annotation core) vs the patient's union
margin fraction, at `d = D_MAIN` (200) microns — a stability check across a patient's
separately drawn regions.

```{r margin-per-ann-vs-union, eval = have_margin, fig.height=4.5}
pa <- peri_per_ann |>
  filter(region == "margin", threshold_um == D_MAIN) |>
  select(patient_id, annotation, cd45_over_inside)
un <- peri_union |>
  filter(region == "margin", threshold_um == D_MAIN) |>
  select(patient_id, union_cd45 = cd45_over_inside)
cmp <- inner_join(pa, un, by = "patient_id") |>
  filter(is.finite(cd45_over_inside), is.finite(union_cd45))
if (nrow(cmp) > 1) {
  ggplot(cmp, aes(patient_id, cd45_over_inside)) +
    geom_line(aes(y = union_cd45, group = 1), colour = "grey60") +
    geom_point(aes(y = union_cd45), colour = "grey40", size = 3, shape = 18) +
    geom_jitter(aes(colour = annotation), width = 0.05, size = 2, alpha = 0.85) +
    labs(title = paste0("Margin immune fraction: single annotations vs union (d = ", D_MAIN, " µm)"),
         x = "Patient", y = "CD45+ fraction in margin band (unitless, 0-1)",
         colour = "Annotation") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

### Margin composition (three denominators)

The same three-denominator composition used inside the tumour, applied to the
margin band at `d = D_MAIN` (200) microns:

```
frac_inside = n_<lineage> / n_inside         (of all cells in band)
frac_tumor  = n_<lineage> / n_tumor_inside   (of tumour cells in band)
frac_cd45   = n_<lineage> / n_cd45_inside    (of CD45+ cells in band)
```

```{r margin-composition, eval = have_margin, fig.height=5}
comp <- peri_union |>
  filter(region == "margin", threshold_um == D_MAIN) |>
  region_composition() |>
  pivot_longer(c(frac_inside, frac_tumor, frac_cd45),
               names_to = "normalization", values_to = "frac") |>
  mutate(normalization = factor(
           c(frac_inside = "of all cells", frac_tumor = "of tumour cells",
             frac_cd45 = "of CD45+ cells")[normalization],
           levels = c("of all cells", "of tumour cells", "of CD45+ cells"))) |>
  filter(is.finite(frac))

if (nrow(comp) > 0) {
  ggplot(comp, aes(lineage, frac, fill = lineage)) +
    geom_boxplot(outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = .15, alpha = .5, size = 1) +
    facet_wrap(~ normalization, scales = "free_y") +
    # Named lineage palette, not positional `oi` — see the note on the same plot
    # in clinical_flowpath.Rmd. These are the same populations as that figure and
    # must therefore be the same colours.
    scale_fill_lineage(guide = "none") +
    labs(title = paste0("Immune composition in the invasive-margin band (d = ", D_MAIN, " µm)"),
         subtitle = with_n(NULL, comp$patient_id, "patients"),
         x = NULL, y = "Cell fraction (unitless, 0-1)") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

### Margin cell-type fractions: hot vs cold

Same margin-band composition (d = D_MAIN = 200 µm, union), now split by the clinical
`Immuno-phenotype`: for each lineage, the boxes contrast **hot** vs **cold** tumours.

```
group(patient) = hot / intermediate / cold   (from clinical Immuno-phenotype)
compared per lineage and per denominator (of all cells / tumour cells / CD45+ cells)
```

```{r margin-composition-hotcold, eval = have_margin, fig.height=5}
comp_hc <- peri_union |>
  filter(region == "margin", threshold_um == D_MAIN) |>
  region_composition() |>
  inner_join(clin, by = "patient_id") |>
  filter(!is.na(immuno_phe)) |>
  mutate(immuno_phe = hotcold_order(immuno_phe)) |>   # cold -> intermediate -> hot
  pivot_longer(c(frac_inside, frac_tumor, frac_cd45),
               names_to = "normalization", values_to = "frac") |>
  mutate(normalization = factor(
           c(frac_inside = "of all cells", frac_tumor = "of tumour cells",
             frac_cd45 = "of CD45+ cells")[normalization],
           levels = c("of all cells", "of tumour cells", "of CD45+ cells"))) |>
  filter(is.finite(frac))

if (nrow(comp_hc) > 0 && dplyr::n_distinct(na.omit(comp_hc$immuno_phe)) > 1) {
  ggplot(comp_hc, aes(lineage, frac, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA, position = position_dodge(width = .75)) +
    geom_point(position = position_jitterdodge(jitter.width = .12, dodge.width = .75),
               alpha = .6, size = 1) +
    facet_wrap(~ normalization, scales = "free_y") +
    scale_colour_manual(values = hotcold_cols(levels(comp_hc$immuno_phe)),
                        name = "Immuno-phenotype") +   # hot = red, cold = light blue
    labs(title = paste0("Invasive-margin cell-type fractions: hot vs cold (d = ", D_MAIN, " µm)"),
         subtitle = sprintf("%d patients with a margin band and an immuno-phenotype",
                            dplyr::n_distinct(comp_hc$patient_id)),
         x = NULL, y = "Cell fraction in margin band (unitless, 0-1)") +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

### Notes

- Band width is physical: each polygon is buffered by `d / 0.325` pixel units; the
  `um_per_unit` column echoes the 0.325 µm/px conversion used.
- Geojson-only: a margin band needs a real boundary, so only patients with a
  pathologist geojson are covered (no `Out_of_annotation` fallback).
- Fraction vs density: a thin band can read immune-rich by fraction yet sparse by
  density (per mm²) — the Immunoscore unit.
- Small cohort; Spearman with reported `n` throughout.


```{r export-pdfs, include=FALSE}
# Every plot above is also rendered to PDF by the dev = c("png", "pdf") chunk option;
# this collects them into output/figures/<SLUG>/. Runs last, so they all exist by now.
export_pdf_figures(SLUG)
```
