---
title: "Molecular — hot/cold vs bulk RNA — Massimo1 — FlowPath, selected regions"
author: "sceriff0"
date: "2026-08-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
# renv: tidyverse + here (in the lockfile), sf + jsonlite for the margin geometry in
# part 3, and immunedeconv (GitHub: omnideconv/immunedeconv, NOT Bioconductor) for
# part 2 — see the method-configuration chunk there for the install recipe. Each part
# degrades to a message if its optional input is absent; none of them is required.
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(forcats)
  library(here)
  library(sf)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the analysis script that used to pull plot_theme.R in never runs —
# which silently left those plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, the per-arm neoplastic_* / ihc_* objects, counts_data)
# helpers. load_data.R prints colnames() on load; silence it. validation_helpers.R pulls
# in cell_tables.R (export-schema adapter) and plot_theme.R (house figure style, which
# is where hotcold_cols()/hotcold_order() come from).
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

SLUG <- "molecular_massimo1"
# WHICH ARM THIS PAGE IS ABOUT. Three arms phenotype the same slides three ways
# (code/arms.R) and load_data.R deliberately defines NO bare `ihc_data`: a page that
# forgot to name its arm would otherwise produce a figure indistinguishable from a
# figure about a different arm. Change this one constant to re-point the page.
ARM      <- "massimo1"
ihc_data <- ihc_for(ARM)

# The bulk-RNA and deconvolution sections stand on their own; only the chunks that
# JOIN an IHC quantity need the arm's cells. Gating just those keeps the RNA half of
# the page rendering when the arm's tree is not on disk.
have_arm <- nrow(ihc_data) > 0
if (!have_arm)
  cat("!! ARM `", ARM, "` HAS NO CELLS at ", arm_spec(ARM)$root_path,
      "\n   The IHC-vs-RNA and IHC-vs-deconvolution comparisons are skipped;",
      "\n   the bulk-RNA sections still render.\n", sep = "")
neoplastic <- neoplastic_for(ARM)
```

> **Arm: `massimo1`.** FlowPath's `_selected` per-region export, with a real whole-slide `_all` tier behind the union scope. Pathologist scores are the ORIGINAL read (`thr_head&neck.xlsx`).
>
> This page is one of three, identical in construction and differing only in which
> arm it reads. The arms' regions were drawn in **separate annotation sessions**, so
> `ANNOTATION_2` here is not `ANNOTATION_2` on the other pages — compare them by
> PATIENT, never region by region. `code/arms.R` states the rule in full.

```{r body, results = "asis", echo = FALSE}
# THE BODY IS SPLICED WITH knit_child(text = ...), NOT THE `child =` CHUNK OPTION.
#
# That is not a style choice. workflowr's fig.path option hook does
#   options$fig.path <- create_figure_path(knitr::current_input())
# and inside a `child =` chunk knitr::current_input() returns the CHILD. So every
# arm's figures would land in figure/molecular_body/ — ONE directory shared by all
# three parents — and whichever page knitted last would silently overwrite the
# others. export_pdf_figures(SLUG) then copies those same overwritten files into
# all three output/figures/<slug>/ directories, so a panel labelled massimo1 would
# show massimo2's cells with nothing to indicate it.
#
# knit_child(text = readLines(...)) has no input file of its own, so
# current_input() stays the PARENT: figures go to figure/<parent>/ as they should,
# and workflowr's "custom fig.path was ignored" warning disappears too — that
# warning was the visible symptom of the collision, not the problem itself.
# tests/testthat/test-child-documents.R pins this idiom.
cat(knitr::knit_child(
      text  = readLines(here::here("analysis", "_children", "molecular_body.Rmd")),
      quiet = TRUE),
    sep = "\n")
```

```{r export-pdfs, include=FALSE}
# Every plot above is also rendered to PDF by the dev = c("png", "pdf") chunk option;
# this collects them into output/figures/<SLUG>/. Runs last, so they all exist by now.
export_pdf_figures(SLUG)
```
