---
title: "Molecular — quanTIseq vs IHC, whole slide and inside each annotation"
author: "sceriff0"
date: "2026-10-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r setup, include=FALSE}
# renv: tidyverse + here (in the lockfile), sf for the point-in-polygon behind the
# annotation scopes, and immunedeconv (GitHub: omnideconv/immunedeconv, NOT
# Bioconductor) for quanTIseq. The page degrades to a visible message if either the
# cells or immunedeconv are absent.
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(forcats)
  library(here)
  library(sf)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set: the
# figure defaults below call fig_width().
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/with_n()
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# load_data.R prints per-arm LOADED messages; silence them. validation_helpers.R
# holds the deconvolution <-> IHC mapping, deconv_scopes.R the cell scopes.
invisible(capture.output(suppressMessages(source(here("code", "load_data.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "deconv_scopes.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

SLUG <- "molecular_quantiseq"
# THIS PAGE IS CROSS-ARM, SO IT DECLARES NO SINGLE `ARM`. Every scope it draws names
# its own arm in DECONV_SCOPES (code/deconv_scopes.R), and each arm's cells are taken
# with ihc_for(<that arm>) — there is still no bare `ihc_data`. Panels from different
# arms are compared by PATIENT only: the arms' regions were drawn in separate sessions.
ARMS   <- unique(DECONV_SCOPES$arm)
METHOD <- "quantiseq"
cells_by_arm <- lapply(stats::setNames(ARMS, ARMS), ihc_for)

# The quanTIseq half stands on its own; every chunk that touches an IHC quantity is
# gated on at least one arm having cells on disk.
have_cells <- any(vapply(cells_by_arm, nrow, integer(1)) > 0)
```

> **Arms: `massimo1` and `massimo2`, one method: quanTIseq.** The per-arm molecular
> pages compare every deconvolution method against one arm's whole slide. This page
> turns that around: one method, and every cell set the IHC fraction can be taken
> over — each arm's whole slide, and the cells inside each arm's pathologist
> annotation. The two arms' annotations are different polygons over the same tissue,
> so panels are compared by PATIENT, never region by region.

One question, asked once per population quanTIseq can estimate: **does the
quanTIseq fraction from bulk RNA track the fraction of cells the IHC panel counts,
and does the answer depend on which cells are counted?**

## Shared keys: the id bridge and the hot/cold call

Bulk RNA is keyed by CRF id and IHC by slide id. `crf2slide` maps one to the other
(`crf_to_slide_map()` over `data/clinical_data.xlsx`: `ID CRF PRESERVE` to
`ID PATIENT`), and `clin` carries the clinical `Immuno-phenotype` used to colour
every point below. The line printed is the number of clinical rows and how many
carry each of the two immune columns.

```{r shared-keys}
crf2slide <- crf_to_slide_map(clinical_data)

clin <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`)) |>
  select(patient_id,
         hot_score  = any_of("HOT score"),
         immuno_phe = any_of("Immuno-phenotype")) |>
  mutate(hot_score = suppressWarnings(as.numeric(hot_score)))

cat("clinical rows:", nrow(clin),
    "| with a numeric HOT score:", sum(is.finite(clin$hot_score)),
    "| with an immuno-phenotype:", sum(!is.na(clin$immuno_phe)), "\n")
```

## quanTIseq on the bulk RNA

quanTIseq is run by `immunedeconv::deconvolute()` on the gene-by-sample matrix from
`data/counts.RData`, on the TPM scale when gene lengths were found and on the
DESeq2-normalised scale otherwise (`expr_scales`, built in `code/load_data.R`;
quanTIseq was fit on TPM, so the printed `scale` matters for absolute values and
much less for the cross-patient ranking used here). The result is one fraction per
cell type and sample, keyed into slide space with `crf2slide`; samples with no
slide are dropped and counted in the printed line.

```{r quantiseq-run}
EXPR_SCALE <- if (!is.null(expr_scales$tpm)) "tpm" else "normalized"
expr_mat   <- expr_scales[[EXPR_SCALE]]
storage.mode(expr_mat) <- "double"

qs_long <- NULL
if (requireNamespace("immunedeconv", quietly = TRUE)) {
  qs_long <- tryCatch(
    immunedeconv::deconvolute(expr_mat, METHOD) |>
      pivot_longer(-cell_type, names_to = "sample", values_to = "score") |>
      mutate(method     = METHOD,
             crf_id     = norm_id(sample),
             patient_id = unname(crf2slide[crf_id])),   # crf id -> slide id
    error = function(e) {
      cat("**quanTIseq FAILED:**", conditionMessage(e), "\n")
      NULL
    })
} else {
  # cat(), not message(): message = FALSE on this page would swallow it and leave a
  # page with no figures and no reason.
  cat("**immunedeconv not installed — quanTIseq skipped, so every comparison below",
      "is empty.** Install it with",
      "`remotes::install_github(\"omnideconv/immunedeconv\")` and re-knit.\n")
}

n_unmapped <- if (is.null(qs_long)) 0L else
  n_distinct(qs_long$crf_id[is.na(qs_long$patient_id)])
if (!is.null(qs_long)) qs_long <- filter(qs_long, !is.na(patient_id))

have_qs    <- !is.null(qs_long) && nrow(qs_long) > 0
have_pairs <- have_qs && have_cells
# The populations quanTIseq can estimate, in the gate tree's order (macro ->
# intermediate -> leaf -> cross-cutting). Every table and figure below keeps it.
# Bound here, ungated, because the IHC-only tables read it too.
qs_lineages <- if (have_qs)
  intersect(deconv_comparison_lineages$lineage, deconv_contributions(qs_long)$lineage) else
  character(0)
cat("expression matrix:", nrow(expr_mat), "genes x", ncol(expr_mat), "samples",
    "| scale:", EXPR_SCALE,
    "| quanTIseq cell types:", if (have_qs) n_distinct(qs_long$cell_type) else 0L,
    "| patients with a slide:", if (have_qs) n_distinct(qs_long$patient_id) else 0L,
    "|", n_unmapped, "RNA sample(s) had no slide crosswalk\n")
```

## Which quanTIseq cell type is compared with which IHC population

Each quanTIseq label is assigned to the IHC populations it can be compared with by
`deconv_contributions()` (`code/validation_helpers.R`), the same rules the per-arm
molecular pages print in full. A label either **is** a population (its direct
match) or is **summed into** a composite; quanTIseq's fractions are of all cells,
so its labels may be added. The B-cell and myeloid labels have no direct match
because the IHC panel has no B or myeloid marker — they are compared only as their
sum, `Immune_other`.

```{r quantiseq-map, eval = have_qs}
qs_contrib <- deconv_contributions(qs_long)
qs_scores  <- deconv_lineage_scores(qs_long)

dash <- function(x) ifelse(is.na(x) | x == "", "—", x)
qs_long |>
  distinct(cell_type) |>
  left_join(
    qs_contrib |>
      mutate(lineage = factor(lineage, levels = deconv_comparison_lineages$lineage)) |>
      arrange(lineage) |>
      group_by(cell_type) |>
      summarise(first  = lineage[1],
                is_pop = paste(lineage[route == "direct"], collapse = ", "),
                summed = paste(lineage[route == "sum"],    collapse = ", "),
                .groups = "drop"),
    by = "cell_type") |>
  arrange(first, cell_type) |>
  transmute(method = method_label(METHOD), cell_type,
            is_pop = dash(is_pop), summed = dash(summed)) |>
  knitr::kable(col.names = c("method", "cell type", "is the population", "summed into"),
               caption = "quanTIseq cell types and the IHC populations each feeds.")
```

The populations those labels add up to, each with its gate in the FlowPath
phenotyping tree and the tree leaves counted for it on the IHC side
(`deconv_comparison_lineages`). There is one scatter per row of this table in every
scope below.

```{r quantiseq-populations, eval = have_qs}
deconv_comparison_lineages |>
  filter(lineage %in% qs_lineages) |>
  select(lineage, level, ihc_gate, ihc_phenotypes) |>
  knitr::kable(col.names = c("population", "level", "IHC gate", "IHC leaves summed"),
               caption = "The populations compared on this page.")
```

## Which cells the IHC fraction is taken over

The IHC fraction is computed four times, once per **scope** in `DECONV_SCOPES`
(`code/deconv_scopes.R`). The table is that registry, with the number of cells each
scope's arm has on disk (`ihc_for(arm)`: one row per physical cell per patient —
`data/massimo1/FlowPath_csv_all/` for `massimo1`, the pooled and de-duplicated
`data/massimo2/csv/` for `massimo2`).

```{r scopes}
DECONV_SCOPES |>
  mutate(cells_loaded = vapply(arm, function(a) nrow(cells_by_arm[[a]]), integer(1)),
         polygons = c(union = "annotation_all/", region = "annotation/ (dissolved)")[tier] |>
           replace_na("none")) |>
  select(label, arm, kept, polygons, cells_loaded) |>
  knitr::kable(col.names = c("scope", "arm", "cells kept", "polygon set", "cells in the arm"),
               format.args = list(big.mark = ","),
               caption = "The four cell scopes. A scope whose arm has 0 cells on disk draws nothing below.")
```

Three things to hold on to when reading across scopes:

- **The two whole-slide scopes are different numbers.** The arms gate the same
  slides at different thresholds, so the same slide has a different CD8T fraction
  in each.
- **`massimo1` has one annotation scope.** Its `annotation_all` polygon is the exact
  union of that patient's `annotation_selected` regions, so a separate scope for the
  selected regions would redraw the same cells.
- **A slide with no polygon in a scope's set is cut by FlowPath's own inside/outside
  call** — the `Out_of_annotation` column of that slide's csv. This is the rule for
  this page only: the clinical pages count such a `massimo2` slide whole.

For each scope, every cell of the arm is marked inside or outside: by
point-in-polygon of its centroid against the scope's polygon set (`sf`), or by the
csv's `Out_of_annotation` flag where the patient has no polygon in that set
(`flag`); a whole-slide scope keeps every cell (`all_cells`). The table counts, per
scope and patient, the cells the arm holds and the cells the scope keeps, and names
the rule that decided.

```{r scope-ihc, eval = have_cells}
# One pass per scope: mark, count, and keep only the per-patient fractions — the
# per-cell frames are several hundred thousand rows each and are not held.
scope_ihc <- lapply(stats::setNames(DECONV_SCOPES$scope, DECONV_SCOPES$scope),
                    function(s) deconv_scope_ihc(s, cells_by_arm[[deconv_scope(s)$arm]]))
scope_frac <- bind_rows(map(scope_ihc, "fractions"))
scope_leaf <- bind_rows(map(scope_ihc, "leaves"))
scope_inv  <- bind_rows(map(scope_ihc, "inventory"))
scope_lab  <- stats::setNames(DECONV_SCOPES$label, DECONV_SCOPES$scope)

scope_inv |>
  mutate(scope = scope_lab[scope]) |>
  knitr::kable(col.names = c("scope", "patient", "decided by", "cells in the arm",
                             "cells kept", "% kept"),
               format.args = list(big.mark = ","),
               caption = "Cells behind each scope, per patient. `sf` = inside the scope's polygons, `flag` = the csv's own Out_of_annotation call (no polygon for that patient), `all_cells` = no restriction.")
```

Within each scope, `ihc_comparison_fraction()` counts the kept cells on each
population's tree leaves and divides by **all kept cells** of that patient — the
denominator quanTIseq's fractions share. The table is the median of that
per-patient fraction, per population and scope, for the populations quanTIseq can
estimate (every IHC population if quanTIseq did not run); it shows how far the
annotation moves the IHC side before any RNA is involved.

```{r scope-fractions, eval = have_cells}
scope_frac |>
  filter(!length(qs_lineages) | lineage %in% qs_lineages) |>
  mutate(lineage = factor(lineage, levels = deconv_comparison_lineages$lineage),
         scope   = factor(scope_lab[scope], levels = DECONV_SCOPES$label)) |>
  group_by(lineage, scope) |>
  summarise(median_frac = median(frac_all), .groups = "drop") |>
  pivot_wider(names_from = scope, values_from = median_frac, names_sort = TRUE) |>
  arrange(lineage) |>
  knitr::kable(digits = 4,
               caption = "IHC side: median per-patient fraction of all kept cells, per population and scope.")
```

## quanTIseq vs IHC, per population and scope

For each scope, the quanTIseq score of a population (the sum of the labels the
mapping table assigns to it) is inner-joined on patient and population to that
scope's IHC fraction (`deconv_pair_with_ihc()`). Agreement is the cross-patient
Spearman correlation (`paired_spearman()`: `NA` under 3 patients or when either
side is constant). Each cell of the table is rho with the number of patients it was
computed over.

```{r quantiseq-concordance, eval = have_pairs}
qs_paired <- scope_frac |>
  group_split(scope) |>
  map_dfr(function(f) deconv_pair_with_ihc(qs_scores, select(f, -scope)) |>
            mutate(scope = f$scope[1], .before = 1)) |>
  left_join(clin, by = "patient_id")

qs_rho <- deconv_scope_concordance(qs_paired) |>
  mutate(lineage = factor(lineage, levels = qs_lineages),
         scope   = factor(scope, levels = DECONV_SCOPES$scope))

qs_rho |>
  mutate(cell = ifelse(is.finite(rho), sprintf("%.2f (n = %d)", rho, n),
                       sprintf("NA (n = %d)", n)),
         scope = factor(scope_lab[as.character(scope)], levels = DECONV_SCOPES$label)) |>
  select(lineage, scope, cell) |>
  pivot_wider(names_from = scope, values_from = cell, names_sort = TRUE) |>
  arrange(lineage) |>
  knitr::kable(caption = "Spearman rho, quanTIseq score vs IHC fraction of all kept cells, per population and scope (n = patients).")
```

The same rho values as a grid: one tile per population (rows, in the gate tree's
order) and scope (columns), coloured on a diverging scale centred at 0. Each column
states the number of patients paired in that scope. With this few patients a tile
is a description of the ranking, not a test.

```{r quantiseq-concordance-heatmap, eval = have_pairs, fig.height = 5}
scope_n <- qs_paired |>
  group_by(scope) |>
  summarise(n = n_distinct(patient_id), .groups = "drop")
scope_axis <- stats::setNames(
  sprintf("%s\n(n = %d)", sub(" — ", "\n", scope_lab[scope_n$scope]), scope_n$n),
  scope_n$scope)

ggplot(qs_rho, aes(scope, fct_rev(lineage), fill = rho)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(aes(label = ifelse(is.na(rho), "·", sprintf("%.2f", rho))),
            size = pt_text(6.5)) +
  scale_x_discrete(labels = scope_axis, position = "top") +
  scale_fill_div(midpoint = 0, limits = c(-1, 1), name = "Spearman rho",
                 guide = guide_cbar()) +
  labs(title = "quanTIseq vs IHC: agreement by population and cell scope",
       subtitle = with_n("Spearman rho across patients, per scope",
                         qs_paired$patient_id, "patients"),
       x = NULL, y = NULL) +
  theme_paper_tile()
```

## One scatter per population {.tabset}

One tab per scope, and inside it one figure per population in the gate tree's
order. Each point is a patient: x is the IHC fraction of all cells the scope keeps
(FlowPath phenotypes from that scope's arm), y is the quanTIseq fraction from that
patient's bulk RNA. The blue line is a least-squares fit drawn to show direction
only. Points
are coloured by the clinical `Immuno-phenotype`. The title carries the population's
gate, the subtitle the scope, its Spearman rho and its n. Each plot is exported as
its own PDF.

```{r quantiseq-scatter, eval = have_pairs, results = "asis", fig.width = fig_width("single"), fig.height = 3.4}
for (s in DECONV_SCOPES$scope) {
  ds <- filter(qs_paired, scope == s)
  if (!nrow(ds)) next
  cat("\n\n### ", scope_lab[[s]], "\n\n", sep = "")
  for (lin in intersect(qs_lineages, unique(ds$lineage))) {
    print(plot_deconv_scope_pair(filter(ds, lineage == lin), s, lin, METHOD))
    cat("\n\n")
  }
}
```

## Single phenotype labels

Everything above compares **sums** of FlowPath phenotype labels: `CD8T` is
`T cytotoxic` + `Activated T cytotoxic` + `CD8+ T reg`. This section takes each
label on its own. `DECONV_LEAVES` (`code/deconv_scopes.R`) sets every label
against the quanTIseq score of the most specific population that contains it, so
several labels share one quanTIseq number: quanTIseq has a single CD8 fraction, the
phenotyping tree has three CD8 labels, and the panels show which label that number
follows. `Immune` is taken whole (not split by CD3 sign) and the four CD45− labels
all face quanTIseq's `uncharacterized cell`, which is the only non-immune term it
has.

```{r leaf-map}
DECONV_LEAVES |>
  left_join(select(deconv_comparison_lineages, lineage, ihc_gate), by = "lineage") |>
  knitr::kable(col.names = c("FlowPath phenotype label", "quanTIseq population on y",
                             "that population's gate"),
               caption = "Each phenotype label and the quanTIseq population it is correlated with.")
```

Per scope and patient, the label's fraction is the kept cells carrying exactly that
`phenotype` label over all kept cells (`ihc_leaf_fraction()`; a label a patient
lacks is 0). It is inner-joined on patient to the quanTIseq score of the label's
population, and each cell of the table is the cross-patient Spearman rho with its n.
Any phenotype label found in the cells but absent from the table above is named in
the printed line.

```{r leaf-concordance, eval = have_pairs}
leaf_unmapped <- unique(unlist(map(scope_ihc, "unmapped")))
cat("phenotype labels with no row in DECONV_LEAVES:",
    if (length(leaf_unmapped)) paste(leaf_unmapped, collapse = ", ") else "none", "\n")

leaf_paired <- scope_leaf |>
  inner_join(qs_scores, by = c("patient_id", "lineage")) |>
  filter(is.finite(score), is.finite(ihc_frac)) |>
  left_join(clin, by = "patient_id") |>
  mutate(leaf = factor(leaf, levels = DECONV_LEAVES$leaf))

deconv_scope_concordance(leaf_paired, by = "leaf") |>
  mutate(cell = ifelse(is.finite(rho), sprintf("%.2f (n = %d)", rho, n),
                       sprintf("NA (n = %d)", n)),
         scope = factor(scope_lab[scope], levels = DECONV_SCOPES$label)) |>
  select(leaf, lineage, scope, cell) |>
  pivot_wider(names_from = scope, values_from = cell, names_sort = TRUE) |>
  arrange(leaf) |>
  knitr::kable(col.names = c("phenotype label", "quanTIseq population",
                             levels(factor(scope_lab[unique(leaf_paired$scope)],
                                           levels = DECONV_SCOPES$label))),
               caption = "Spearman rho, quanTIseq population score vs the fraction of cells carrying one phenotype label, per label and scope (n = patients).")
```

### One scatter per phenotype label {.tabset}

One tab per scope, one figure per phenotype label. x is the fraction of the scope's
cells carrying that label, y the quanTIseq score of the population named in the
title; one point per patient, coloured by the clinical `Immuno-phenotype`, with a
least-squares fit for direction. Labels that share a population share their y
values, so only x differs between those figures.

```{r leaf-scatter, eval = have_pairs, results = "asis", fig.width = fig_width("single"), fig.height = 3.4}
for (s in DECONV_SCOPES$scope) {
  ds <- filter(leaf_paired, scope == s)
  if (!nrow(ds)) next
  cat("\n\n#### ", scope_lab[[s]], "\n\n", sep = "")
  for (lf in levels(droplevels(ds$leaf))) {
    dl <- filter(ds, leaf == lf)
    print(plot_deconv_scope_pair(dl, s, dl$lineage[1], METHOD, leaf = lf))
    cat("\n\n")
  }
}
```

## Notes

- Every statistic is a rank correlation over a handful of patients. Read the grid
  for pattern across scopes, not for significance.
- The populations overlap (CD8T sits inside Cytotoxic, T_total and Immune_total),
  so their rho values are not independent.
- `Immune_other` puts the sum of quanTIseq's B, macrophage, monocyte, dendritic and
  neutrophil fractions against every CD45+ CD3− CD56− cell; `NonImmune` puts
  quanTIseq's `uncharacterized cell` against every CD45− cell.
- An annotation scope changes the IHC side only. The bulk RNA is one sample per
  patient and is the same in all four scopes.
- `massimo1_inverted` has no scope here: it re-classifies `massimo1`'s cells and
  borrows its polygons, so it is a third phenotyping rather than a third annotation.

```{r export-pdfs, include=FALSE}
# Every plot above is also rendered to PDF by the dev = c("png", "pdf") chunk option;
# this collects them into output/figures/<SLUG>/. Runs last, so they all exist by now.
export_pdf_figures(SLUG)
```
