---
title: "Paper figures — panel assembly sheet"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(readxl)     # not attached by tidyverse; the clinical table is .xlsx
})

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/
source(here("code", "pdf_export.R"))

SLUG <- "paper_figures"

# Every panel below is guarded on its own input, so this page renders on a machine
# that has none of them and tells you exactly what is missing. That is the point of
# the page: it is a checklist you can knit.
suppressWarnings(suppressMessages({
  have_helpers <- tryCatch({ source(here("code", "validation_helpers.R")); TRUE },
                           error = function(e) FALSE)
  if (have_helpers) source(here("code", "paper_figures.R"))
  if (have_helpers) source(here("code", "all_slide.R"))
}))

# Where a hand-made panel (screenshot, exported schematic) is dropped in.
MANUAL_DIR <- here("output", "figures", "manual")
dir.create(MANUAL_DIR, recursive = TRUE, showWarnings = FALSE)
```

This page is **not an analysis**. It is the sheet the manuscript figures are cut
from: one panel per plot, at publication proportions, exported as vector PDFs to
`output/figures/paper_figures/` so they can be placed and labelled by hand in
Affinity. No panel here computes anything a analysis page does not already
compute — each is a re-cut of an existing quantity in the shape its legend asks
for. Where a panel *cannot* be computed (a screenshot, a microscope crop, a
schematic) the page says so, says exactly how to make it, and picks it up
automatically once you drop the file in.

**Two kinds of panel.**

- **Computed** — built here from data and exported as a PDF. If the data is
  absent the panel says which file it wanted.
- **To supply** — a screenshot or an image from the imaging team. Save it as
  `output/figures/manual/<panel-id>.png` (or `.pdf`) and re-knit; it appears
  inline below its instructions, and the instructions disappear.

```{r panel-slot}
# A panel that cannot be computed. Prints the acquisition instructions until the
# file exists, then prints the file instead. The id IS the filename, so there is
# never a question of what to name a screenshot — the page tells you, and the same
# string is what it looks for.
#
# Deliberately checks several extensions: a QuPath screenshot is a PNG, an exported
# schematic is a PDF, and a microscope crop may be either. Whichever arrives first
# is the one used.
panel_slot <- function(id, what, howto, source = NULL) {
  exts  <- c("png", "pdf", "jpg", "jpeg", "tif", "tiff", "svg")
  found <- file.path(MANUAL_DIR, paste0(id, ".", exts))
  found <- found[file.exists(found)]

  cat("\n\n**Panel `", id, "` — ", what, "**\n\n", sep = "")
  if (length(found)) {
    # Route the supplied file through knitr's own fig.path. A markdown link
    # straight at output/figures/manual/ would resolve while knitting and break in
    # the built site, because workflowr writes the HTML into docs/ and nothing
    # outside docs/ is published. Copying it here makes the panel a first-class
    # site figure that travels with the page.
    dest_dir <- knitr::opts_chunk$get("fig.path")
    ok <- FALSE
    if (!is.null(dest_dir) && nzchar(dest_dir)) {
      dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
      dest <- file.path(sub("/+$", "", dest_dir), basename(found[1]))
      ok   <- file.copy(found[1], dest, overwrite = TRUE)
      if (ok) cat("![", what, "](", dest, ")\n\n", sep = "")
    }
    if (!ok) cat("_Supplied but not copied into the site:_ `",
                 file.path("output", "figures", "manual", basename(found[1])), "`\n\n", sep = "")
    cat("_Source file:_ `output/figures/manual/", basename(found[1]), "`\n\n", sep = "")
    return(invisible(TRUE))
  }
  cat("> **TO SUPPLY.** Save as `output/figures/manual/", id, ".png` (or `.pdf`) and re-knit.\n>\n", sep = "")
  for (line in howto) cat(">  - ", line, "\n", sep = "")
  if (!is.null(source)) cat(">\n> _Source:_ ", source, "\n", sep = "")
  cat("\n")
  invisible(FALSE)
}
```

```{r load-cells}
# The cell source for every spatial and immune-fraction panel: the all-slide
# layout, which is the only export where the pathologist polygon and the cell table
# are matched region by region. See code/all_slide.R for the layout and for the
# "no annotation directory means everything is inside" rule.
as_cells <- tibble::tibble(); as_polys <- NULL; as_per <- tibble::tibble()
as_union <- tibble::tibble()
have_cells <- FALSE; geom_failure <- NULL

if (have_helpers && dir.exists(file.path(ALL_SLIDE_DIR, "csv"))) {
  as_cells <- all_slide_cells(ALL_SLIDE_DIR)
  if (nrow(as_cells)) {
    # all_slide_annotations() errors without sf, on purpose: a script that asked for
    # geometry should not quietly get something else. This page is a checklist and
    # has to knit anywhere, so it catches that and says what was lost — the panels
    # then fall back to the export flag, which is a DIFFERENT membership rule and
    # therefore reported, not absorbed.
    as_polys <- tryCatch(
      all_slide_annotations(ALL_SLIDE_DIR, patient_ids = unique(as_cells$patient_id)),
      error = function(e) { geom_failure <<- conditionMessage(e); NULL })
    as_per     <- all_slide_metrics(as_cells, as_polys, "per_annotation")
    as_union   <- all_slide_metrics(as_cells, as_polys, "union")
    have_cells <- TRUE
  }
}

if (have_cells) {
  if (!is.null(geom_failure))
    cat("> **Polygons were not read**, so membership below falls back to each",
        "export's own out-of-annotation flag and every density/area column is `NA`.",
        "Reason: ", geom_failure, "\n\n")
  else if (is.null(as_polys))
    cat("> **No polygons found** — every patient is treated as whole-slide.\n\n")
  knitr::kable(all_slide_inventory(as_cells),
               caption = "What was read from data/all_slide/ — one row per patient x region.")
} else {
  cat("**No cells.** This page expects the all-slide export at `data/all_slide/`:\n\n",
      "```\ndata/all_slide/csv/<patient>/<patient>_<A|B|C>.csv\n",
      "data/all_slide/annotation/<patient>/<patient>_<A|B|C>.geojson\n```\n\n",
      "On the cluster that tree is\n",
      "`/hpcnfs/techunits/imaging/work/ATTEND/Mirage/all-slide_new` — symlink or copy it:\n\n",
      "```sh\nln -s /hpcnfs/techunits/imaging/work/ATTEND/Mirage/all-slide_new data/all_slide\n```\n",
      sep = "")
}
```

---

# Figure 2 — the MIRAGE pipeline

## 2(a) Pipeline stage graph

**Already drawn, and better than a redraw.** `mirage/docs/figures/pipeline-schematic.html`
is Supplementary Figure S1: publication-shaped, self-contained, with every module's
defaults as chips. Export it to PDF from a browser at A4 width and place it directly.

Two things in the drafted legend contradict that schematic and need fixing in the
*text*, not the figure:

- **There are four Nextflow steps, not five.** `ParamUtils.STEPS` and
  `Layout.CHECKPOINT_STEPS` both define `preprocessing → registration →
  segmentation → postprocessing`. Quantification and GeoJSON export are both
  *inside* postprocessing.
- **The anchor is a configurable nuclear-counterstain channel**, not DAPI by
  construction — S1's own samplesheet shows `nuclear_markers = [DAPI, CELLTOX]`.
  "DAPI throughout this study" is the accurate phrasing.

```{r fig2a, results="asis"}
panel_slot(
  "fig2a_pipeline_schematic",
  "MIRAGE five-module / four-step pipeline graph",
  c("Open `mirage/docs/figures/pipeline-schematic.html` in a browser.",
    "Print to PDF, A4 landscape, background graphics ON, margins none.",
    "Crop to the panel you want — the sheet holds several; Fig 2(a) is the step ladder plus the module table."),
  source = "`mirage/docs/figures/pipeline-schematic.html` (Supplementary Figure S1)")
```

## 2(b) Illumination correction across a tile seam

```{r fig2b, results="asis"}
panel_slot(
  "fig2b_basic_seam",
  "Tile-boundary crop, before vs after BaSiC flatfield/darkfield correction",
  c("**Do not use the pipeline's preprocess QC PNGs.** They are written at `preprocess_qc_scale_factor = 0.25`; a tile seam is a few-pixel intensity step and does not survive a quarter-scale downsample.",
    "Take matched full-resolution crops at the SAME coordinates from the converted (pre-BaSiC) and preprocessed (post-BaSiC) OME-TIFFs.",
    "Centre the crop on a mosaicing seam — a visible grid line in the uncorrected image.",
    "Use one channel, one display range, identical for both crops. A different contrast stretch would manufacture the effect.",
    "Record the pixel size from the OME-TIFF metadata so the scale bar is real."),
  source = "imaging team — needs the raw and preprocessed images, not published QC output")
```

---

# Figure 3 — MIRAGE → FlowPath hand-off

## 3(a) Interop hand-off

```{r fig3a, results="asis"}
panel_slot(
  "fig3a_interop_handoff",
  "MIRAGE → cells.geojson → FlowPath, the hand-off only",
  c("Three boxes, not a stage graph — Fig 2(a) is the stage graph and this panel must not repeat it.",
    "Show the artefact in the middle: `cells.geojson`, geometry + per-cell features in one file.",
    "Draw it in the S1/S2/S3 house style so the figure set reads as one artefact — copy the shared CSS block verbatim from any of `mirage/docs/figures/*.html` (it is duplicated across all three on purpose).",
    "A real excerpt of `qupath-extension-flowpath/cells.geojson` makes a good inset."),
  source = "new; `qupath-extension-flowpath/ARCHITECTURE.md` and `mirage/docs/spatialdata.md` describe the contract")
```

## 3(b) Interactive gating on MIRAGE-derived features

```{r fig3b, results="asis"}
panel_slot(
  "fig3b_gating_threshold",
  "One user-drawn gate on a threshold plot of MIRAGE per-cell features",
  c("QuPath ≥ 0.7.0, cells loaded, then `Extensions → FlowPath - GatingTree` (Ctrl+G).",
    "Add a root gate → threshold (1D). Pick a marker with clean separation — CD45 is the panel's cleanest.",
    "Drag the threshold handle so it is visibly ON the histogram; the panel's claim is that gating is interactive, so the handle must be in frame.",
    "Capture the histogram pane, not the whole application window.",
    "The catalog docs currently ship `assets/screenshots/placeholder.png` for exactly this figure — nothing exists yet, this is yours to take."),
  source = "screenshot — FlowPath GatingTree")
```

## 3(c) Sequential hierarchical phenotype tree

```{r fig3c, results="asis"}
panel_slot(
  "fig3c_phenotype_tree",
  "The sequential gating scheme and its terminal populations",
  c("Draw from the ACTUAL `flowpath.json` used for the study, not from memory.",
    "Sequential, not Boolean: a CD45-only pan-immune root, then the T/NK axis.",
    "Shade the CD45+CD3−CD56− catch-all node — the coverage caveat in the legend depends on it being visually marked.",
    "Reconcile the terminal-population count against the table printed below before writing a number into the legend."),
  source = "the study's `flowpath.json`")
```

The project's own record of what the gate tree emits is below. It is the join
table every analysis reads phenotypes through, so **if the figure and this table
disagree, the analyses are keyed to this one** — fix the code or the figure, but
do not leave them different.

```{r fig3c-table}
if (have_helpers) {
  fp <- dplyr::filter(phenotype_lineage_labels,
                      !grepl("_", phenotype_clean))     # FlowPath spellings only
  knitr::kable(fp, caption = paste(
    "FlowPath gate-tree labels as recorded in code/cell_tables.R, and the lineage",
    "each collapses to. Myeloid/Macrophage_M2 appear on the mirage side only —",
    "check the study panel.yaml before repeating the legend's 'no myeloid markers' caveat."))
}
```

## 3(d) Spatial phenotype map — **computed**

Cells coloured by phenotype call, in the geojson pixel frame, with the
pathologist's polygon outlined. The whole-slide panel and its inset are exported
as **separate PDFs** on purpose — place, scale and label them in Affinity rather
than accepting a composite baked in R.

```{r fig3d-map, fig.width=6.2, fig.height=5.2}
map_pid <- NULL
if (have_cells) {
  # The exemplar: whichever patient has the most cells, so the map is dense enough
  # to read at figure size. Override by hand when the paper wants a specific case.
  map_pid <- as_cells |> count(patient_id, sort = TRUE) |> slice(1) |> pull(patient_id)
  p <- paper_phenotype_map(as_cells, patient_id = map_pid, annots = as_polys,
                           title = paste("Patient", map_pid, "— phenotype calls"))
  if (!is.null(p)) print(p)
}
if (is.null(map_pid)) cat("_No cells — see the loader note at the top of the page._\n")
```

```{r fig3d-inset, fig.width=4.4, fig.height=4.0}
# The inset comes from the SAME function, so it cannot drift from the map above.
# The window is a fixed fraction of the field centred on the densest patch: an
# inset that lands on empty slide shows nothing, and choosing it by eye each time
# makes the figure unreproducible.
if (have_cells && !is.null(map_pid)) {
  cp <- dplyr::filter(as_cells, slide_key(patient_id) == slide_key(map_pid))
  xy <- cell_centroids_px(cp)
  ok <- is.finite(xy$x) & is.finite(xy$y)
  if (any(ok)) {
    cx <- median(xy$x[ok]); cy <- median(xy$y[ok])
    w  <- diff(range(xy$x[ok])) * 0.12; h <- diff(range(xy$y[ok])) * 0.12
    p <- paper_phenotype_map(as_cells, patient_id = map_pid, annots = as_polys,
                             zoom = c(cx - w, cx + w, cy - h, cy + h),
                             title = paste0("Patient ", map_pid, " — inset"))
    if (!is.null(p)) print(p)
  }
}
```

---

# Figure 4 — cross-panel registration accuracy

## 4(a) Two-colour DAPI overlay, before and after

```{r fig4a, results="asis"}
panel_slot(
  "fig4a_dapi_overlay",
  "Two-colour DAPI overlay of two separately-acquired panels, before vs after registration",
  c("The pipeline already makes this: `GENERATE_REGISTRATION_QC` emits DAPI-overlay QC whenever `reg_qc >= 1`, which is the default (`reg_qc = 2`).",
    "But it writes at `qc_scale_factor = 0.25`. For a 'nuclei superimposed' claim, take a full-resolution crop at matched coordinates from the pre- and post-registration images instead.",
    "Two colours, one panel per channel, same display range before and after.",
    "Pick a field where the pre-registration drift is visible at figure size — a corner usually drifts more than the centre.",
    "Record the pixel size for the scale bar."),
  source = "mirage `GENERATE_REGISTRATION_QC` output, or full-res crops from the same run")
```

## 4(b) and 4(c) Target registration error and Dice, by configuration — **computed**

The arm-ranking panels. There are **two possible sources**, and they are not
equivalent:

1. **The real arm sweep** (`data/registration_arms/`) — the study slides registered
   once per configuration. This is what the figure should use.
2. **The synthetic benchmark sweep** (`data/benchmark/param_matrix.csv`) — mirage's
   `benchmarks/` runs, whose moving panels are shifted copies of channel 0. A
   residual measured there reflects a known injected offset, not real
   cross-acquisition difficulty.

This chunk prefers (1) and falls back to (2), and says which it used — because
which one produced the figure changes what the figure is allowed to claim.

```{r fig4bc, fig.width=7.5, fig.height=4.8}
reg_dir  <- here("data", "benchmark")
arm_figs <- list(); arm_source <- NULL

if (have_helpers && dir.exists(here("data", "registration_arms"))) {
  suppressMessages(source(here("code", "registration_arms.R")))
  .man <- arm_manifest()
  .seg <- if (nrow(.man)) read_arms_seg_qc(.man) else tibble::tibble()
  if (nrow(.seg)) {
    arm_figs   <- build_arm_figs(.seg, tibble::tibble(), .man)[c("01_final_residual_um_by_arm",
                                                                "02_final_dice_by_arm")]
    arm_source <- "real"
  }
}
if (!length(arm_figs) && have_helpers && file.exists(file.path(reg_dir, "param_matrix.csv"))) {
  source(here("code", "registration_accuracy_plots.R"))
  arm_figs   <- build_reg_figs(reg_dir)[c("06_tre_by_arm", "07_dice_by_arm")]
  arm_source <- "synthetic"
}
arm_figs <- arm_figs[!vapply(arm_figs, is.null, logical(1))]
for (p in arm_figs) print(p)
```

```{r fig4bc-note, results="asis"}
if (identical(arm_source, "real")) {
  cat("**Source: the real arm sweep** on the study slides — the right source for",
      "the manuscript. Arms are ranked at each run's FINAL transform, because",
      "`reg_micro_reg` 0 and 1 emit no `micro` stage and ranking on a fixed stage",
      "would drop those arms. See [registration arms](registration_arms.html) for",
      "the stage ladder, the pair-fraction trust check, and why the `rigid` stage",
      "cannot be compared across depths.\n")
} else if (identical(arm_source, "synthetic")) {
  cat("> **Source: the SYNTHETIC sweep.** These arms come from",
      "`data/benchmark/param_matrix.csv`, whose moving panels are shifted copies of",
      "channel 0. Usable for cost and scaling; **not** a real-tissue accuracy claim.",
      "Put the per-arm runs under `data/registration_arms/` and re-knit to replace",
      "this panel with the real one.\n")
} else {
  cat("_No arm data._ Either symlink the per-arm runs under",
      "`data/registration_arms/<arm>/` (preferred — see",
      "[registration arms](registration_arms.html)), or drop mirage's `paper_data`",
      "CSVs into `data/benchmark/`.\n")
}
```

> **Two legend corrections this sweep forces.**
>
> - The presets are **not** "two accuracy presets differing in feature-matching
>   resolution, not in the feature matcher". `memory_mode` selects the matcher:
>   `low` = BRISK/RANSAC, `high` = SuperPoint/SuperGlue. It is a different matcher.
> - Micro-registration is **not** on/off. `reg_micro_reg` is a nested depth
>   (`0` none, `1` micro-rigid, `2` + micro non-rigid), so a preset × depth sweep has
>   **six** arms, not four.
>
> **And there is no clean "rigid-only" baseline in a depth-crossed sweep.** mirage's
> `rigid` QC stage means affine ∘ micro-rigid whenever `reg_micro_reg ≥ 1`, so a
> rigid-only baseline exists only in the depth-0 arms. The `native` stage is the
> honest no-registration baseline.

---

# Figure 5 — proof-of-concept concordance

## 5(a) Registered, phenotyped cases — **computed**

The same renderer as 3(d), one panel per case. Exported one PDF per case.

```{r fig5a, fig.width=5.4, fig.height=4.6}
if (have_cells) {
  for (pid in sort(unique(as_cells$patient_id))) {
    p <- paper_phenotype_map(as_cells, patient_id = pid, annots = as_polys,
                             title = paste("Patient", pid))
    if (!is.null(p)) print(p)
  }
}
```

## 5(b) Imaging immune fraction, hot vs cold — **computed**

```{r fig5b-groups}
# WHICH VARIABLE DEFINES HOT AND COLD. The clinical table carries two different
# things: a continuous `HOT score` and a categorical `Immuno-phenotype`. The
# legend's selection axis is the score, so the score is used here and the top/bottom
# three cases become the groups. That IS a threshold — the legend cannot also claim
# "no threshold" while showing two boxes. Set HOTCOLD_SOURCE to "immuno_phe" to
# group by the clinical category instead.
HOTCOLD_SOURCE <- "hot_score"

groups <- NULL
clin_path <- here("data", "clinical_data.xlsx")
if (file.exists(clin_path) && have_cells) {
  # Same idiom as molecular_hot_cold.Rmd: select(any_of(...)) so a clinical table missing
  # either column loses that column rather than aborting the chunk.
  clin <- readxl::read_excel(clin_path) |>
    dplyr::filter(!is.na(`ID PATIENT`)) |>
    dplyr::mutate(patient_id = slide_key(`ID PATIENT`)) |>
    dplyr::select(patient_id,
                  hot_score  = dplyr::any_of("HOT score"),
                  immuno_phe = dplyr::any_of("Immuno-phenotype")) |>
    dplyr::filter(patient_id %in% slide_key(as_union$patient_id))
  if ("hot_score" %in% names(clin))
    clin$hot_score <- suppressWarnings(as.numeric(clin$hot_score))

  if (HOTCOLD_SOURCE == "hot_score" && "hot_score" %in% names(clin) &&
      sum(is.finite(clin$hot_score)) >= 2) {
    k <- min(3, floor(sum(is.finite(clin$hot_score)) / 2))
    r <- rank(-clin$hot_score, ties.method = "first")
    groups <- clin |>
      dplyr::mutate(group = dplyr::case_when(r <= k ~ "hot",
                                             r > dplyr::n() - k ~ "cold",
                                             TRUE ~ NA_character_)) |>
      dplyr::filter(!is.na(group)) |>
      dplyr::select(patient_id, group)
  } else if ("immuno_phe" %in% names(clin)) {
    groups <- clin |> dplyr::transmute(patient_id, group = immuno_phe) |>
      dplyr::filter(!is.na(group))
  }
}
if (!is.null(groups)) knitr::kable(groups, caption = "Cases as grouped for panel 5(b).")
```

```{r fig5b, fig.width=3.4, fig.height=3.6}
if (!is.null(groups) && nrow(as_union) > 0) {
  p <- paper_immune_fraction_hotcold(as_union, groups)
  if (!is.null(p)) print(p)
} else {
  cat("_Needs `data/clinical_data.xlsx` (for the hot/cold axis) and the all-slide cells._\n")
}
```

No test is applied and no coefficient is printed — with three cases a side the six
points *are* the evidence, and `paper_immune_fraction_hotcold()` deliberately never
calls `paired_spearman()`. Every other page on this site reports rho freely; this
panel is the one place that must not, because its legend says so.

## 5(c) Imaging vs bulk-RNA deconvolution — **computed**

```{r fig5c, fig.width=5.6, fig.height=4.6}
paired_path <- here("output", "paired_deconv.rds")
if (have_helpers && file.exists(paired_path)) {
  p <- paper_deconv_scatter(readRDS(paired_path), method = "quantiseq")
  if (!is.null(p)) print(p)
} else {
  cat("_Needs `output/paired_deconv.rds`._ Knit [the molecular page](molecular_hot_cold.html)",
      "first — its last chunk caches the paired frame so this panel does not have to",
      "re-run immunedeconv.\n")
}
```

No fit line and no coefficient, for the reason above and one more: the axes use
different denominators (imaging counts cells, deconvolution estimates a mixture
fraction), so a regression line would invite exactly the absolute-agreement
reading the legend disclaims. **The populations plotted are fixed by
`deconv_to_lineage()` and there are four** — CD4T, CD8T, Treg, NK. That answers
the legend's `[author to supply: full list of populations plotted]`.

---

# Additional files

```{r addfile-2}
if (have_helpers && file.exists(file.path(reg_dir, "param_matrix.csv"))) {
  pm <- readr::read_csv(file.path(reg_dir, "param_matrix.csv"), show_col_types = FALSE)
  at <- .reg_arm_table(pm)
  if (!is.null(at))
    knitr::kable(dplyr::arrange(at, tre), digits = 4, caption = paste(
      "Additional file 2 — per-arm registration accuracy.",
      "TRE and DAPI-overlap Dice for every benchmarked configuration."))
} else {
  cat("**Additional file 2** (per-arm TRE and Dice table) needs `data/benchmark/param_matrix.csv`.\n")
}
```

```{r addfile-4}
if (have_helpers)
  knitr::kable(paper_lineage_table(), caption = paste(
    "Additional file 4 — the imaging-phenotype and deconvolution-cell-type",
    "vocabularies, and the lineage each maps to. Generated from the same objects",
    "the analyses join on, so it cannot drift from what was computed. The six",
    "'(unmapped)' deconvolution types are precisely the catch-all the legend",
    "calls Tier 3."))
```

**Additional files 1, 3, 5 and 6.**

- **1 — full pipeline configuration.** `mirage/docs/parameters.md` is already the
  complete per-module table. Missing from it: the compartment-to-marker map and a
  default-versus-recommended comparison; both still have to be written.
- **3 — Tier 2 / Tier 3 scatters.** Blocked: the tier vocabulary does not exist in
  this codebase. Its content maps onto `cd45_over_inside` (Tier 1),
  `deconv_to_lineage()` (Tier 2) and `Immune_other` (Tier 3) — settle the naming,
  then this is a re-cut of the table above.
- **5 — epitope / tissue-integrity monitoring.** Per-round DAPI retention exists
  nowhere; imaging-team deliverable. For the CD8+ ⊂ CD3+ biological-consistency check,
  the machinery is [marker QC](marker_qc.html), which computes worst-case
  conditional co-positivity per marker pair straight from the cell tables. Two things
  to note: it is **exclusivity**-shaped (it asks whether two markers co-occur *more*
  than they should) while CD8 ⊂ CD3 is an **inclusion** constraint (CD8+CD3− should be
  rare), so it needs a negated-marker pair added; and it reads the exports directly, so
  unlike mirage-side QC it works on data this project actually has.
- **6 — computational performance.** Fully covered by
  [pipeline benchmarks](benchmark_pipeline.html); nothing to add.

---

## Exported files

```{r manifest}
out_dir <- here("output", "figures", SLUG)
if (dir.exists(out_dir)) {
  f <- list.files(out_dir, pattern = "\\.pdf$")
  if (length(f)) knitr::kable(tibble::tibble(panel = f),
                              caption = "Vector PDFs written for Affinity.")
}
cat("\nComputed panels land in `output/figures/", SLUG, "/`.",
    " Hand-made panels go in `output/figures/manual/` named as the slots above.\n", sep = "")
```

```{r export-pdfs, include=FALSE}
export_pdf_figures(SLUG)
```
