---
title: "Paper figures — panel assembly sheet"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(tidyr)
  library(readr)
  library(purrr)
  library(ggplot2)
  library(stringr)
  library(forcats)
  library(here)
  library(readxl)     # not attached by tidyverse; the clinical table is .xlsx
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the analysis script that used to pull plot_theme.R in never runs —
# which silently left those plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/
source(here("code", "pdf_export.R"))

SLUG <- "paper_figures"

# Every panel below is guarded on its own input, so this page renders on a machine
# that has none of them and tells you exactly what is missing. That is the point of
# the page: it is a checklist you can knit.
suppressWarnings(suppressMessages({
  have_helpers <- tryCatch({ source(here("code", "validation_helpers.R")); TRUE },
                           error = function(e) FALSE)
  if (have_helpers) source(here("code", "paper_figures.R"))
  if (have_helpers) source(here("code", "arm_cells.R"))
}))

# Where a hand-made panel (screenshot, exported schematic) is dropped in.
MANUAL_DIR <- here("output", "figures", "manual")
dir.create(MANUAL_DIR, recursive = TRUE, showWarnings = FALSE)
```

This page is **not an analysis**. It is the sheet the manuscript figures are cut
from: one panel per plot, at publication proportions, exported as vector PDFs to
`output/figures/paper_figures/` so they can be placed and labelled by hand in
Affinity. No panel here computes anything a analysis page does not already
compute — each is a re-cut of an existing quantity in the shape its legend asks
for. Where a panel *cannot* be computed (a screenshot, a microscope crop, a
schematic) the page says so, says exactly how to make it, and picks it up
automatically once you drop the file in.

**Two kinds of panel.**

- **Computed** — built here from data and exported as a PDF. If the data is
  absent the panel says which file it wanted.
- **To supply** — a screenshot or an image from the imaging team. Save it as
  `output/figures/manual/<panel-id>.png` (or `.pdf`) and re-knit; it appears
  inline below its instructions, and the instructions disappear.

```{r panel-slot}
# A panel that cannot be computed. Prints the acquisition instructions until the
# file exists, then prints the file instead. The id IS the filename, so there is
# never a question of what to name a screenshot — the page tells you, and the same
# string is what it looks for.
#
# Deliberately checks several extensions: a QuPath screenshot is a PNG, an exported
# schematic is a PDF, and a microscope crop may be either. Whichever arrives first
# is the one used.
panel_slot <- function(id, what, howto, source = NULL) {
  exts  <- c("png", "pdf", "jpg", "jpeg", "tif", "tiff", "svg")
  found <- file.path(MANUAL_DIR, paste0(id, ".", exts))
  found <- found[file.exists(found)]

  cat("\n\n**Panel `", id, "` — ", what, "**\n\n", sep = "")
  if (length(found)) {
    # Route the supplied file through knitr's own fig.path. A markdown link
    # straight at output/figures/manual/ would resolve while knitting and break in
    # the built site, because workflowr writes the HTML into docs/ and nothing
    # outside docs/ is published. Copying it here makes the panel a first-class
    # site figure that travels with the page.
    dest_dir <- knitr::opts_chunk$get("fig.path")
    ok <- FALSE
    if (!is.null(dest_dir) && nzchar(dest_dir)) {
      dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)
      dest <- file.path(sub("/+$", "", dest_dir), basename(found[1]))
      ok   <- file.copy(found[1], dest, overwrite = TRUE)
      if (ok) cat("![", what, "](", dest, ")\n\n", sep = "")
    }
    if (!ok) cat("_Supplied but not copied into the site:_ `",
                 file.path("output", "figures", "manual", basename(found[1])), "`\n\n", sep = "")
    cat("_Source file:_ `output/figures/manual/", basename(found[1]), "`\n\n", sep = "")
    return(invisible(TRUE))
  }
  cat("> **TO SUPPLY.** Save as `output/figures/manual/", id, ".png` (or `.pdf`) and re-knit.\n>\n", sep = "")
  for (line in howto) cat(">  - ", line, "\n", sep = "")
  if (!is.null(source)) cat(">\n> _Source:_ ", source, "\n", sep = "")
  cat("\n")
  invisible(FALSE)
}
```

```{r load-cells}
# The cell source for every spatial and immune-fraction panel: the all-slide
# layout, where the pathologist polygon and the cell table are matched region by
# region. See code/arms.R for the three arms and code/arm_cells.R for the layout
# rules, including "no annotation directory means everything is inside".
#
# The manuscript panels are built from ONE arm, named here. They cannot be pooled
# across arms: the arms' regions are drawn independently, so an ANNOTATION_2 row from
# each is a different polygon and stacking them would average regions that do not
# overlap.
ARM      <- "massimo2"
as_spec  <- if (have_helpers) arm_spec(ARM) else NULL

as_cells <- tibble::tibble(); as_polys <- NULL; as_per <- tibble::tibble()
as_union <- tibble::tibble(); as_ucells <- tibble::tibble()
have_cells <- FALSE; geom_failure <- NULL

if (have_helpers && dir.exists(as_spec$region_csv$path)) {
  as_cells  <- arm_cells(as_spec)
  as_ucells <- arm_union_tier_cells(as_spec)
  if (nrow(as_cells)) {
    # arm_annotations() errors without sf, on purpose: a script that asked for
    # geometry should not quietly get something else. This page is a checklist and
    # has to knit anywhere, so it catches that and says what was lost — the panels
    # then fall back to the export flag, which is a DIFFERENT membership rule and
    # therefore reported, not absorbed.
    # An arm with no union tier hands back an EMPTY tibble with no columns, and
    # `$patient_id` on that warns on every knit.
    pids <- unique(c(as_cells$patient_id,
                     if ("patient_id" %in% names(as_ucells)) as_ucells$patient_id))
    as_polys <- tryCatch(arm_annotations(as_spec, "region", patient_ids = pids),
      error = function(e) { geom_failure <<- conditionMessage(e); NULL })
    as_upoly <- tryCatch(arm_annotations(as_spec, "union", patient_ids = pids),
      error = function(e) NULL)
    .prom    <- arm_promote_unregioned(as_spec, as_cells, as_polys, as_ucells, as_upoly)
    as_cells <- .prom$cells; as_polys <- .prom$polys
    as_per   <- arm_metrics(as_spec, as_cells, as_polys, "per_annotation")
    as_union <- arm_metrics(as_spec, as_cells, as_polys, "union",
                            union_cells = as_ucells, union_polys = as_upoly)
    have_cells <- TRUE
  }
}

if (have_cells) {
  if (!is.null(geom_failure))
    cat("> **Polygons were not read**, so membership below falls back to each",
        "export's own out-of-annotation flag and every density/area column is `NA`.",
        "Reason: ", geom_failure, "\n\n")
  else if (is.null(as_polys))
    cat("> **No polygons found** — every patient is treated as whole-slide.\n\n")
  knitr::kable(arm_inventory(as_cells),
               caption = paste0("What was read from data/", as_spec$root,
                                "/ — one row per patient x region."))
} else {
  cat("**No cells.** This page expects arm `", ARM, "` at `data/", ARM, "/`:\n\n",
      "```\ndata/massimo2/csv/<patient>/<patient>_<A|B|C>.csv\n",
      "data/massimo2/annotation/<patient>/<patient>_<A|B|C>.geojson\n```\n\n",
      "On the cluster these trees are under the ATTEND share — symlink them:\n\n",
      "```sh\nln -s <share>/Massimo2 data/massimo2\n",
      "ln -s <share>/Massimo1 data/massimo1\n```\n",
      sep = "")
}
```

---

# Figure 2 — the MIRAGE pipeline

## 2(a) Pipeline stage graph

**Already drawn, and better than a redraw.** `mirage/docs/figures/pipeline-schematic.html`
is Supplementary Figure S1: publication-shaped, self-contained, with every module's
defaults as chips. Export it to PDF from a browser at A4 width and place it directly.

Two things in the drafted legend contradict that schematic and need fixing in the
*text*, not the figure:

- **There are four Nextflow steps, not five.** `ParamUtils.STEPS` and
  `Layout.CHECKPOINT_STEPS` both define `preprocessing → registration →
  segmentation → postprocessing`. Quantification and GeoJSON export are both
  *inside* postprocessing.
- **The anchor is a configurable nuclear-counterstain channel**, not DAPI by
  construction — S1's own samplesheet shows `nuclear_markers = [DAPI, CELLTOX]`.
  "DAPI throughout this study" is the accurate phrasing.

```{r fig2a, results="asis"}
panel_slot(
  "fig2a_pipeline_schematic",
  "MIRAGE five-module / four-step pipeline graph",
  c("Open `mirage/docs/figures/pipeline-schematic.html` in a browser.",
    "Print to PDF, A4 landscape, background graphics ON, margins none.",
    "Crop to the panel you want — the sheet holds several; Fig 2(a) is the step ladder plus the module table."),
  source = "`mirage/docs/figures/pipeline-schematic.html` (Supplementary Figure S1)")
```

## 2(b) Illumination correction across a tile seam

```{r fig2b, results="asis"}
panel_slot(
  "fig2b_basic_seam",
  "Tile-boundary crop, before vs after BaSiC flatfield/darkfield correction",
  c("**Do not use the pipeline's preprocess QC PNGs.** They are written at `preprocess_qc_scale_factor = 0.25`; a tile seam is a few-pixel intensity step and does not survive a quarter-scale downsample.",
    "Take matched full-resolution crops at the SAME coordinates from the converted (pre-BaSiC) and preprocessed (post-BaSiC) OME-TIFFs.",
    "Centre the crop on a mosaicing seam — a visible grid line in the uncorrected image.",
    "Use one channel, one display range, identical for both crops. A different contrast stretch would manufacture the effect.",
    "Record the pixel size from the OME-TIFF metadata so the scale bar is real."),
  source = "imaging team — needs the raw and preprocessed images, not published QC output")
```

---

# Figure 3 — MIRAGE → FlowPath hand-off

## 3(a) Interop hand-off

```{r fig3a, results="asis"}
panel_slot(
  "fig3a_interop_handoff",
  "MIRAGE → cells.geojson → FlowPath, the hand-off only",
  c("Three boxes, not a stage graph — Fig 2(a) is the stage graph and this panel must not repeat it.",
    "Show the artefact in the middle: `cells.geojson`, geometry + per-cell features in one file.",
    "Draw it in the S1/S2/S3 house style so the figure set reads as one artefact — copy the shared CSS block verbatim from any of `mirage/docs/figures/*.html` (it is duplicated across all three on purpose).",
    "A real excerpt of `qupath-extension-flowpath/cells.geojson` makes a good inset."),
  source = "new; `qupath-extension-flowpath/ARCHITECTURE.md` and `mirage/docs/spatialdata.md` describe the contract")
```

## 3(b) Interactive gating on MIRAGE-derived features

```{r fig3b, results="asis"}
panel_slot(
  "fig3b_gating_threshold",
  "One user-drawn gate on a threshold plot of MIRAGE per-cell features",
  c("QuPath ≥ 0.7.0, cells loaded, then `Extensions → FlowPath - GatingTree` (Ctrl+G).",
    "Add a root gate → threshold (1D). Pick a marker with clean separation — CD45 is the panel's cleanest.",
    "Drag the threshold handle so it is visibly ON the histogram; the panel's claim is that gating is interactive, so the handle must be in frame.",
    "Capture the histogram pane, not the whole application window.",
    "The catalog docs currently ship `assets/screenshots/placeholder.png` for exactly this figure — nothing exists yet, this is yours to take."),
  source = "screenshot — FlowPath GatingTree")
```

## 3(c) Sequential hierarchical phenotype tree

```{r fig3c, results="asis"}
panel_slot(
  "fig3c_phenotype_tree",
  "The sequential gating scheme and its terminal populations",
  c("Draw from the ACTUAL `flowpath.json` used for the study, not from memory.",
    "Sequential, not Boolean: a CD45-only pan-immune root, then the T/NK axis.",
    "Shade the CD45+CD3−CD56− catch-all node — the coverage caveat in the legend depends on it being visually marked.",
    "Reconcile the terminal-population count against the table printed below before writing a number into the legend."),
  source = "the study's `flowpath.json`")
```

The project's own record of what the gate tree emits is below. It is the join
table every analysis reads phenotypes through, so **if the figure and this table
disagree, the analyses are keyed to this one** — fix the code or the figure, but
do not leave them different.

```{r fig3c-table}
if (have_helpers) {
  fp <- dplyr::filter(phenotype_lineage_labels,
                      !grepl("_", phenotype_clean))     # FlowPath spellings only
  knitr::kable(fp, caption = paste(
    "FlowPath gate-tree labels as recorded in code/cell_tables.R, and the lineage",
    "each collapses to. Myeloid/Macrophage_M2 appear on the mirage side only —",
    "check the study panel.yaml before repeating the legend's 'no myeloid markers' caveat."))
}
```

---

# Figure 4 — cross-panel registration accuracy

## 4(a) Two-colour DAPI overlay, before and after

```{r fig4a, results="asis"}
panel_slot(
  "fig4a_dapi_overlay",
  "Two-colour DAPI overlay of two separately-acquired panels, before vs after registration",
  c("The pipeline already makes this: `GENERATE_REGISTRATION_QC` emits DAPI-overlay QC whenever `reg_qc >= 1`, which is the default (`reg_qc = 2`).",
    "But it writes at `qc_scale_factor = 0.25`. For a 'nuclei superimposed' claim, take a full-resolution crop at matched coordinates from the pre- and post-registration images instead.",
    "Two colours, one panel per channel, same display range before and after.",
    "Pick a field where the pre-registration drift is visible at figure size — a corner usually drifts more than the centre.",
    "Record the pixel size for the scale bar."),
  source = "mirage `GENERATE_REGISTRATION_QC` output, or full-res crops from the same run")
```

## 4(b) and 4(c) Target registration error and Dice, by configuration — **computed**

The arm-ranking panels. There are **two possible sources**, and they are not
equivalent:

1. **The real arm sweep** (`data/registration_arms/`) — the study slides registered
   once per configuration. This is what the figure should use.
2. **The synthetic benchmark sweep** (`data/benchmark/param_matrix.csv`) — mirage's
   `benchmarks/` runs, whose moving panels are shifted copies of channel 0. A
   residual measured there reflects a known injected offset, not real
   cross-acquisition difficulty.

This chunk prefers (1) and falls back to (2), and says which it used — because
which one produced the figure changes what the figure is allowed to claim.

```{r fig4bc, fig.width=panel_width(150), fig.height=panel_height(96)}
reg_dir  <- here("data", "benchmark")
arm_figs <- list(); arm_source <- NULL

.man <- tibble::tibble(); .seg <- tibble::tibble(); .val <- tibble::tibble()
if (have_helpers && dir.exists(here("data", "registration_arms"))) {
  suppressMessages(source(here("code", "registration_arms.R")))
  .man <- arm_manifest()
  .seg <- if (nrow(.man)) read_arms_seg_qc(.man) else tibble::tibble()
  .val <- if (nrow(.man)) read_arms_valis(.man)  else tibble::tibble()
  if (nrow(.seg)) {
    # (b) is VALIS's own reported error and (c) the independent nucleus overlap,
    # each at the arm's final transform, each with the no-registration and
    # rigid-only baselines read from the depth-0 runs. See build_arm_paper_figs().
    arm_figs   <- build_arm_paper_figs(.seg, .val, .man)[c("tre_by_arm", "dice_by_arm")]
    arm_source <- "real"
  }
}
if (!length(arm_figs) && have_helpers && file.exists(file.path(reg_dir, "param_matrix.csv"))) {
  source(here("code", "registration_accuracy_plots.R"))
  arm_figs   <- build_reg_figs(reg_dir)[c("06_tre_by_arm", "07_dice_by_arm")]
  arm_source <- "synthetic"
}
arm_figs <- arm_figs[!vapply(arm_figs, is.null, logical(1))]
for (p in arm_figs) print(p)
```

```{r fig4bc-note, results="asis"}
if (identical(arm_source, "real")) {
  cat("**Source: the real arm sweep** on the study slides — the right source for",
      "the manuscript. Arms are ranked at each run's FINAL transform, because",
      "`reg_micro_reg` 0 and 1 emit no `micro` stage and ranking on a fixed stage",
      "would drop those arms. (b) is the error VALIS reports from its own feature",
      "correspondences; (c) is the matched-nucleus Dice from mirage's staged QC, which",
      "never sees those features. The two baselines are the `native` / `original` and",
      "`rigid` stages of the **depth-0 runs only** — at depth ≥ 1 `rigid` already",
      "contains micro-rigid refinement, so it is not \"rigid only\" there. The tiled",
      "(STARE) arm writes no VALIS summary, so it cannot appear on (b); to keep (b)",
      "and (c) the same arms, as the legend says, it is left off (c) too. It stays on",
      "the [registration arms](registration_arms.html) page and in Additional file 2.",
      "See [registration arms](registration_arms.html) for the stage ladder, the",
      "pair-fraction trust check, and why `rigid` cannot be compared across depths.\n")
  if (!is.null(arm_figs$tre_by_arm))
    cat("\nVALIS error metric on (b): `", arm_figs$tre_by_arm$labels$y, "`.\n", sep = "")
} else if (identical(arm_source, "synthetic")) {
  cat("> **Source: the SYNTHETIC sweep.** These arms come from",
      "`data/benchmark/param_matrix.csv`, whose moving panels are shifted copies of",
      "channel 0. Usable for cost and scaling; **not** a real-tissue accuracy claim.",
      "Put the per-arm runs under `data/registration_arms/` and re-knit to replace",
      "this panel with the real one.\n")
} else {
  cat("_No arm data._ Either symlink the per-arm runs under",
      "`data/registration_arms/<arm>/` (preferred — see",
      "[registration arms](registration_arms.html)), or drop mirage's `paper_data`",
      "CSVs into `data/benchmark/`.\n")
}
```

> **What the sweep says about the legend.**
>
> - "Two VALIS accuracy presets differing in feature-matching resolution, not in
>   the feature matcher" is **correct**. Every `memory_mode` tier uses SuperPoint +
>   SuperGlue (mirage `bin/utils/valis_config.py`, `MEMORY_PRESETS`); `low` detects
>   on a 256 px processed image and fits the non-rigid field at 1024 px, `high` at
>   2048 / 4096 px. An earlier note on this page said `low` was BRISK/RANSAC; that
>   came from a comment in mirage's benchmark config, not from the shipped code.
> - Micro-registration is **not** on/off. `reg_micro_reg` is a nested depth
>   (`0` none, `1` micro-rigid, `2` + micro non-rigid), so a preset × depth sweep has
>   **six** arms, not four. The legend should say "three micro-registration depths".
> - The **rigid-only baseline is read from the depth-0 arms only**, because mirage's
>   `rigid` QC stage means affine ∘ micro-rigid whenever `reg_micro_reg ≥ 1`. The
>   no-registration baseline is the `native` stage (QC) / `original` column (VALIS)
>   of the same runs. Both are drawn in grey on (b) and (c) and are not arms.

---

# Figure 5 — proof-of-concept concordance

## The hot/cold axis — shared by (a), (b) and (c)

```{r hotcold-groups}
# WHICH VARIABLE DEFINES HOT AND COLD. The clinical table carries two different
# things: a continuous `HOT score` and a categorical `Immuno-phenotype`. The
# legend's selection axis is the score, so the score is used and the top/bottom
# three cases become the groups. That IS a threshold on rank — the legend may say
# "no threshold on the score" but not "no grouping". Set HOTCOLD_SOURCE to
# "immuno_phe" to group by the clinical category instead; it is a different
# variable and the figure must then say so.
#
# ONE derivation, paper_hotcold_groups(), feeds all three panels — and
# figures/fig5.R — so (c) cannot colour by one variable while (b) groups by another.
HOTCOLD_SOURCE <- "hot_score"

groups <- NULL; cases <- character(0)
if (have_cells)
  groups <- paper_hotcold_groups(here("data", "clinical_data.xlsx"),
                                 patient_ids = as_union$patient_id,
                                 source = HOTCOLD_SOURCE)
if (!is.null(groups)) {
  cases <- paper_representative_cases(groups, available = as_cells$patient_id)
  knitr::kable(dplyr::arrange(groups, group, dplyr::desc(hot_score)),
               caption = "Cases as grouped for panels 5(a)–(c), by hotscore rank.")
}
```

## 5(a) Registered, phenotyped cases — **computed**

`paper_phenotype_map()`, one panel per case. The two shown first are the
representative pair the legend asks for — the most extreme hotscore in each
group among the cases with imaging — derived from the same ranking as (b), so the
choice is reproducible rather than an `[author to supply]`. The remaining cases
follow for completeness.

```{r fig5a-representative, results="asis"}
.case <- function(g) if (g %in% names(cases)) cases[[g]] else "—"
if (length(cases))
  cat("Representative cases: **hot = ", .case("hot"), "**, **cold = ", .case("cold"),
      "**.\n", sep = "")
```

```{r fig5a, fig.width=panel_width(108), fig.height=panel_height(92)}
if (have_cells) {
  order <- c(unname(cases), setdiff(sort(unique(as_cells$patient_id)), unname(cases)))
  for (pid in order) {
    grp <- if (!is.null(groups)) groups$group[slide_key(groups$patient_id) == slide_key(pid)] else character(0)
    p <- paper_phenotype_map(as_cells, patient_id = pid, annots = as_polys,
                             title = paste0("Patient ", pid,
                                            if (length(grp)) paste0(" (", grp[1], ")") else ""))
    if (!is.null(p)) print(p)
  }
}
```

## 5(b) Imaging immune fraction, hot vs cold — **computed**

```{r fig5b, fig.width=panel_width(68), fig.height=panel_height(72)}
if (!is.null(groups) && nrow(as_union) > 0) {
  p <- paper_immune_fraction_hotcold(as_union, groups)
  if (!is.null(p)) print(p)
} else {
  cat("_Needs `data/clinical_data.xlsx` (for the hot/cold axis) and the all-slide cells._\n")
}
```

No test is applied and no coefficient is printed — with three cases a side the six
points *are* the evidence, and `paper_immune_fraction_hotcold()` deliberately never
calls `paired_spearman()`. Every other page on this site reports rho freely; this
panel is the one place that must not, because its legend says so.

## 5(c) Imaging vs bulk-RNA deconvolution — **computed**

```{r fig5c, fig.width=panel_width(144), fig.height=panel_height(112)}
paired_path <- here("output", "paired_deconv.rds")
if (have_helpers && file.exists(paired_path)) {
  # `groups` is (b)'s axis. Passing it is what makes hot in (b) mean hot in (c).
  p <- paper_deconv_scatter(readRDS(paired_path), method = "quantiseq", groups = groups)
  if (!is.null(p)) print(p)
} else {
  cat("_Needs `output/paired_deconv.rds`._ Knit [a molecular page](molecular_massimo2.html)",
      "first — its last chunk caches the paired frame so this panel does not have to",
      "re-run immunedeconv.\n")
}
```

No fit line and no coefficient, for the reason above and one more: the axes use
different denominators (imaging counts cells, deconvolution estimates a mixture
fraction), so a regression line would invite exactly the absolute-agreement
reading the legend disclaims. **The populations plotted are fixed by
`deconv_to_lineage()` and there are four** — CD4T, CD8T, Treg, NK. That answers
the legend's `[author to supply: full list of populations plotted]`.

---

# Additional files

```{r addfile-2}
# The same source rule as Fig 4(b)/(c): the real arm sweep first, the synthetic
# sweep as a labelled fallback. The table is the figure's numbers — VALIS TRE, the
# µm residual, matched-nucleus Dice and the pair fraction — per arm, plus the two
# baseline rows, so the legend's "[author to supply: per-arm values]" is read off
# here rather than from the plot.
if (identical(arm_source, "real")) {
  at <- arm_paper_table(.seg, .val, .man)
  dir.create(here("output"), showWarnings = FALSE, recursive = TRUE)
  readr::write_csv(at, here("output", "additional_file_2_registration_arms.csv"))
  knitr::kable(at, digits = c(NA, NA, NA, 0, 0, NA, 5, NA, 2, 3, 2), caption = paste(
    "Additional file 2 — per-arm registration accuracy on the study slides.",
    "Medians over slides at each arm's final transform; `no registration` and",
    "`rigid only` are the native and affine-only stages of the depth-0 runs.",
    "Also written to output/additional_file_2_registration_arms.csv."))
} else if (have_helpers && file.exists(file.path(reg_dir, "param_matrix.csv"))) {
  pm <- readr::read_csv(file.path(reg_dir, "param_matrix.csv"), show_col_types = FALSE)
  at <- .reg_arm_table(pm)
  if (!is.null(at))
    knitr::kable(dplyr::arrange(at, tre), digits = 4, caption = paste(
      "Additional file 2 (SYNTHETIC sweep fallback) — per-arm registration accuracy.",
      "Replace with the real arm sweep before submission."))
} else {
  cat("**Additional file 2** (per-arm TRE and Dice table) needs the real arm sweep",
      "under `data/registration_arms/` (or, as a fallback, `data/benchmark/param_matrix.csv`).\n")
}
```

```{r addfile-4}
# Additional file 4: the population vocabulary, generated from the join objects the
# analyses use, so it cannot drift from what was computed. The legend calls the
# three groupings tiers; in this codebase they are cd45_over_inside (pan-immune),
# comparable_lineages (the four mapped lineages) and Immune_other (the catch-all).
if (have_helpers) {
  lt <- paper_lineage_table() |>
    dplyr::mutate(tier = dplyr::case_when(
      lineage %in% comparable_lineages ~ "Tier 2 — mappable lineage",
      lineage == "Immune_other"        ~ "Tier 3 — non-T/NK immune",
      lineage == "(unmapped)"          ~ "Tier 3 — non-T/NK immune (no imaging lineage)",
      TRUE                             ~ "not compared"))
  dir.create(here("output"), showWarnings = FALSE, recursive = TRUE)
  readr::write_csv(lt, here("output", "additional_file_4_lineage_mapping.csv"))
  knitr::kable(lt, caption = paste(
    "Additional file 4 — every phenotype call and every deconvolution cell type,",
    "with the lineage and comparison tier it is assigned to.",
    "Also written to output/additional_file_4_lineage_mapping.csv."))
}
```

**Additional files 1, 3, 5 and 6.**

- **1 — full pipeline configuration.** `mirage/docs/parameters.md` is already the
  complete per-module table. Missing from it: the compartment-to-marker map and a
  default-versus-recommended comparison; both still have to be written.
- **3 — Tier 2 / Tier 3 scatters.** Blocked: the tier vocabulary does not exist in
  this codebase. Its content maps onto `cd45_over_inside` (Tier 1),
  `deconv_to_lineage()` (Tier 2) and `Immune_other` (Tier 3) — settle the naming first.
- **4 — vocabulary mapping table.** Generated above from `paper_lineage_table()`,
  with the tier each row belongs to.
- **5 — epitope / tissue-integrity monitoring.** Per-round DAPI retention exists
  nowhere; imaging-team deliverable. For the CD8+ ⊂ CD3+ biological-consistency check,
  the machinery is [marker QC](marker_qc.html), which computes worst-case
  conditional co-positivity per marker pair straight from the cell tables. Two things
  to note: it is **exclusivity**-shaped (it asks whether two markers co-occur *more*
  than they should) while CD8 ⊂ CD3 is an **inclusion** constraint (CD8+CD3− should be
  rare), so it needs a negated-marker pair added; and it reads the exports directly, so
  unlike mirage-side QC it works on data this project actually has.
- **6 — computational performance.** Fully covered by
  [pipeline benchmarks](benchmark_pipeline.html); nothing to add.

---

## Exported files

```{r manifest}
out_dir <- here("output", "figures", SLUG)
if (dir.exists(out_dir)) {
  f <- list.files(out_dir, pattern = "\\.pdf$")
  if (length(f)) knitr::kable(tibble::tibble(panel = f),
                              caption = "Vector PDFs written for Affinity.")
}
cat("\nComputed panels land in `output/figures/", SLUG, "/`.",
    " Hand-made panels go in `output/figures/manual/` named as the slots above.\n", sep = "")
```

```{r export-pdfs, include=FALSE}
export_pdf_figures(SLUG)
```
