---
title: "Registration accuracy (landmark-free, in-pipeline)"
author: "sceriff0"
date: "2026-07-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      dev = c("png", "pdf"))
source(here("code", "pdf_export.R"))                     # export_pdf_figures()
reg_dir  <- here("data", "benchmark")
have_reg <- any(file.exists(file.path(reg_dir,
  c("registration_valis_rtre.csv", "registration_accuracy.csv", "param_matrix.csv"))))
```

## Starting data and notation

Two **independent, landmark-free** estimates of how well the Mirage pipeline
registers each moving slide onto its reference, plus a cost view and an explicit
agreement check. Every figure is built by sourcing
`code/registration_accuracy_plots.R` and rendered inline from the CSVs in
`data/benchmark/` — no PNGs on disk. Figures whose input is absent are silently
skipped.

**Where the numbers come from.** All tables are produced by mirage
`benchmarks/analysis/make_tables.py` (the `paper_data/` set) — drop them into
`data/benchmark/` and re-knit.

- **`registration_valis_rtre.csv`** — VALIS's *own* feature error per moving
  slide (`original_rTRE` → `rigid_rTRE` → `non_rigid_rTRE`, with `n_matches`;
  some VALIS builds emit `*_D` raw distances instead). rTRE is **relative** (a
  fraction of the image diagonal), so it is unitless and comparable across image
  sizes. Auto-harvested by `make_tables.py` from the summary CSVs VALIS writes
  during `register()` — the same numbers the pipeline's QC report shows
  ("Registration Accuracy (Valis rTRE)").
- **`registration_accuracy.csv`** — an *independent* re-scoring by DAPI-nucleus
  overlap (`bin/warp_seg_qc.py`, `reg_qc=2`, harvested by `make_tables.py`). Per
  (run, moving, stage ∈ {native, rigid, non_rigid, micro}): `dice_matched` (areal
  Dice over matched nuclei) and `displacement_um_p50/p90` (centroid residual, in
  microns). This does not trust VALIS's features — a second method for the same
  question.
- **`param_matrix.csv`** — one wide row per run: the headline registration
  metrics (`reg_displacement_um_p50`, `reg_dice_matched`) and the VALIS median
  (`valis_non_rigid_D`/`valis_non_rigid_rTRE`) joined to cost (`cpu_hours`).
- **`feature_dist/*.json`** (optional/legacy) — before/after matched-feature
  distance per slide; only present if a run set `enable_feature_error`.

The finding to look for: the VALIS feature rTRE (§1) and the segmentation-overlap
Dice/displacement (§2) are computed by **different methods** yet should agree that
rigid → non-rigid registration tightens alignment — §5 plots that agreement
directly. Convergence of independent estimates is the point.

```{r build}
reg_figs <- list()
if (have_reg) {
  source(here("code", "registration_accuracy_plots.R"))
  cat(length(reg_figs), "figure(s) built from", basename(reg_dir), "data.\n")
} else {
  cat("No registration data found — add mirage `paper_data` CSVs",
      "(`registration_valis_rtre.csv` / `registration_accuracy.csv` /",
      "`param_matrix.csv`) to `data/benchmark/` and re-knit.")
}
```

```{r fig-notes}
fig_notes <- c(
  "01_valis_rtre_by_stage" = paste(
    "VALIS self-reported feature error (`registration_valis_rtre.csv`: `original_rTRE`,",
    "`rigid_rTRE`, `non_rigid_rTRE`, or the `*_D` distance variants), one line per moving",
    "slide across the stages. rTRE is relative to the image diagonal (unitless); lower = better."),
  "01b_valis_n_matches" = paste(
    "Feature-match count per slide (`n_matches`) — the evidence behind each rTRE.",
    "A slide with few matches has a low-confidence estimate."),
  "02_overlap_dice_by_stage" = paste(
    "Independent overlap check: `dice_matched` from `registration_accuracy.csv` per",
    "stage (native → rigid → non_rigid → micro), box across slides, faint per-slide",
    "lines. Computed from DAPI-nucleus overlap, not VALIS features; higher = better."),
  "02b_displacement_um_by_stage" = paste(
    "Matched-nucleus centroid residual in microns (`displacement_um_p50`/`_p90`) per",
    "stage. Physical units; lower = tighter alignment. Same source table as above."),
  "03_feature_distance_reduction" = paste(
    "(Legacy) percent drop in mean matched-feature distance before → after registration",
    "(`feature_dist/*.json`); only present if a run enabled `enable_feature_error`."),
  "04_accuracy_vs_cost" = paste(
    "Accuracy vs cost: `reg_displacement_um_p50` vs `cpu_hours` from `param_matrix.csv`,",
    "one point per config. Lower-left is better — less residual for fewer CPU-hours."),
  "05_valis_vs_overlap_agreement" = paste(
    "Agreement of the two independent estimates: VALIS feature error",
    "(`valis_non_rigid_rTRE`/`_D`) vs matched-nucleus Dice (`reg_dice_matched`) per run",
    "(`param_matrix.csv`). Two different methods; a clear trend = they corroborate each other.")
)
```

## Figures

Each figure is preceded by its derivation note. Figures whose CSV was not present
in `data/benchmark/` are silently absent.

```{r figures, results="asis", fig.width=9, fig.height=5.5}
if (length(reg_figs)) {
  for (nm in names(reg_figs)) {
    title <- gsub("_", " ", sub("^[0-9]+[a-z]?_", "", nm))
    cat("\n\n### ", title, "\n\n", sep = "")
    if (!is.na(fig_notes[nm])) cat(fig_notes[[nm]], "\n\n")
    print(reg_figs[[nm]])
    cat("\n\n")
  }
}
```

## Notes

- **Two independent methods.** §1 is VALIS grading itself from feature
  correspondences; §2 re-scores the same registrations by nucleus overlap. They
  are not derived from each other, so agreement (§5) is meaningful.
- **rTRE is relative, displacement is physical.** rTRE (§1) is a fraction of the
  image diagonal; `displacement_um` (§2) is in microns. Read §1 for cross-size
  comparability, §2 for an interpretable physical residual.
- Absolute-scale registration cost on synthetic sweep offsets is covered in the
  [pipeline benchmarks](benchmarks.html); this page uses the real patient slides.

```{r export-pdfs, include=FALSE}
export_pdf_figures("registration_accuracy")
```
