---
title: "Registration arms — choosing a configuration on the real slides"
author: "sceriff0"
date: "2026-08-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      dev = c("png", "pdf"))
source(here("code", "pdf_export.R"))

SLUG <- "registration_arms"

have_arms <- dir.exists(here("data", "registration_arms"))
if (have_arms) suppressMessages(source(here("code", "registration_arms.R")))
```

## Starting data and notation

The study slides, registered **once per configuration**. Three pages now read
registration accuracy and they answer different questions — this one answers
*which configuration do we ship, and what does it buy?*

| page | data | question |
|---|---|---|
| [Registration accuracy](benchmark_registration.html) | mirage's `benchmarks/` sweep on **synthetic** images with a known injected offset | how does cost and accuracy scale? |
| [Run QC](registration_run_qc.html) | **one** run on the study slides | was this cohort registered well enough to analyse? |
| **this page** | the study slides, **N runs**, one per arm | which configuration, and what does each knob buy? |

The manuscript's registration-accuracy figure needs *this* page, not the sweep:
an arm ranking measured on real tissue rather than on a synthetic offset.

### The axes

**Two VALIS knobs, crossed:**

- **`memory_mode`** — the VALIS accuracy preset. `low` is BRISK/RANSAC at small
  dimensions; `high` is SuperPoint/SuperGlue at larger ones. These are **different
  feature matchers**, not one matcher at two resolutions — worth stating plainly,
  because a figure legend that says otherwise is describing a different experiment.
- **`reg_micro_reg`** — micro-registration **depth**, nested, *not* a boolean:
  `0` = none, `1` = micro-rigid only (refines `slide.M`), `2` = additionally micro
  non-rigid (`register_micro`).

Crossing them gives 2 × 3 = **six arms**, not the 2 × 2 an on/off reading would.

**Plus a backend comparator, on a different axis:**

- **`registration_method = tiled`** — STARE, at defaults. A *different backend*:
  JVM-free, internally tiled, runnable on a laptop. `memory_mode` and
  `reg_micro_reg` are **VALIS-only params**, so a tiled arm has neither and carries
  `NA` for both. It is not a seventh cell of the preset × depth grid; it is one point
  of comparison against the whole grid.

**Why a cross-backend comparison is legitimate at all.** The segmentation-overlap
scorer is method-agnostic — `bin/warp_seg_qc.py` takes `--method` and builds its
warper from either a VALIS registrar pickle or a STARE transform manifest — so the
metric is the same measurement in both cases. That is the one thing that makes
"which backend aligns better" answerable here.

**What is not comparable is the stage axis.** `lib/WarpBackends.groovy` declares one
stage list per backend and they are different vocabularies:

| backend | stages |
|---|---|
| `valis` | `native → rigid → non_rigid → micro` |
| `tiled` | `native → rigid → refined` |

Only `native` is shared as both a spelling and a meaning. `rigid` is shared as a
**word** and not as an **operation**: for VALIS it is the affine transform (composed
with micro-rigid at depth ≥ 1); for STARE it is the coarse global anchor before mesh
refinement.

**And each backend reports its own error in its own units, in its own file.** VALIS
writes rTRE (a fraction of the image diagonal) or a raw distance to
`registered/summary/*.csv`; STARE writes TRE in **pixels** to
`qc/registration/*_tre.json`, with a per-tile breakdown VALIS has no equivalent of.
Those get separate figures here, never a shared axis — a pixel TRE beside a
diagonal-relative rTRE is a unit error dressed up as a comparison.

### Where the numbers come from

Every arm is a normal mirage run at `reg_qc = 2`, so each emits the staged
segmentation-overlap QC: DAPI segmented on the **native** image, cells paired
**once** at the rigid anchor, then those same pairs re-scored after each stage.
Because the pairing is held fixed, a change between stages is geometry and not a
change in which cells got compared.

```
data/registration_arms/<arm>/<patient>/qc/registration/*_seg_qc.json
data/registration_arms/<arm>/<patient>/registered/summary/*.csv   (optional)
data/registration_arms/arms.csv                                   (optional, preferred)
```

Every reader here is `run_qc.R`'s, called once per arm tree — the artifacts are the
same artifacts, only the number of trees changed.

```{r manifest}
arms <- if (have_arms) arm_manifest() else tibble::tibble()
seg  <- if (nrow(arms)) read_arms_seg_qc(arms) else tibble::tibble()

if (nrow(arms)) {
  knitr::kable(dplyr::select(arms, arm, arm_dir, backend, memory_mode, micro_reg),
               caption = paste("The arms found, and how each was labelled.",
                               "A `tiled` arm carries no preset and no micro depth —",
                               "those params do not exist on that backend."))
} else {
  cat("**No arm runs found.** This page expects one mirage outdir per arm under",
      "`data/registration_arms/`:\n\n```\ndata/registration_arms/valis_high_micro2/<patient>/qc/registration/*_seg_qc.json\n```\n\n",
      "Symlink each run in, then re-knit.\n")
}
```

> **Label the arms explicitly.** Directory names are parsed for `high`/`low` and a
> micro depth, but a mislabelled arm does not fail — it produces a clean figure with
> the conclusion inverted. Write `data/registration_arms/arms.csv` and the parse is
> bypassed:
>
> ```
> arm_dir,backend,memory_mode,micro_reg
> valis_high_micro2,valis,high,2
> valis_low_micro0,valis,low,0
> tiled_defaults,tiled,,
> ```
>
> A directory name containing `tiled` or `stare` is read as the tiled backend and is
> *not* scanned for preset or depth — reading knobs off a run that never had them
> would invent values.

## What may be compared across arms, and what may not

```{r guard, results="asis"}
if (nrow(seg)) {
  cat(comparable_stage_note(seg), "\n\n")
  cat("Stages directly comparable across these arms: ",
      paste0("`", arm_comparable_stages(seg), "`", collapse = ", "), ".\n\n", sep = "")
}
```

Three traps are worth stating before any figure:

1. **A tiled arm's final stage is `refined`, and a VALIS-only stage vocabulary drops
   it.** `read_seg_qc()` factors the stage against the known list; anything unknown
   becomes `NA` and is filtered out. With a VALIS-only list, every STARE `refined`
   row vanished and the tiled arm's "final" stage came out as `rigid` — making the
   backend look far worse than it is, with no error and no missing-data warning. The
   stage vocabulary now covers both backends, and "the last stage" is read from each
   run's own `stage_order` rather than from a shared ordering.

2. **`rigid` is not "rigid only".** mirage defines the QC `rigid` stage as the rigid
   transform *after* `MicroRigidRegistrar` refined it. At `reg_micro_reg = 0` that is
   affine alone; at `≥ 1` it is affine ∘ micro-rigid. Putting `rigid` on one axis
   across a depth-crossed sweep plots two different transforms and reads as a
   micro-registration effect that is really a definition change. **A true rigid-only
   baseline exists only in the depth-0 arms.**
3. **`native` is only nearly comparable.** It is the untransformed segmentation, so
   it looks like a clean no-registration baseline — but the pairing is established at
   the rigid anchor, and arms with different rigid transforms pair different cells.
   Read it as a baseline; do not difference against it.

Which is why the arms are ranked on **each run's final stage** — the transform that
arm actually ships. That stage is not constant: depths 0 and 1 emit no `micro` stage
at all, so ranking on "the micro stage" would silently drop four of six arms.

```{r stages-present}
if (nrow(seg)) {
  knitr::kable(
    seg |> dplyr::count(micro_reg, stage) |>
      tidyr::pivot_wider(names_from = stage, values_from = n, values_fill = 0),
    caption = paste("Stages each depth actually reported. Depths 0 and 1 emit no",
                    "`micro` stage — at depth 1 the micro-rigid refinement has",
                    "already been folded into `rigid`."))
}
```

## Headline: the arm ranking

```{r ranking-table}
if (nrow(seg))
  knitr::kable(arm_ranking_table(seg), digits = 3, caption = paste(
    "Per arm, median across slides at that arm's final transform.",
    "Sorted by residual displacement — lower is better.",
    "`pair_fraction` is the trust check: read it before the ranking."))
```

```{r figures, results="asis", fig.width=7.5, fig.height=4.8}
arm_notes <- c(
  "01_final_residual_um_by_arm" = paste(
    "**The arm ranking.** Matched-nucleus centroid residual (median per slide) at",
    "each arm's final transform, in microns. Physical units and no resolution floor,",
    "so this is the number to quote. One point per slide: with a handful of slides",
    "the points are the evidence, not the box."),
  "02_final_dice_by_arm" = paste(
    "**The same ranking, measured differently.** Areal Dice over the matched pairs.",
    "It is bounded above by segmentation agreement rather than by registration, so",
    "read the ORDER, not the absolute level. Agreement with the residual figure is",
    "corroboration; disagreement is a reason to distrust both."),
  "03_stage_ladder_within_arm" = paste(
    "**What each stage bought, within an arm.** One line per moving slide, log y.",
    "Read DOWN a panel — inside one arm `rigid` has a fixed meaning. Do not read",
    "`rigid` ACROSS panels."),
  "04_pair_fraction_by_arm" = paste(
    "**Is the ranking trustworthy?** Fraction of nuclei paired at the rigid anchor.",
    "mirage's own rule: below ~0.5 the later stages are scored on a biased subset of",
    "cells that happened to land close. An arm that wins on residual while pairing",
    "thinly has not won."),
  "05_knob_effects" = paste(
    "**Which knob moved the result.** Final-transform residual by micro-registration",
    "depth, split by accuracy preset. The presets use different feature matchers",
    "(BRISK/RANSAC vs SuperPoint/SuperGlue), so a preset difference is a matcher",
    "difference."),
  "06_delta_vs_rigid_anchor" = paste(
    "**Did a stage make it worse?** Residual minus the rigid stage's. Above zero",
    "means that stage degraded the alignment. mirage catches-and-continues a failed",
    "micro-registration, so this regression is silent by design — this is where it",
    "surfaces."),
  "08_stare_intrinsic_tre_px" = paste(
    "**STARE grading itself**, before and after mesh refinement. In **pixels** — this",
    "is deliberately a separate figure from the VALIS rTRE above, because the two are",
    "in different units and a shared axis would invite a comparison that is not one."),
  "09_stare_tile_error_map" = paste(
    "**What only the tiled backend can show.** Residual error per tile, so a",
    "systematic regional failure is visible instead of averaged into a slide-level",
    "median. VALIS reports one number per slide and has no equivalent."),
  "07_valis_intrinsic_by_arm" = paste(
    "**VALIS grading itself**, one panel per arm, medians printed. Independent of every",
    "segmentation-overlap metric above — VALIS scoring its own feature correspondences —",
    "so agreement with the ranking figures is corroboration rather than restatement.",
    "**Three stages from two columns:** VALIS's `error_df` is `filename`, `rigid_D`,",
    "`non_rigid_D`; micro-registration has no column of its own because it *updates the",
    "non-rigid field*, so the micro value lives in the difference between the pre-micro",
    "and final summaries rather than in a column.",
    "**A missing `micro` box is the finding** — that arm wrote no pre-micro summary, so",
    "`reg_micro_reg < 2` and micro-registration never ran. A duplicated `non_rigid` box",
    "would instead read as 'micro bought nothing'.",
    "Read DOWN a panel: the stage meanings differ across depths, which is why this",
    "facets rather than sharing one axis. See [run QC](registration_run_qc.html) for the full",
    "column-and-file to stage mapping."))

figs <- if (nrow(seg)) build_arm_figs(seg, read_arms_valis(arms), arms) else list()
for (nm in names(figs)) {
  cat("\n\n### ", gsub("_", " ", sub("^[0-9]+_", "", nm)), "\n\n", sep = "")
  if (!is.na(arm_notes[nm])) cat(arm_notes[[nm]], "\n\n")
  print(figs[[nm]])
  cat("\n\n")
}
if (!length(figs)) cat("\n_No figures — see the loader note above._\n")
```

## Notes

- **This is the page the manuscript's registration figure should cite**, not
  [registration accuracy](benchmark_registration.html). That page reads the synthetic
  sweep, whose moving panels are shifted copies of channel 0, so a residual measured
  there reflects a known injected offset rather than real cross-acquisition
  difficulty. The arm ranking only means what the paper wants it to mean if it was
  measured on tissue.
- **Two independent estimates, again.** The segmentation-overlap metrics (figures 1–4)
  and VALIS's own error (figure 7) are computed by different methods from different
  inputs. Their agreement is the argument; neither alone is.
- **Dice has a ceiling below 1 under perfect registration.** Two independent
  segmentation runs on the same nucleus in different rounds do not produce identical
  outlines. The *delta* between stages is the registration signal; the absolute level
  is not.
- **A cheaper arm that pairs thinly is not cheaper.** `pair_fraction` gates every
  other number on the page — check figure 4 before quoting figure 1.
- **The tiled arm answers a different question from the VALIS grid.** The grid asks
  "which configuration of this backend?"; the tiled arm asks "is the other backend
  competitive at all?". It ran at defaults, so a poor showing is evidence about
  STARE-at-defaults and not about STARE — it has its own knobs
  (`reg_tiled_tile`, `reg_tiled_halo`, `reg_tiled_gate_tre`,
  `reg_tiled_coarse_max_dim`) that this sweep never varied. Say so if the paper
  reports it, or a reader will read a tuned-versus-untuned comparison as
  backend-versus-backend.
- **The tiled backend is the only one that localises its error.** If the paper needs
  to show *where* registration is hard on a slide rather than how hard on average,
  that figure only exists on this path.
- **The ANHIR/ACROBAT landmark evaluation is a different benchmark.** mirage's
  `benchmarks/registration_eval/` measures TRE against **landmark ground truth** on
  public datasets. Nothing on this page is ground-truth-anchored: it is a
  landmark-free internal comparison, which is the right tool for choosing between
  arms and the wrong one for an absolute accuracy claim.

```{r export-pdfs, include=FALSE}
export_pdf_figures(SLUG)
```
