---
title: "Run QC — registration accuracy on the study samples"
author: "sceriff0"
date: "2026-08-10"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

# The house style is sourced HERE, unconditionally and before opts_chunk$set:
# the figure defaults below call fig_width(), and on a page whose data is
# missing the analysis script that used to pull plot_theme.R in never runs —
# which silently left those plots on ggplot2's default grey theme.
source(here("code", "plot_theme.R"))           # house figure style + fig_width()/label_n()
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      # Figure size is NOT hand-picked: fig_width() pins the shrink factor
                      # from rendered width to printed column, which is what keeps type
                      # the same size on paper across every page. See plot_theme.R.
                      fig.width = fig_width("double"), fig.height = fig_height("double"),
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# This page reads ONLY the QC artifacts mirage published for the run, so it does not
# source load_data.R: it renders from a pipeline run alone, with no counts.RData, no
# clinical table and no FlowPath gating.
source(here("code", "run_qc.R"))              # readers + build_run_qc_figs()
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

SLUG <- "registration_run_qc"
qc_root <- RUN_QC_ROOT
```

## Starting data and notation

How well was **this cohort** registered and phenotyped? Everything here is measured
on the real patient slides, by the pipeline, during the run that produced the cells
the other pages analyse.

That is a different question from the one the benchmark pages answer. [Pipeline
benchmarks](benchmark_pipeline.html) and [registration accuracy](benchmark_registration.html)
read the tables mirage's `benchmarks/` sweep produces on **synthetic** images —
rescaled copies with a known injected offset, run to measure cost and scaling. Useful
for choosing parameters, silent about your samples. This page reads
`data/mirage/<patient>/` — the tree a mirage run writes directly — so a run already
copied in needs nothing further.

**Where the numbers come from.**

| artifact | what it is |
|---|---|
| `qc/registration/*_seg_qc.json` | staged **segmentation-overlap** QC (`reg_qc = 2`): matched-nucleus Dice, per-pair IoU and centroid residual, per moving slide per stage |
| `registered/summary/**/*.csv` | **VALIS's own** error, written during `register()` — `rigid_D` and `non_rigid_D` (two metric columns, no micro column) plus `n_matches`. Nextflow preserves `REGISTER`'s `preprocessed/data/` producer subdirectory on publish, so these sit two levels down; the reader recurses |
| `qc/registration/*_tre.json` | **STARE's own** TRE, when the tiled path ran — plus a per-tile spatial map |

**Three signals, and only one comparison is meaningful.** VALIS rTRE and STARE TRE
are both **intrinsic**: each method grading itself, from the very features it
registered on. They are not comparable to each other — only one of them ran — and
neither is evidence on its own, because a method that aligned its own features well
has demonstrated exactly that and nothing more. The segmentation-overlap Dice is
computed from **DAPI nuclei, not features**, by a different method entirely, so it
is the independent check on whichever intrinsic number this run has. §4 plots them
against each other; a run where the self-report looks excellent and the overlap does
not is the failure mode worth catching.

```
stage        what the transform is at that point
-----------  -----------------------------------------------------------
native       none — the raw between-slide offset
rigid        rigid only (post micro-rigid). THE ANCHOR: correspondences are
             established once here and then held fixed, so every later change
             is geometry rather than a different set of cells being paired
non_rigid    rigid + the wave-1 non-rigid field
micro        + the micro residual — what the pipeline actually ships
```

**Two traps in these files.**

- **Two files, four stages, three columns — and the two files are not two
  measurements of the same thing.** `register_micro()` re-runs `measure_error()` and
  **overwrites** `<name>_summary.csv`, composing the micro residual into the same
  displacement field. `<name>_summary_premicro.csv` is a copy taken just before that,
  and it is written **only at `reg_micro_reg = 2`** — below that `register_micro()`
  never runs, so the plain summary already *is* the pre-micro one.

  **Why there is no micro column.** VALIS's `error_df` schema is
  `from`/`filename`, `rigid_D`, `non_rigid_D` — that is the whole of it.
  Micro-registration is not modelled as a fourth stage with its own column; it is an
  **update to the non-rigid field**. So the micro number does not live in a column at
  all: **it lives in the difference between the two files.**

  The stage axis is therefore recovered from *which file* a value came from, not from
  which column:

  | stage | column | file |
  |---|---|---|
  | `rigid` | `rigid_D` | final — unchanged by micro, so either file gives it |
  | `non_rigid` | `non_rigid_D` | **pre-micro** — the only file that still isolates it |
  | `micro` | `non_rigid_D` | **final** — same column, other file |

  Two columns, two files, three stages. That is exactly the reconstruction mirage's own
  report performs (`bin/generate_qc_report.py:_RECONCILE_TRE_SOURCE`), and it sorts the
  two files purely on the `_summary_premicro.csv` filename suffix.

  It is also why reading the two files as "pre-micro" and "post-micro" *panels* looked
  like a duplicate: `rigid_D` is the same number in both, so one of two stages was drawn
  twice, and the one that did differ was given the same label in both panels when the two
  values are two different stages.

  **A blank micro stage is a claim, not a gap.** With no pre-micro file
  (`reg_micro_reg < 2`, the shipped default) the final `non_rigid_D` *is* the pre-micro
  value, so it becomes `non_rigid` and no `micro` box is drawn. Drawing one anyway — a
  byte-for-byte duplicate of `non_rigid` — would read as *"micro bought nothing"*, which
  is a different statement from *"micro did not run"*. mirage makes the same choice on
  the segmentation-overlap side: when micro-registration is skipped, or raises and is
  caught, the QC omits the stage rather than duplicating `non_rigid`.

  **STARE has no micro stage at all.** The tiled backend writes no
  `registered/summary/*.csv`, so it never reaches this axis; its intrinsic TRE comes from
  `*_tre.json` in pixels and is plotted on its own.
- **Pair fraction gates everything above it.** A Dice computed over 30% of the cells
  is a Dice over a thin and probably biased subset. Read §3c before quoting any
  accuracy number from §3.

```{r build}
qc_tables <- run_qc_tables(qc_root)
qc_figs   <- build_run_qc_figs(qc_root, qc_tables)

found <- vapply(qc_tables, nrow, integer(1))
if (sum(found) == 0) {
  cat("No mirage QC found under `", qc_root, "`.\n\n",
      "Copy (or symlink) a pipeline run's `--outdir` to `data/mirage/` and re-knit —\n",
      "the per-patient `qc/` and `registered/summary/` directories\n",
      "are what this page reads.\n", sep = "")
} else {
  cat("artifacts found — ",
      paste(sprintf("%s: %d rows", names(found), found), collapse = " | "), "\n", sep = "")
  path <- if (nrow(qc_tables$stare)) "tiled (STARE)" else
          if (nrow(qc_tables$valis)) "VALIS"          else "unknown"
  cat("registration path in this run: ", path, "\n", sep = "")
}
```

## Headline accuracy per slide

One row per moving slide at the **final** stage — the alignment the cells were
actually quantified on. `dice_matched` is the matched-nucleus overlap and
`disp_um_p50` the median centroid residual in microns, which is the number to quote
as the pipeline's spatial resolution on this cohort. `pair_fraction` is how much of
the slide those two are based on.

```{r headline}
if (nrow(qc_tables$seg_qc)) {
  qc_tables$seg_qc |>
    group_by(patient_id, moving) |>
    slice_max(as.integer(stage), n = 1, with_ties = FALSE) |>
    ungroup() |>
    transmute(patient_id, moving, final_stage = as.character(stage),
              dice = round(dice_matched, 3),
              disp_um_p50 = round(disp_um_p50, 2),
              disp_um_p90 = round(disp_um_p90, 2),
              pair_fraction = round(pair_fraction, 2),
              gain_vs_rigid_dice = round(d_dice_vs_rigid, 3),
              gain_vs_rigid_um   = round(d_disp_um_vs_rigid, 2)) |>
    arrange(desc(disp_um_p50)) |>
    knitr::kable(caption = paste(
      "Registration accuracy per moving slide, worst residual first. The gain_* columns",
      "are the change from the rigid anchor: negative micron gain = the non-rigid and",
      "micro stages tightened the alignment. A low pair_fraction undermines the row."))
}
```

```{r fig-notes}
fig_notes <- c(
  "01_valis_error_by_stage" = paste(
    "VALIS's own error across `original` → `rigid` → `non_rigid` → `micro`, one box per",
    "stage with every slide overlaid and the **median printed**. rTRE is relative to the",
    "image diagonal (unitless, comparable across slide sizes); some builds emit raw",
    "`*_D` distances instead. **Intrinsic** — VALIS scoring its own feature matches.",
    "The four stages come from three columns across two files: `non_rigid` from the",
    "pre-micro summary, `micro` from the final one (same column, micro composed in).",
    "A missing `micro` box means `reg_micro_reg < 2` and micro-registration never ran."),
  "01b_valis_n_matches" = paste(
    "`n_matches`: how many correspondences each rTRE above is based on. A slide with",
    "few matches has a low-confidence estimate no matter how good the number looks."),
  "02_stare_tre" = paste(
    "STARE's intrinsic TRE from `qc/registration/*_tre.json`, in pixels.",
    "`coarse` is the RMS ORB residual after the global rigid M0 (directly comparable",
    "to VALIS's rigid error); the per-tile rigid median and p90 summarise local",
    "misalignment before the mesh; `after mesh` is the post-refinement residual —",
    "STARE's analogue of VALIS's non-rigid error."),
  "02b_stare_tre_map" = paste(
    "The per-tile records behind those percentiles, laid out on the slide grid.",
    "**This is what the tiled path gives that VALIS cannot**: a summary number averages",
    "a hot corner away, and tissue folds and slide edges fail locally. A patch of high",
    "TRE means cells there are misregistered even when the slide's median looks fine."),
  "03_overlap_dice_by_stage" = paste(
    "Matched-nucleus Dice per stage, from `qc/registration/*_seg_qc.json`. Computed by",
    "warping each slide's natively-segmented DAPI nuclei and measuring overlap — it does",
    "not use the features the registration was fitted on, which is what makes it the",
    "independent check. Correspondences are fixed at the rigid anchor, so a change",
    "between stages is geometry, not a different set of pairs."),
  "03b_displacement_um_by_stage" = paste(
    "Matched-nucleus centroid residual in **microns** (`displacement_um_p50`/`_p90`).",
    "Physical units, so this is the figure to quote as the pipeline's spatial",
    "resolution on this cohort — and to compare against cell diameter when deciding",
    "whether per-cell marker assignment across cycles is trustworthy."),
  "03c_pair_fraction" = paste(
    "The fraction of cells that paired between the two slides, per moving slide. Every",
    "Dice and displacement above is computed over this subset. Below the dashed 0.5 line",
    "the pairing is thin and probably biased toward the well-aligned regions, which",
    "flatters the accuracy numbers."),
  "04_intrinsic_vs_overlap" = paste(
    "The self-report against the independent check, one point per moving slide at the",
    "final stage: whichever intrinsic error this run has (STARE TRE, else VALIS) on x,",
    "matched-nucleus Dice on y. They should track. A slide with low intrinsic error and",
    "low Dice is one where the method aligned its own features but not the tissue.")
)
```

## Figures

Each figure is preceded by its derivation note. A figure whose artifact this run did
not produce is simply absent — a VALIS run has no §2, a tiled run has no §1.

```{r figures, results="asis", fig.width=fig_width("double"), fig.height=5.5}
for (nm in names(qc_figs)) {
  title <- gsub("_", " ", sub("^[0-9]+[a-z]?_", "", nm))
  cat("\n\n### ", title, "\n\n", sep = "")
  if (!is.na(fig_notes[nm])) cat(fig_notes[[nm]], "\n\n")
  print(qc_figs[[nm]])
  cat("\n\n")
}
```

## Notes

- **Landmark-free throughout.** No manual fiducials are placed anywhere: the
  intrinsic scores come from the registration's own correspondences and the overlap
  score from DAPI segmentation. That is what makes this runnable on every cohort,
  and also why the independent cross-check in §4 matters more than either number
  alone.
- **Intrinsic scores are not interchangeable.** VALIS rTRE is relative to the image
  diagonal and unitless; STARE TRE is in pixels; the overlap residual is in microns.
  Only the last is a physical distance. Compare across cohorts with the first,
  quote the third.
- **This page says nothing about cost.** Resource scaling, runtime and the
  segmentation-method comparison live on [pipeline benchmarks](benchmark_pipeline.html),
  which needs the sweep — it is not produced by an ordinary run.
- mirage also writes a self-contained HTML QC report per run (`<outdir>/qc/`) with
  the preprocessing, registration and segmentation **images**. This page is the
  numeric complement: it is what you can tabulate, compare across patients and carry
  into the analysis pages, not what you eyeball.

```{r export-pdfs, include=FALSE}
# Every plot above is also rendered to PDF by the dev = c("png", "pdf") chunk option;
# this collects them into output/figures/<SLUG>/. Runs last, so they all exist by now.
export_pdf_figures(SLUG)
```
